### 2024-08-02 data-exercise/data-exercise.qmd
redid repos

---
title: "Data Exercise"
author: "Leonel Salazar, IGE436"
---

# I am not 100% done, i still need to add comments and analysis for the models. I will plan on doing this by the end of the 4th of july weekend.

```{r}

# Load required packages
library(ggplot2)  # For advanced plotting
library(dplyr)    # For data manipulation
library(tidyr)    # For data tidying
library(tidyverse)

```

# Set seed for reproducibility

```{r}

set.seed(789)

```

# Number of observations

```{r}

n <- 3000

```

# Generate variables

```{r}

Gender <- sample(c("Male", "Female"), n, replace = TRUE)
Age <- round(rnorm(n, mean = 45, sd = 10))
Income <- round(rnorm(n, mean = 60000, sd = 20000))
Education <- sample(c("High School", "Bachelor's", "Master's", "PhD"), n, replace = TRUE, prob = c(0.25, 0.4, 0.25, 0.1))
HomeOwner <- sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.7, 0.3))
Retired <- sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.15, 0.85))

```

# Combine into a data frame

```{r}

data <- data.frame(Gender, Age, Income, Education, HomeOwner, Retired)

```

# Summary and structure of the dataset

```{r}

summary(data)
str(data)

```

```{r}

# Convert 'HomeOwner' from character to factor with levels "No" and "Yes"
data$HomeOwner <- factor(data$HomeOwner, levels = c("No", "Yes"))

# Show structure of 'HomeOwner' after conversion
str(data$HomeOwner)

data$Retired <- factor(data$HomeOwner, levels = c("No", "Yes"))

```

```{r}

# Frequency table for categorical variables
table(data$Gender)
table(data$Education)
table(data$HomeOwner)
table(data$Retired)

```

# Created histogram and box plot to check for normality and outliers.
```{r}

# Histograms for Age and Income
par(mfrow=c(1,2))  # Arrange plots in one row, two columns
# Histogram for Age
hist(data$Age, 
     main = "Age Distribution", 
     xlab = "Age", 
     ylab = "Frequency",
     col = "skyblue",  # Histogram color
     border = "black"  # Border color
)

# Histogram for Income
hist(data$Income, 
     main = "Income Distribution", 
     xlab = "Income", 
     ylab = "Frequency",
     col = "lightgreen",  # Histogram color
     border = "black"     # Border color
)

# Box plot for Age
boxplot(data$Age,
        main = "Box Plot of Age",
        ylab = "Age",
        col = "skyblue",  # Box color
        border = "black", # Border color
        horizontal = FALSE  # Vertical box plot
)

# Box plot for Income
boxplot(data$Income,
        main = "Box Plot of Income",
        ylab = "Income",
        col = "lightgreen",  # Box color
        border = "black",    # Border color
        horizontal = FALSE   # Vertical box plot
)

```
# Barplots to explore visually the categorical variables.

```{r}

# Bar plots for categorical variables
par(mfrow=c(2,2))  # Arrange plots in a 2x2 grid
barplot(table(data$Gender), main="Gender Distribution")
barplot(table(data$Education), main="Education Level")
barplot(table(data$HomeOwner), main="Home Ownership")
barplot(table(data$Retired), main="Retirement Status")

```

```{r}

# Model 1: Income ~ Age
model1 <- lm(Age ~ Age, data = data)

# Summary of the model
summary(model1)

str(data)

```

```{r}

# Model 2: Income ~ Age + Gender
model2 <- lm(Income ~ Age + Gender, data = data)

# Summary of the model
summary(model2)

```

```{r}

# Model 3: HomeOwner ~ Age + Income + Education
model3 <- glm(HomeOwner ~ Age + Income + Education, data = data, family = binomial)

# Summary of the model
summary(model3)

```

```{r}

# Model 4: Retired ~ Age + Income + Gender + Education
model4 <- glm(Retired ~ Age + Income + Gender + Education, data = data, family = binomial)

# Summary of the model
summary(model4)

```

### 2024-08-02 presentation-exercise/presentation-exercise.qmd
redid repos

---
title: "Presentation Exercise"
---

Placeholder file for the future data/results presentation exercise.

# load libraries

```{r}

library(ggplot2)
library(tidyverse)
library(here)
library(plotly)
library(flextable)

```

# Import data

```{r}

data <- read.csv("piedata.csv", header = TRUE, sep = ",")

data$Category <- as.factor(data$Category)
data$Value <- as.numeric(data$Value)

str(data)

```

# Create the Graph

```{r}

blue_palette <- scales::brewer_pal(palette = "Blues")(4)

ggplot(data, aes(x = "", y = Value, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() +
  scale_fill_manual(values = blue_palette) +  # Use the blue color palette
  labs(title = "RESIDENCES ARE THE LEADING LOCATION OF GUN THEFTS FROM CARS") +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, margin = margin(b = 20)))  # Center title and adjust bottom margin

```

```{r}
set_flextable_defaults(
  font.color = "black",
  border.color = "lightblue",
  theme_fun = "theme_box")

dat <- data.frame(
  Category = c("Other", "Parking Lot", "Residence", "Road"),
  Value = c(12, 32, 36, 20 )

)
flextable(dat)

```

```{r}

init_flextable_defaults()
flextable(dat)
```

### 2024-08-02 tidytuesday-exercise/tidytuesday-exercise(draft).qmd
redid repos

---
title: "Tidy Tuesday Exercise"
---

Placeholder file for the future Tidy Tuesday exercise.

```{r}
library(tidyverse)
library(ggplot2)
library(knitr)
library(dplyr)
library(here)
```

```{r}
#correctly removed first row and replaced with correct labels or variable names
Data <- read.csv("C:/Users/Leonel/Desktop/MSDA/MS Data Analytics/Current Class/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/tidytuesday-exercise/finalists.csv", header = T, na.strings = "?",
stringsAsFactors = T)
```

```{r}
str(Data)
```

```{r}
view(Data)
```

```{r}
# Select all columns except 4
contestants_data <- dplyr::select(Data, -c(4,5))
```

```{r}
contestants_data <- contestants_data %>%
   mutate(Birthday = as.Date(Birthday, format = "%d-%b-%y")) %>%
   mutate(Birthday = format(Birthday, "%d-%m-%y"))
```

```{r}
view(contestants_data)
```

```{r}
str(contestants_data)
```

```{r}
# Convert Contestant to character to clean it
contestants_data$Contestant <- as.character(contestants_data$Contestant)

# Convert Contestant column to UTF-8 encoding to handle encoding issues
contestants_data$Contestant <- iconv(contestants_data$Contestant, to = "UTF-8")

# Remove non-printable characters and unwanted symbols
contestants_data$Contestant <- gsub("[^[:print:]]", "", contestants_data$Contestant)

# Remove specific unwanted characters (e.g., quotes and slashes) if needed
contestants_data$Contestant <- gsub("[\"/]", "", contestants_data$Contestant)

# Convert Contestant back to factor
contestants_data$Contestant <- as.factor(contestants_data$Contestant)

# Ensure Birthday is treated as a factor (if needed)
contestants_data$Birthday <- as.factor(contestants_data$Birthday)

# Remove rows with NA values in Contestant, Birthday, and Season columns (if needed)
contestants_data_clean <- contestants_data %>%
  drop_na(Contestant, Birthday, Season)
```

```{r}
str(contestants_data_clean)
```
```{r}
#contestants_data_clean$Contestant <- as.numeric(contestants_data_clean$Contestant)
```

```{r}
view(contestants_data_clean)
```

```{r}
contestants_data_clean <- contestants_data_clean %>%
  drop_na(Birthplace)

```

```{r}
View(contestants_data_clean)
```

```{r}
write.csv(contestants_data_clean, "C:/Users/Leonel/Desktop/MSDA/MS Data Analytics/Current Class/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/tidytuesday-exercise/contestants_data_clean.csv", row.names = FALSE)
```

```{r}
library(ggplot2)

# Convert Contestant to numeric (just for plotting)
contestants_data_clean$Contestant <- as.numeric(contestants_data_clean$Contestant)

# Create histogram of Contestant numbers
ggplot(contestants_data_clean, aes(x = Contestant)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Contestants",
       x = "Contestant (Numeric Representation)",
       y = "Count") +
  theme_minimal()
```

```{r}
contestants_data_clean$Contestant_Num <- as.numeric(contestants_data_clean$Contestant)

ggplot(contestants_data_clean, aes(x = factor(Season), y = Contestant_Num)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Box Plot of Contestants by Season",
       x = "Season",
       y = "Contestant (Numeric Representation)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### 2024-08-02 tidytuesday-exercise/tidytuesday-exercise.qmd
redid repos

---
title: "tidytuesday-exercise"

---

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(knitr)
library(dplyr)
library(here)
library(caret)  # For machine learning models
library(rpart)  # For decision trees
library(randomForest)  # For random forest
library(gbm)  # For gradient boosting
library(e1071)  # For SVM
```

```{r}

# Correctly removed first row and replaced with correct labels or variable names
Data <- read.csv("C:/Users/Leonel/Desktop/MSDA/MS Data Analytics/Current Class/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/tidytuesday-exercise/finalists.csv", header = TRUE, na.strings = "?",
stringsAsFactors = TRUE)
```

```{r}
# Display the structure of the dataset
str(Data)

```

```{r}
# View the data
view(Data)

```

```{r}
# Select all columns except 4 and 5
contestants_data <- dplyr::select(Data, -c(4,5))

```

```{r}
# Convert Birthday to date format and reformat
contestants_data <- contestants_data %>%
   mutate(Birthday = as.Date(Birthday, format = "%d-%b-%y")) %>%
   mutate(Birthday = format(Birthday, "%d-%m-%y"))

```

```{r}
# View the transformed data
view(contestants_data)

```

```{r}
# Display the structure of the transformed data
str(contestants_data)

```

```{r}

# Clean the Contestant column
contestants_data$Contestant <- as.character(contestants_data$Contestant)
contestants_data$Contestant <- iconv(contestants_data$Contestant, to = "UTF-8")
contestants_data$Contestant <- gsub("[^[:print:]]", "", contestants_data$Contestant)
contestants_data$Contestant <- gsub("[\"/]", "", contestants_data$Contestant)
contestants_data$Contestant <- as.factor(contestants_data$Contestant)
contestants_data$Birthday <- as.factor(contestants_data$Birthday)

# Remove rows with NA values in Contestant, Birthday, and Season columns
contestants_data_clean <- contestants_data %>%
  drop_na(Contestant, Birthday, Season)

```

```{r}

# Display the structure of the cleaned data
str(contestants_data_clean)

```

```{r}
# View the cleaned data
view(contestants_data_clean)

```

```{r}

# Remove rows with NA values in Birthplace
contestants_data_clean <- contestants_data_clean %>%
  drop_na(Birthplace)

```

```{r}

# View the cleaned data
view(contestants_data_clean)

```

```{r}
# Save the cleaned data to a CSV file
write.csv(contestants_data_clean, "C:/Users/Leonel/Desktop/MSDA/MS Data Analytics/Current Class/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/tidytuesday-exercise/contestants_data_clean.csv", row.names = FALSE)

```

```{r}
# Load ggplot2 for plotting
library(ggplot2)

# Convert Contestant to numeric for plotting
contestants_data_clean$Contestant <- as.numeric(contestants_data_clean$Contestant)

# Create histogram of Contestant numbers
ggplot(contestants_data_clean, aes(x = Contestant)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "black") +
  labs(title = "Distribution of Contestants",
       x = "Contestant (Numeric Representation)",
       y = "Count") +
  theme_minimal()

```

```{r}
# Add numeric representation of Contestant
contestants_data_clean$Contestant_Num <- as.numeric(contestants_data_clean$Contestant)

# Create box plot of Contestants by Season
ggplot(contestants_data_clean, aes(x = factor(Season), y = Contestant_Num)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Box Plot of Contestants by Season",
       x = "Season",
       y = "Contestant (Numeric Representation)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}

# Split the data into training and testing sets
set.seed(123)
train_index <- suppressWarnings(createDataPartition(contestants_data_clean$Contestant_Num, p = 0.8, list = FALSE))
train_data <- contestants_data_clean[train_index, ]
test_data <- contestants_data_clean[-train_index, ]

# Linear Regression
lm_model <- suppressWarnings(train(Contestant_Num ~ ., data = train_data, method = "lm"))
lm_pred <- suppressWarnings(predict(lm_model, test_data))
lm_rmse <- RMSE(lm_pred, test_data$Contestant_Num)

# Decision Tree
tree_model <- suppressWarnings(train(Contestant_Num ~ ., data = train_data, method = "rpart"))
tree_pred <- suppressWarnings(predict(tree_model, test_data))
tree_rmse <- RMSE(tree_pred, test_data$Contestant_Num)

# Random Forest
rf_model <- suppressWarnings(train(Contestant_Num ~ ., data = train_data, method = "rf"))
rf_pred <- suppressWarnings(predict(rf_model, test_data))
rf_rmse <- RMSE(rf_pred, test_data$Contestant_Num)

# Gradient Boosting
gbm_model <- suppressWarnings(train(Contestant_Num ~ ., data = train_data, method = "gbm", verbose = FALSE))
gbm_pred <- suppressWarnings(predict(gbm_model, test_data))
gbm_rmse <- RMSE(gbm_pred, test_data$Contestant_Num)

# Support Vector Machine
svm_model <- suppressWarnings(train(Contestant_Num ~ ., data = train_data, method = "svmRadial"))
svm_pred <- suppressWarnings(predict(svm_model, test_data))
svm_rmse <- RMSE(svm_pred, test_data$Contestant_Num)

# Combine RMSE results into a dataframe for comparison
results <- suppressWarnings(data.frame(
  Model = c("Linear Regression", "Decision Tree", "Random Forest", "Gradient Boosting", "Support Vector Machine"),
  RMSE = c(lm_rmse, tree_rmse, rf_rmse, gbm_rmse, svm_rmse)
))

print(results)

```

| Model                     | RMSE       |
|---------------------------|------------|
| Linear Regression         | 0.000000   |
| Decision Tree             | 23.2707546 |
| Random Forest             | 0.8732771  |
| Gradient Boosting         | 3.2988005  |
| Support Vector Machine    | 42.7373149 |

### Analysis

1. **Linear Regression**:
   - **RMSE: 0.000000**
   - The RMSE value of zero suggests perfect prediction, which is highly unusual. This might indicate an issue with the model or the way it was applied. It's worth checking the implementation for any possible errors or overfitting.

2. **Decision Tree**:
   - **RMSE: 23.2707546**
   - This relatively high RMSE indicates that the Decision Tree model is not performing well on this dataset. Decision trees can overfit to the training data if not pruned properly, and this might be a case of overfitting or lack of sufficient depth to capture the complexity of the data.

3. **Random Forest**:
   - **RMSE: 0.8732771**
   - The Random Forest model performs quite well, with a very low RMSE. This suggests that the ensemble approach of averaging multiple decision trees helps in capturing the data's patterns more effectively than a single decision tree.

4. **Gradient Boosting**:
   - **RMSE: 3.2988005**
   - Gradient Boosting also shows good performance, though not as strong as Random Forest. This method sequentially builds models to correct the errors of previous models, which often leads to high accuracy, but it may require careful tuning.

5. **Support Vector Machine (SVM)**:
   - **RMSE: 42.7373149**
   - The high RMSE for the SVM indicates that this model is not suitable for this particular dataset or problem. SVMs can be very effective but often require specific tuning and may not perform well with certain types of data or without proper parameter optimization.

### Conclusion

- **Best Performing Model**: The **Random Forest** model is the best performing among the ones compared, with an RMSE of 0.8732771, indicating a good balance between bias and variance and an ability to generalize well on unseen data.
- **Possible Issues**: The RMSE of zero for Linear Regression should be investigated as it is highly unusual and suggests a perfect fit which is rare in practical scenarios.
- **Room for Improvement**: While Gradient Boosting also performs well, further tuning of its parameters could potentially improve its performance. The Decision Tree and SVM models do not perform as well and may need different configurations or may not be suitable for this specific task.

The next steps could involve:
1. Investigating the Linear Regression model for any anomalies.
2. Fine-tuning the Random Forest and Gradient Boosting models further.
3. Considering additional preprocessing steps or feature engineering to improve the overall performance of the models.

### 2024-08-02 Group10/manuscript_final.qmd
updating

---
title: "Honey Bee Colonies Impacted by Varroa, American Fouldbrood and Global Warming"
subtitle: ""
author: Edward Cruz, Jr, Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(ggplot2)
library(dplyr)
library(scales)
library(showtext)
library(ggimage)
```

# Summary/Abstract

Group Ten is pursuing the historical data collected by various public agencies to determine if the impact of Varroa, American Foulbrood, and global warming has demonstrated an increase in hive losses across the United States and Texas. The research will be accomplished using data sets derived from the agencies National Agricultural Statistics Service, Agricultural Statistics Board, and United States Department of Agriculture (USDA). Data sets containing several years of hive losses based on varroa and bacterium losses including recent possible environmental thermal global warming. A visualization of outcomes using R demonstrating and validating possible detrimental effects on honeybee colonies in the United States and Texas wrought by the negative impact of mites, bacterium, and global warming that will affect honey production and inevitably impact food production.

{{< pagebreak >}}

# Introduction

```{r}
# Create a data frame with the composition of honey
honey_data <- data.frame(
  component = c("Fructose", "Glucose", "Water", "Maltose", "Trisaccharides, Carbohydrates", "Sucrose", "Minerals, Vitamins, Enzymes"),
  percentage = c(38.5, 31.0, 17.1, 7.2, 4.2, 1.5, 0.5)
)

# Define colors for the segments
honey_data$color <- c("#FFA726", "#FB8C00", "#FFD54F", "#FFB74D", "#90CAF9", "#F06292", "#BA68C8")

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```

Honeybees have been nature's pollinators for centuries, with documented relationships with humans dating back to ancient Egyptians and Hindus. Historically, humans have maintained beehives, using honey as medicine in cultures such as the Egyptians, Assyrians, Chinese, Greeks, and Romans. The natural antibacterial properties of honey made it a valuable treatment for wounds, preventing infection, a practice used by Romans and Russians during World War I. Honeybees and other pollinators are critical for food production and nutritional security, yet bees face a variety of survival challenges. Currently, Varroa mites impact bee colonies and this mite infestation, a tiny red-brown parasite that can live on adult honeybees and reproduce on larvae and pupae in the developing brood. Another major threat is American Foulbrood Disease (AFB), caused by the bacterium Paenibacillus larvae. A disease that is fatal to honeybee larvae and found worldwide. The only effective control measure is to incinerate and destroy infected hives and live bees mitigating the infectious spread to other colonies. In addition, the exploration of climate change impacting honeybee colony losses has only recently been researched. While there are correlations between higher winter temperatures and greater colony losses, the effects of warmer autumn and winter temperatures on colony population dynamics and age structure as potential causes of reduced colony survival have not yet been fully investigated.Index Catalog // USDA Economics, Statistics and Market Information System. (n.d.-b). Index Catalog // USDA Economics, Statistics and Market Information System. (n.d.-a). , USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports. (n.d.).\
<https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en> <https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/>

You can add options to executable code like this

```{r}
library(ggplot2)
library(sf)
library(maps)
```

```{r}
states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
```

```{r}
varroa_data <- data.frame(
  state = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
            "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
            "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
            "maine", "maryland", "massachusetts", "michigan", "minnesota", 
            "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
            "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
            "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", 
            "south carolina", "south dakota", "tennessee", "texas", "utah", 
            "vermont", "virginia", "washington", "west virginia", "wisconsin", "wyoming"),
  year = c("1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995")
)
```

```{r}
states <- merge(states, varroa_data, by.x = "ID", by.y = "state", all.x = TRUE)
```

```{r}
ggplot(data = states) +
  geom_sf(aes(fill = year), color = "#292929", size = 0.2) +
  scale_fill_manual(
    values = c(
      "1986 - 1987" = "#CD2626", 
      "1988 - 1989" = "#CD9B1D", 
      "1990 - 1991" = "#FFD700", 
      "1992 - 1993" = "#00CD00", 
      "1994 - 1995" = "#104E8B"
    ),
    na.value = "#757575"
  ) +
  theme_minimal() +
  labs(
    title = "Spread of Varroa Mites",
    fill = "Year of First Appearance"
  )

```

Written By:sdns6mchl4. (2016, February 24). Varroa mite spread in the United States. Beesource Beekeeping Forums. https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/ \## General Background Information

[**Uncapped Honey Floresville,Texas Hive**]{.underline}

![](images/Bee%20Frame%202024-01.jpg){width="674"}

[**Capped Honey one Month Later same hive frame- Italian bees Floresville, Texas**]{.underline}

![](images/Bee%20Frame%20Capped%20Honey.jpg){fig-align="center" width="697"}

## Description of data and data source

Bee colonies maintained by beekeepers are considered livestock by the USDA due to their ability to produce honey, a consumable food item, and their essential role in assisting farmers with pollination crop seasons. Given the importance of bee colonies in agriculture, it was logical to source data from the following two authoritative websites: 1. USDA National Agricultural Statistics Service (NASS): This site provides comprehensive agricultural data, including statistics on honey production and colony health. 2. Bee Informed Partnership: This site offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations. Index Catalog // USDA Economics, Statistics and Market Information System. (n.d.-a). https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports. (n.d.). https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/

## Questions/Hypotheses to be addressed

Hypothesis: "The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.

[**Bacterium Infection Foul Brood.**]{.underline}

![](images/American-Foul-Brood-Comb-1024x768-1.jpg)

[**Dead bees resulting from extreme heat found in hive**]{.underline}.

![](images/dead-bees-hive-colony-collapse-600nw-2289939105.webp){width="679"}

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}

#knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

We got our data from the United States Department of Agriculture (USDA).

## Data import and cleaning

We decided to clean out our data from a few different datasets. We had to remove blank spaces and columns that were not pertinent to our analysis. We then filtered out other observations that did not directly deal with the data we are exploring. We are looking for cause of death to bee colonies and how they are affected by mites and climate change so we wanted to single out data that represented the losses so we can explore the different states by year and determine how the colonies were affected.

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(knitr)
```

```{r}

library(here)

# Install and load the here package
#install.packages("here")
library(here)

# Create a marker file (only needed once)
file.create("C:/Users/Leonel/Documents/DA 6833/Group Project/.here")

# Set the root directory manually

# Read the CSV file
data <- read.csv(here("Databystate.csv"))

data <- read.csv(here("Databystate.csv"))

#data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/DatabyState.csv")
```

```{r}
View(data)
```

```{r}
# Select all columns except 3, 6, and 9
Data_Clean <- dplyr::select(data, -c(3, 6, 9))
```

```{r}
# Output cleaned data file to a csv file.

write.csv(Data_Clean, "Data_Clean.csv", row.names = FALSE)

#write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Clean.csv")
```

```{r}
view(Data_Clean)
```

```{r}
# Assuming your data frame is named "data"

# Filter for rows where "Loss" or "Deadout" is present in any column (case-insensitive)

library(stringr)  # Load stringr package for regular expressions
data_filtered <- Data_Clean[rowSums(sapply(data, grepl, pattern = c("Loss"), ignore.case = TRUE)) > 0, ]
```

```{r}
# Output cleaned data file to a csv file.

write.csv(data_filtered, "Data_Filtered.csv", row.names = FALSE)

#write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Filtered.csv")
```

```{r}
view(data_filtered)
```

```{r}
# View data str to see what class they are.

str(data_filtered)
```

```{r}

# Load necessary libraries
library(ggplot2)
library(here)

str(data_filtered)

# Load the dataset
data_filtered <- read.csv(here::here("Data_Filtered.csv"), stringsAsFactors = FALSE)

# Convert 'Value' to numeric
data_filtered$Value <- as.numeric(gsub(",", "", data_filtered$Value))

# Handle missing values by replacing NAs with the median value
data_filtered$Value[is.na(data_filtered$Value)] <- median(data_filtered$Value, na.rm = TRUE)

# Boxplot of Value by Year and Data.Item
ggplot(data_filtered, aes(x = factor(Year), y = Value, fill = Data.Item)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boxplot of Value by Year and Data.Item", x = "Year", y = "Value") +
  theme_minimal()

# Scatterplot of Value by State.ANSI
ggplot(data_filtered, aes(x = State.ANSI, y = Value, color = factor(Year))) +
  geom_point() +
  labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
  theme_minimal()

```

```{r}
# Load necessary library
library(ggplot2)
library(scales)

# Assuming data_filtered is already loaded and has the required columns

# Create the plot
p <- ggplot(data_filtered, aes(x = Period, y = Value, fill = Data.Item)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boxplot of Value by Period and Data.Item", x = "Period", y = "Value")

# Print the plot
print(p)

```

\*\*\*\*\*\*\*\*\*\*\*\*\*\*\ Leo provided additional code below and works perfectly\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*\*

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(cluster)
library(car)
library(factoextra)
library(rpart)
library(rpart.plot)
library(randomForest)
library(e1071)

```

```{r}
# Load your dataset
# Assuming your data frame is named 'data_hcny'

data_hcny <- read.csv("C:/Users/Leonel/Documents/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/hcnydata_cleaned.csv")

data_clean <- na.omit(data_hcny)

# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)
```

```{r}
# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)
```

```{r}
# Visualizations

# Bar Plot for varroa_mites by State
ggplot(data_clean, aes(x = State, y = Varroa_Mites)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Varroa Mites by State", y = "Percentage", x = "State")
```

```{r}
# Stacked Bar Plot
data_long <- data_clean %>%
  gather(key = "factor", value = "percentage", -State)

ggplot(data_long, aes(x = State, y = percentage, fill = factor)) +
  geom_bar(stat = "identity") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Factors Affecting Bees by State", y = "Percentage", x = "State")
```

```{r}
# Heatmap
data_melt <- melt(data_clean, id.vars = "State")

ggplot(data_melt, aes(x = variable, y = State, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Factors Affecting Bees", y = "State", x = "Factor")
```

```{r}
# Pie Chart for the United States
us_data <- data_clean %>%
  filter(State == "United States") %>%
  dplyr::select(-State) %>%
  gather(key = "factor", value = "percentage")

ggplot(us_data, aes(x = "", y = percentage, fill = factor)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  theme_void() +
  labs(title = "Distribution of Factors Affecting Bees in the United States")
```

```{r}
# Correlation Analysis
cor_data <- cor(data_clean[,-1])
print("Correlation Matrix:")
print(cor_data)
```

```{r}
# Comparative Analysis
national_average <- data_clean %>%
  filter(State == "United States") %>%
  dplyr::select(-State)

state_differences <- data_clean %>%
  filter(State != "United States") %>%
  mutate(across(-State, ~ . - national_average))

print("State Differences from National Average:")
print(state_differences)
```

```{r}
# Assumption Checks for Descriptive Analysis
print("Assumption Checks:")
```

```{r}
# Normality check
shapiro_results <- lapply(data_clean[-1], function(column) shapiro.test(column))
print("Shapiro-Wilk Test for Normality:")
print(shapiro_results)
```

```{r}
# Homogeneity of variances
levene_results <- leveneTest(Varroa_Mites ~ State, data = data_clean)
print("Levene's Test for Homogeneity of Variances:")
print(levene_results)
```

```{r}
# Cluster Analysis
data_cluster <- data_clean %>%
  filter(State != "United States") %>%
  dplyr::select(-State)

hc <- hclust(dist(data_cluster), method = "complete")
plot(hc, labels = data_clean$state[data_clean$state != "United States"], main = "Hierarchical Clustering of States")
```

```{r}
# Principal Component Analysis (PCA)
data_pca <- data_clean %>%
  filter(State != "United States") %>%
  dplyr::select(-State) %>%
  scale()
```

```{r}
pca_result <- prcomp(data_pca, center = TRUE, scale. = TRUE)
summary(pca_result)
```

```{r}
fviz_pca_var(pca_result, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))

```

```{r}
# Decision Tree
tree_model <- rpart(Varroa_Mites ~ Other_Parasites_1. + Disease_2. + Pesticides + Other_3. + Unknown, data = data_clean, method = "anova")
rpart.plot(tree_model)
```

```{r}
# Support Vector Machine (SVM)
svm_model <- svm(Varroa_Mites ~ Other_Parasites_1. + Disease_2. + Pesticides + Other_3. + Unknown, data = data_clean)
summary(svm_model)
```

```{r}
# Assuming predictions and data_clean$varroa_mites are already created

# Plot the data
plot(predictions, data_clean$varroa_mites, 
     xlab = "Predictions", 
     ylab = "Varroa Mites", 
     main = "Predictions vs Varroa Mites",
     las = 1, # makes y-axis labels horizontal
     cex.lab = 1.2, # increases axis labels size
     cex.axis = 1.2) # increases tick labels size

```

![](images/_20220406_on_varroadestructor_minden_00463080.jpg){fig-align="center" width="541"}

```{r}
# Write the cleaned data to a new CSV file

library(here)

# Create a marker file (only needed once)
file.create("C:/Users/Leonel/Documents/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/.here")

here()

write.csv(hcny_CleanDraft, here("hcnydata_cleaned.csv"), row.names = FALSE)

#write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", row.names = FALSE)
```

```{r}
str(hcnydata_cleaned)
```

Varroa Mite

![](images/Varrpa%201.webp){fig-align="center" width="307"}

![](images/varroa-mite-bee.webp){fig-align="center" width="286"}

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

[**`Texas Bee gathering pollen from natural Texas foliage`**]{.underline}

![](images/Bee%20Feeding.jpg){fig-align="center" width="609"}

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2024-08-02 Group 10 Final Project/manuscript_final-2.qmd
trying to get webpage to work

---
title: "Honey Bee Colonies Impacted by Varroa, American Foulbrood and Global Warming"
subtitle: ""
author: Edward Cruz, Jr., Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
date: today
format:
  html:
    toc: false
    number-sections: true
    highlight-style: github
editor: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# load a few R packages
library(here)
library(knitr)
library(ggplot2)
library(dplyr)
library(scales)
library(showtext)
library(ggimage)

```

# Summary/Abstract

Group Ten is conducting a comprehensive analysis of historical data from
various public agencies to evaluate the impact of Varroa mites, American
Foulbrood, and global warming on hive losses in the United States.This
research utilizes extensive datasets from the National Agricultural
Statistics Service, the Agricultural Statistics Board, and the United
States Department of Agriculture (USDA), encompassing multiple years of
data. The data highlights the topic on hive losses attributed to mites,
bacterial infections, and environmental factors related to global
warming. Through advanced data visualization techniques in R, we aim to
demonstrate and validate the detrimental effects of these factors on
honeybee colonies, highlighting the consequent implications for honey
production and broader food security. ***Dennis and Kemp's study on
honey bee hive collapse [@hivecollapse] provides important insights into
Allee effects and ecological resilience.***

{{< pagebreak >}}

# Introduction

Honeybees have served as nature's pollinators for centuries. With their
relationship with humans documented as far back as ancient Egyptian and
Hindu cultures. Historically, humans have maintained beehives and
utilized honey for its medicinal properties in various civilizations,
including the Egyptians, Assyrians, Chinese, Greeks, and Romans. Honey's
natural antibacterial qualities made it a valuable treatment for wounds.
This was a practice continued by Romans and Russians during World War I.
Honeybees and other pollinators are essential for food production and
nutritional security. But even though they face numerous survival
challenges Varroa mites pose a significant threat to bee colonies. These
tiny red-brown parasites live on adult honeybees and reproduce on larvae
and pupae. Another major threat is American Foulbrood Disease (AFB).
This is caused by the bacterium Paenibacillus larvae, which is fatal to
honeybee larvae and requires incineration of infected hives to prevent
its spread. The impact of climate change on honeybee colony losses is a
growing area of research. While correlations have been found between
higher winter temperatures and increased colony losses, the effects of
warmer autumn and winter temperatures on colony dynamics and survival
require further investigation. ***Dennis and Kemp's study on honey bee
hive collapse [@hivecollapse] provides important insights into Allee
effects and ecological resilience.***[@usda_honey_bees; @usda_nass]

### Composition of Honey

The first image is a pie chart that illustrates the composition of honey
by percentage. The main components are:

-   **Fructose**: 38.5%
-   **Glucose**: 31%
-   **Water**: 17.1%
-   **Maltose**: 7.2%
-   **Other components**:
    -   Sucrose: 1.5%
    -   Minerals, Vitamins, Enzymes: 0.5%
    -   Trisaccharides, Carbohydrates: 4.2%

This composition indicates that honey is primarily made up of sugars,
specifically fructose and glucose, with water being the third major
component. The presence of minerals, vitamins, and enzymes, though in
smaller amounts, adds nutritional value to honey. This detailed
breakdown underscores honey's role as a natural sweetener with
additional health benefits beyond its primary carbohydrate content.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create a data frame with the composition of honey
honey_data <- data.frame(
  component = c("Fructose", "Glucose", "Water", "Maltose", "Trisaccharides, Carbohydrates", "Sucrose", "Minerals, Vitamins, Enzymes"),
  percentage = c(38.5, 31.0, 17.1, 7.2, 4.2, 1.5, 0.5)
)

# Define colors for the segments
honey_data$color <- c("#FFA726", "#FB8C00", "#FFD54F", "#FFB74D", "#90CAF9", "#F06292", "#BA68C8")

# Plot with ggplot2
ggplot(honey_data, aes(x = "", y = percentage, fill = component)) +
  geom_bar(width = 0.8, stat = "identity", color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = honey_data$color) +
  geom_text(aes(x=1.8, label = paste0(percentage, "%")), position = position_stack(vjust = 0.6), size = 3, color = "black") +
  labs(
    title = "COMPOSITION OF HONEY",
    subtitle = "Illustration of honey components by percentage",
    fill = NULL
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5, family = "lobster"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(sf)
library(maps)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
states <- st_as_sf(map("state", plot = FALSE, fill = TRUE))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
varroa_data <- data.frame(
  state = c("alabama", "alaska", "arizona", "arkansas", "california", "colorado", 
            "connecticut", "delaware", "florida", "georgia", "hawaii", "idaho", 
            "illinois", "indiana", "iowa", "kansas", "kentucky", "louisiana", 
            "maine", "maryland", "massachusetts", "michigan", "minnesota", 
            "mississippi", "missouri", "montana", "nebraska", "nevada", "new hampshire", 
            "new jersey", "new mexico", "new york", "north carolina", "north dakota", 
            "ohio", "oklahoma", "oregon", "pennsylvania", "rhode island", 
            "south carolina", "south dakota", "tennessee", "texas", "utah", 
            "vermont", "virginia", "washington", "west virginia", "wisconsin", "wyoming"),
  year = c("1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995",
           "1986 - 1987", "1988 - 1989", "1990 - 1991", "1992 - 1993", "1994 - 1995")
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
states <- merge(states, varroa_data, by.x = "ID", by.y = "state", all.x = TRUE)
```

### Spread of Varroa Mites

The second image is a map illustrating the spread of Varroa mites in the
United States from 1986 to 1995. The map uses different colors to
represent the year of the first appearance of Varroa mites in each
state:

-   **1986 - 1987**: Red
-   **1988 - 1989**: Yellow
-   **1990 - 1991**: Green
-   **1992 - 1993**: Blue
-   **1994 - 1995**: Grey
-   **NA**: States where data is not available

The map reveals the progressive spread of Varroa mites from the
mid-1980s to the mid-1990s, affecting bee colonies across the country.
The earliest appearances were concentrated in specific regions and
gradually spread to more states over time. This visual representation
highlights the increasing geographical distribution of Varroa mites and
underscores the growing challenge they pose to beekeeping and
agricultural industries in the United States.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = states) +
  geom_sf(aes(fill = year), color = "#292929", size = 0.2) +
  scale_fill_manual(
    values = c(
      "1986 - 1987" = "#CD2626", 
      "1988 - 1989" = "#CD9B1D", 
      "1990 - 1991" = "#FFD700", 
      "1992 - 1993" = "#00CD00", 
      "1994 - 1995" = "#104E8B"
    ),
    na.value = "#757575"
  ) +
  theme_minimal() +
  labs(
    title = "Spread of Varroa Mites",
    fill = "Year of First Appearance"
  )

```

***[@varroamap] Varroa Map of infestation periods***

```{r}
# Load necessary libraries
library(tidyverse)

# Create a data frame with the provided data
data <- tibble(
  Year = 1901:2023,
  `Earth's surface` = c(-0.15, -0.43, -1.4, -0.86, -1.02, -0.29, -0.54, 0.06, -0.59, 0.4, 0.01, -1.79, -0.48, -0.18, -0.57, -1.17, -1.96, -0.15, -0.47, -0.95, 1.78, 0.01, -0.38, -1.43, 0.5, -0.07, 0.13, -0.1, -1.17, -0.04, 1.52, -0.29, 0.97, 2.08, -0.12, 0.13, -0.47, 1.16, 1.24, -0.13, 0.64, -0.18, 0.05, -0.19, -0.27, 0.93, -0.1, -0.41, 0, -0.63, -0.9, 0.25, 1.35, 1.31, -0.33, 0.32, 0.02, -0.09, 0.09, -0.58, -0.15, -0.12, 0.24, -0.35, -0.33, -0.53, -0.26, -0.7, -0.52, -0.41, -0.36, -0.65, 0.27, 0.24, -0.52, -0.55, 0.53, -0.97, -1.14, 0.37, 1.1, -0.67, -0.14, -0.04, -0.72, 1.3, 1.31, 0.61, -0.18, 1.49, 1.14, 0.58, -0.76, 0.85, 0.63, -0.13, 0.18, 2.21, 1.86, 1.25, 1.68, 1.19, 1.24, 1.08, 1.62, 2.23, 1.63, 0.27, 0.37, 0.96, 1.16, 3.26, 0.41, 0.52, 2.38, 2.9, 2.53, 1.5, 0.66, 2.35, 2.49, 1.37, 2.39),
  `Lower troposphere (UAH)` = c(rep(NA, 78), 0.049093434, 0.731593434, 1.102093434, 0.008593434, 0.034093434, 0.292093434, 0.110593434, 1.220593434, 1.054093434, 0.698593434, 0.548593434, 1.451593434, 1.031593434, 0.484093434, -0.174406566, 1.043593434, 0.797593434, 0.502093434, 0.457093434, 1.691593434, 1.796593434, 1.351093434, 1.255093434, 0.994093434, 1.223593434, 0.769093434, 1.303093434, 1.444093434, 1.765093434, 0.467593434, 0.524593434, 0.929593434, 1.013593434, 1.955593434, 0.655093434, 0.763093434, 2.234593434, 2.030593434, 2.252593434, 1.645093434, 1.133593434, 1.970593434, 2.059093434, 1.567093434, NA),
  `Lower troposphere (RSS)` = c(rep(NA, 78), -0.808971111, 0.178178889, 0.658778889, -0.579771111, -0.383871111, -0.217371111, -1.053021111, 0.737978889, 1.111328889, 0.557978889, 0.221828889, 1.280228889, 1.043228889, 0.585278889, -0.261321111, 1.078178889, 0.772628889, 0.278078889, 0.418178889, 1.911578889, 1.882928889, 1.471328889, 1.536578889, 1.144628889, 1.465628889, 1.058528889, 1.583078889, 1.749278889, 1.858928889, 0.503378889, 0.595928889, 0.997028889, 1.109378889, 2.677328889, 0.696278889, 0.781478889, 2.547878889, 2.476328889, 2.350478889, 1.793978889, 1.221428889, 2.290628889, 2.485328889, 1.596878889, 1.886228889)
)
# Pivot the data for plotting
data_long <- data %>%
  pivot_longer(cols = c(`Earth's surface`, `Lower troposphere (UAH)`, `Lower troposphere (RSS)`), 
               names_to = "Measurement", values_to = "Temperature")

# Create the plot
ggplot(data_long, aes(x = Year, y = Temperature, color = Measurement)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  theme_minimal() +
  labs(
    title = "Temperature Changes Over the Years",
    x = "Year",
    y = "Temperature Anomaly (°C)",
    color = "Measurement"
  ) +
  scale_color_manual(values = c("Earth's surface" = "#254976", "Lower troposphere (UAH)" = "#4A8C1D", "Lower troposphere (RSS)" = "#B5153A"))

```

[@epa_climate_indicator] ***Global warming trends from 1900-2024***

***Spread of Varroa Mite by Year of First Apperance [@varroamap]***
Written By:sdns6mchl4. (2016, February 24). Varroa mite spread in the
United States. Beesource Beekeeping Forums.
https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/

\## General Background Information

### **Uncapped Honey Floresville,Texas Hive**

![](/images/Bee%20Frame%202024.jpg){fig-align="center"}

### **Capped Honey one Month Later same hive frame- Italian bees Floresville, Texas**

![](/images/Bee%20Frame%20Capped%20Honey.jpg){fig-align="center"}

## Description of data and data source

Bee colonies maintained by beekeepers are considered livestock by the
USDA due to their ability to produce honey, a consumable food item, and
their essential role in assisting farmers with pollination crop seasons.
Given the importance of bee colonies in agriculture, it was logical to
source data from the following two authoritative websites: 1. USDA
National Agricultural Statistics Service (NASS): This site provides
comprehensive agricultural data, including statistics on honey
production and colony health. 2. Bee Informed Partnership: This site
offers detailed insights and research on bee colony management and
health, contributing valuable information on the status and trends of
bee populations. Index Catalog // USDA Economics, Statistics and Market
Information System. (n.d.-a).
https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
USDA - National Agricultural Statistics Service - Surveys - honey bee
surveys and reports. (n.d.).
https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/

## Questions/Hypotheses to be addressed

Hypothesis: "The negative impacts of mites, bacterium, and global
warming have detrimental effects on honeybee colonies in the United
States and Texas, which in turn will lead to a decline in honey
production and negatively impact food production." This hypothesis can
be tested and validated through a visualization of outcomes using R,
demonstrating the relationship between these factors and their effects
on honeybee colonies.

### **Bacterium Infection Foul Brood**

![](/images/American-Foul-Brood-Comb-1024x768-1.jpg){fig-align="center"}

### **Dead bees resulting from extreme heat found in hive**

![](/images/dead-bees-hive-colony-collapse-600nw-2289939105.webp)

#Citation section To cite other work (important everywhere, but likely
happens first in introduction), make sure your references are in the
bibtex file specified in the YAML header above and have the right bibtex
key. Then you can include like this:

Examples of reproducible research projects can for instance be found in
[@mckay2020; @mckay2020a].

#Here are the BibTex entries for citations:

Using this in RMarkdown: #Make sure to save these entries in a file
names'reference.bib' and include it in your R Markdown document like
this:

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning
processes, and the analysis approaches. You might want to provide a
shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not
created with code (if you can do it with code, do it). @fig-schematic is
an example of some - completely random/unrelated - schematic that was
generated with Biorender. We store those figures in the `assets` folder.

```{r}

#knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

We got our data from the United States Department of Agriculture (USDA).

## Data import and cleaning

We decided to clean out our data from a few different datasets. We had
to remove blank spaces and columns that were not pertinent to our
analysis. We then filtered out other observations that did not directly
deal with the data we are exploring. We are looking for cause of death
to bee colonies and how they are affected by mites and climate change so
we wanted to single out data that represented the losses so we can
explore the different states by year and determine how the colonies were
affected.

```{r}
library(readxl)
library(tidyverse)
library(ggplot2)
library(knitr)
```

```{r}

library(here)

# Install and load the here package
#install.packages("here")
library(here)

here()
# Create a 'data' directory within the project
dir.create(here("data"))

# Read the CSV file
data <- read.csv(here("Databystate.csv"))

#data <- read.csv("C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/DatabyState.csv")
```

```{r}
# Select all columns except 3, 6, and 9
Data_Clean <- dplyr::select(data, -c(3, 6, 9))
```

```{r}
# Output cleaned data file to a csv file.

write.csv(Data_Clean, "Data_Clean.csv", row.names = FALSE)

#write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Clean.csv")
```

```{r}
# Assuming your data frame is named "data"

# Filter for rows where "Loss" or "Deadout" is present in any column (case-insensitive)

library(stringr)  # Load stringr package for regular expressions
data_filtered <- Data_Clean[rowSums(sapply(data, grepl, pattern = c("Loss"), ignore.case = TRUE)) > 0, ]
```

```{r}
# Output cleaned data file to a csv file.

write.csv(data_filtered, "Data_Filtered.csv", row.names = FALSE)

#write.csv(Data_Clean, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II #Repository/P2-Practicum-II-Portfolio-EdwardCruz/docs/Databystate_Filtered.csv")
```

```{r}
# View data str to see what class they are.

str(data_filtered)
```

### Boxplot of Value by Year and Data Item

The third image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" from 2015 to 2017. The boxplot
provides insights into: - The central tendency and spread of losses due
to both disorders. - Presence of outliers indicates that there are
occasional, significantly higher losses. - Comparative analysis shows
that both types of losses have varying distributions each year.

This indicates that while both types of losses are present each year,
the magnitude and frequency of extreme loss events can differ,
suggesting variability in the factors contributing to colony losses.

```{r}

# Load necessary libraries
library(ggplot2)
library(here)

str(data_filtered)

# Load the dataset
data_filtered <- read.csv(here::here("Data_Filtered.csv"), stringsAsFactors = FALSE)

# Convert 'Value' to numeric
data_filtered$Value <- as.numeric(gsub(",", "", data_filtered$Value))

# Handle missing values by replacing NAs with the median value
data_filtered$Value[is.na(data_filtered$Value)] <- median(data_filtered$Value, na.rm = TRUE)

# Boxplot of Value by Year and Data.Item
ggplot(data_filtered, aes(x = factor(Year), y = Value, fill = Data.Item)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boxplot of Value by Year and Data.Item", x = "Year", y = "Value") +
  theme_minimal()
```

### Scatterplot of Value by State ANSI

The fourth image is a scatterplot showing values for different states
(ANSI coded) across the years 2015 to 2017. Key points include: - Each
color represents a different year (2015: Red, 2016: Green, 2017:
Blue). - The scatterplot reveals the distribution of values across
states and highlights any trends or patterns over the three years.

This scatterplot indicates that while some states experience
consistently high or low values, others show significant variability
year over year. This could suggest regional differences in factors
affecting bee colony health.

```{r}
# Scatterplot of Value by State.ANSI
ggplot(data_filtered, aes(x = State.ANSI, y = Value, color = factor(Year))) +
  geom_point() +
  labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
  theme_minimal()

```

### Boxplot of Value by Period and Data Item

The fifth image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
THRU DEC**

Key insights include: - Seasonal trends in colony losses, with some
periods showing higher median losses or more variability. - Outliers
indicate extreme loss events occurring in specific periods.

This seasonal analysis suggests that certain times of the year are more
prone to higher losses, potentially due to environmental factors,
resource availability, or management practices.

```{r}
# Load necessary library
library(ggplot2)
library(scales)

# Assuming data_filtered is already loaded and has the required columns

# Create the plot
p <- ggplot(data_filtered, aes(x = Period, y = Value, fill = Data.Item)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = comma) +
  labs(title = "Boxplot of Value by Period and Data.Item", x = "Period", y = "Value")

# Print the plot
print(p)

```

### Boxplot of Value by Period and Data Item

The fifth image is a boxplot comparing the values for "LOSS, COLONY
COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
THRU DEC**

Key insights include: - Seasonal trends in colony losses, with some
periods showing higher median losses or more variability. - Outliers
indicate extreme loss events occurring in specific periods.

This seasonal analysis suggests that certain times of the year are more
prone to higher losses, potentially due to environmental factors,
resource availability, or management practices.

### Overall Analysis

-   The composition of honey indicates its high sugar content, essential
    for energy but also highlighting the presence of beneficial
    components like enzymes and vitamins.
-   The spread of Varroa mites map highlights the widespread and growing
    impact of these pests over time, emphasizing the need for ongoing
    management and control efforts.
-   The boxplots and scatterplots provide a detailed view of the
    variability and trends in colony losses due to different factors,
    both over time and across different regions. These plots suggest
    that while losses are a consistent issue, their magnitude and causes
    can vary widely, pointing to the need for tailored strategies to
    address colony health.

These visualizations collectively provide a comprehensive overview of
the challenges faced by bee colonies, from composition and health to
external threats like Varroa mites and seasonal variations in losses.

### **VARROA MITE EXPOSED**

![](/images/_20220406_on_varroadestructor_minden_00463080.jpg){fig-align="center"}

### **VARROA MITE**

![](/images/Varrpa%201.webp){fig-align="center" width="527"}

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(reshape2)
library(cluster)
library(car)
library(factoextra)
library(rpart)
library(rpart.plot)
library(randomForest)
library(e1071)

```

```{r}
# Load your dataset
# Assuming your data frame is named 'data_hcny'

data_hcny <- read.csv(here("hcny_CleanDraft.csv"))

#data_hcny <- read.csv("C:/Users/Leonel/Desktop/MSDA/MS Data Analytics/Current Class/DA 6833/Practicum 2 Github/leonelsalazar-P2-portfolio/Group Project/hcnydata_cleaned.csv")

data_clean <- na.omit(data_hcny)

```

```{r}
# Descriptive Statistics
print("Descriptive Statistics:")
summary(data_clean)
str(data_clean)
```

```{r}
# Visualizations

# Bar Plot for varroa_mites by State
ggplot(data_clean, aes(x = state, y = varroa_mites)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Varroa Mites by State", y = "Percentage", x = "State")
```

```{r}
str(data_clean)
```

```{r}
# Stacked Bar Plot
data_long <- data_clean %>%
  gather(key = "factor", value = "percentage", -state)

ggplot(data_long, aes(x = state, y = percentage, fill = factor)) +
  geom_bar(stat = "identity") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Factors Affecting Bees by State", y = "Percentage", x = "State")
```

```{r}
# Heatmap
data_melt <- melt(data_clean, id.vars = "state")

ggplot(data_melt, aes(x = variable, y = state, fill = value)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Factors Affecting Bees", y = "State", x = "Factor")
```

```{r}
# Pie Chart for the United States
us_data <- data_clean %>%
  filter(state == "United States") %>%
  dplyr::select(-state) %>%
  gather(key = "factor", value = "percentage")

ggplot(us_data, aes(x = "", y = percentage, fill = factor)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  theme_void() +
  labs(title = "Distribution of Factors Affecting Bees in the United States")
```

```{r}
# Correlation Analysis
cor_data <- cor(data_clean[,-1])
print("Correlation Matrix:")
print(cor_data)
```

```{r}
# Assumption Checks for Descriptive Analysis
print("Assumption Checks:")
```

```{r}
# Normality check
shapiro_results <- lapply(data_clean[-1], function(column) shapiro.test(column))
print("Shapiro-Wilk Test for Normality:")
print(shapiro_results)
```

```{r}
# Homogeneity of variances

levene_results <- leveneTest(varroa_mites ~ state, data = data_clean)
print("Levene's Test for Homogeneity of Variances:")
print(levene_results)
```

```{r}
# Cluster Analysis
data_cluster <- data_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state)

hc <- hclust(dist(data_cluster), method = "complete")
plot(hc, labels = data_clean$state[data_clean$state != "United States"], main = "Hierarchical Clustering of States")
```

```{r}
# Principal Component Analysis (PCA)
data_pca <- data_clean %>%
  filter(state != "United States") %>%
  dplyr::select(-state) %>%
  scale()
```

```{r}
pca_result <- prcomp(data_pca, center = TRUE, scale. = TRUE)
summary(pca_result)
```

```{r}
fviz_pca_var(pca_result, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))

```

```{r}
str(data_clean)
```

```{r}
# Support Vector Machine (SVM)
svm_model <- svm(varroa_mites ~ other_pests + disease + pesticides + other + unknown, data = data_clean)
summary(svm_model)
```

```{r}
# Assuming predictions and data_clean$varroa_mites are already created

# Example data
data_clean <- data.frame(
  Varroa_Mites = c(5, 10, 15, 20),
  other_variable = c(2, 4, 6, 8)
)

# Fit a linear model
model <- lm(Varroa_Mites ~ other_variable, data = data_clean)

# Generate predictions
predictions <- predict(model, newdata = data_clean)

# Plot the data
plot(predictions, data_clean$varroa_mites, 
     xlab = "Predictions", 
     ylab = "Varroa Mites", 
     main = "Predictions vs Varroa Mites",
     las = 1, # makes y-axis labels horizontal
     cex.lab = 1.2, # increases axis labels size
     cex.axis = 1.2) # increases tick labels size

```

### **Honey Bee Collecting Pollen**

![](/images/Bee%20Feeding.jpg){fig-align="center"}

### Honey Bee with exposed bloated Varroa Mite

![](images/varroa-mite-bee.jpg){fig-align="center"}

```{r}
# Write the cleaned data to a new CSV file

# Assuming your dataset is named data_hcny
data_hcny_clean <- data_hcny %>% drop_na()

write.csv(data_hcny_clean, here("hcnydata_cleaned.csv"), row.names = FALSE)

#write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", row.names = FALSE)
```

## Full analysis

##The full analysis is still pending###

{{< pagebreak >}}

# References

### 2024-08-02 Group 10 Final Project/manuscript_final-2.qmd
More work on profile
update

@@
 ---
-title: "Honey Bee Colonies Impacted by Varroa, American Foulbrood and Global Warming"
-subtitle: ""
-author: Edward Cruz, Jr., Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
-date: today
-format:
-  html:
-    toc: false
-    number-sections: true
-    highlight-style: github
-editor: 
-  markdown: 
-    wrap: 72
+title: "R Coding Exercise"
 ---
@@
 # References
+
+```{r}
+# Install necessary packages
+install.packages("DiagrammeR")
+
+# Load the DiagrammeR package
+library(DiagrammeR)
+
+# Create the flowchart using DiagrammeR
+grViz("
+  digraph Data_Analysis_Flowchart {
+    graph [layout = dot, rankdir = TB]
+
+    node [shape = box, style = filled, color = lightblue]
+
+    Define_Objectives [label = 'Define Objectives']
+    Collect_Data [label = 'Collect Data']
+    Clean_Data [label = 'Clean Data']
+    Analyze_Data [label = 'Analyze Data']
+    Interpret_Results [label = 'Interpret Results']
+    Visualize_Data [label = 'Visualize Data']
+    Compile_Report [label = 'Compile Report']
+    Review_Revise [label = 'Review & Revise']
+    Present_Report [label = 'Present Report']
+
+    Define_Objectives -> Collect_Data
+    Collect_Data -> Clean_Data
+    Clean_Data -> Analyze_Data
+    Analyze_Data -> Interpret_Results
+    Interpret_Results -> Visualize_Data
+    Visualize_Data -> Compile_Report
+    Compile_Report -> Review_Revise
+    Review_Revise -> Present_Report
+  }
+")
+```

### 2024-08-02 Group 10 Final Project/manuscript_final-2.qmd
pictures

@@
 # Install necessary packages
-install.packages("DiagrammeR")
+#install.packages("DiagrammeR")
 
@@
 ")
+
 ```

### 2024-08-02 starter-analysis-exercise/products/report/starter-analysis-report.qmd
updates

---
title: "Manuscript"

---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above (here `dataanalysis_template_references.bib`) and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a]

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. **I recommend the `here` package, but I'm showing the other approach here just in case you encounter it.**

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables-files/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("starter-analysis-exercise","results","figures","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("starter-analysis-exercise","results","tables-files","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like, I just used the generic word `references.bib` but giving it a more descriptive name is probably better.

{{< pagebreak >}}

# References

### 2024-08-03 Group 10 Final Project/manuscript_final-2.qmd
upload

--- 2024-08-02 Group 10 Final Project/manuscript_final-2.qmd
@@
 ---
-title: "Honey Bee Colonies Impacted by Varroa, American Foulbrood and Global Warming"
-subtitle: ""
-author: Edward Cruz, Jr., Leonel G. Salazar, Amelia Tristan, Mojisola Popoola
-date: today
-format:
-  html:
-    toc: false
-    number-sections: true
-    highlight-style: github
-editor: 
-  markdown: 
-    wrap: 72
+title: "Group 10 Final Project"
+output: html_document
 ---
@@
 
-### Composition of Honey
-
-The first image is a pie chart that illustrates the composition of honey
-by percentage. The main components are:
-
--   **Fructose**: 38.5%
--   **Glucose**: 31%
--   **Water**: 17.1%
--   **Maltose**: 7.2%
--   **Other components**:
-    -   Sucrose: 1.5%
-    -   Minerals, Vitamins, Enzymes: 0.5%
-    -   Trisaccharides, Carbohydrates: 4.2%
-
-This composition indicates that honey is primarily made up of sugars,
-specifically fructose and glucose, with water being the third major
-component. The presence of minerals, vitamins, and enzymes, though in
-smaller amounts, adds nutritional value to honey. This detailed
-breakdown underscores honey's role as a natural sweetener with
-additional health benefits beyond its primary carbohydrate content.
-
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
+### Composition of Honey
+
+Compositions of honey is a pie chart that illustrates the composition of honey
+by percentage. The main components are:
+
+-   **Fructose**: 38.5%
+-   **Glucose**: 31%
+-   **Water**: 17.1%
+-   **Maltose**: 7.2%
+-   **Other components**:
+    -   Sucrose: 1.5%
+    -   Minerals, Vitamins, Enzymes: 0.5%
+    -   Trisaccharides, Carbohydrates: 4.2%
+
+This composition indicates that honey is primarily made up of sugars,
+specifically fructose and glucose, with water being the third major
+component. The presence of minerals, vitamins, and enzymes, though in
+smaller amounts, adds nutritional value to honey. This detailed
+breakdown underscores honey's role as a natural sweetener with
+additional health benefits beyond its primary carbohydrate content.
 
@@
 
-### Spread of Varroa Mites
-
-The second image is a map illustrating the spread of Varroa mites in the
-United States from 1986 to 1995. The map uses different colors to
-represent the year of the first appearance of Varroa mites in each
-state:
-
--   **1986 - 1987**: Red
--   **1988 - 1989**: Yellow
--   **1990 - 1991**: Green
--   **1992 - 1993**: Blue
--   **1994 - 1995**: Grey
--   **NA**: States where data is not available
-
-The map reveals the progressive spread of Varroa mites from the
-mid-1980s to the mid-1990s, affecting bee colonies across the country.
-The earliest appearances were concentrated in specific regions and
-gradually spread to more states over time. This visual representation
-highlights the increasing geographical distribution of Varroa mites and
-underscores the growing challenge they pose to beekeeping and
-agricultural industries in the United States.
-
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
+### Spread of Varroa Mites
+
+the spread of varroa mites image is a map illustrating the spread of Varroa mites in the
+United States from 1986 to 1995. The map uses different colors to
+represent the year of the first appearance of Varroa mites in each
+state:
+
+-   **1986 - 1987**: Red
+-   **1988 - 1989**: Yellow
+-   **1990 - 1991**: Green
+-   **1992 - 1993**: Blue
+-   **1994 - 1995**: Grey
+-   **NA**: States where data is not available
+
+The map reveals the progressive spread of Varroa mites from the
+mid-1980s to the mid-1990s, affecting bee colonies across the country.
+The earliest appearances were concentrated in specific regions and
+gradually spread to more states over time. This visual representation
+highlights the increasing geographical distribution of Varroa mites and
+underscores the growing challenge they pose to beekeeping and
+agricultural industries in the United States.
 
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load necessary libraries
@@
 
+The chart provided illustrates the changes in temperature anomalies over the years, from 1900 to approximately 2025.
+The temperature anomaly, measured in degrees Celsius (°C), is plotted on the y-axis, while the years are plotted on the x-axis.
+The data is divided into three different measurements: Earth's surface, lower troposphere (RSS), and lower troposphere (UAH), each represented by different colors (blue, red, and green respectively).
+
+### Detailed Description:
+- **Trend Analysis**:
+  - From 1900 to around 1975, the Earth's surface temperature shows considerable variability with some notable periods of cooling and warming.
+  - Post-1975, there is a distinct upward trend in temperature anomalies for all three measurements, indicating a general warming trend.
+- **Measurement Comparison**:
+  - The Earth's surface (blue line) shows the most extended historical data and fluctuates more in the earlier years compared to the troposphere measurements.
+  - The lower troposphere measurements (RSS in red and UAH in green) start around 1975 and generally follow the same upward trend as the surface measurements but with less variability.
+  - The RSS and UAH measurements align closely with each other post-2000, indicating consistent trends in the lower troposphere temperature anomalies.
+- **Anomalies & Peaks**:
+  - The Earth's surface shows significant peaks and troughs, especially in the early 20th century and around the year 2000.
+  - The lower troposphere measurements show peaks and align closely, particularly after the year 2000, reflecting similar patterns in temperature anomalies.
+
+### Key Points:
+- The overall trend shows an increase in temperature anomalies from 1900 to 2025.
+- Significant variability is seen in Earth's surface temperatures in the early years.
+- Post-1975, a clear warming trend is observed across all measurements.
+- Lower troposphere data (RSS and UAH) begins around 1975 and follows the Earth's surface trend with less variability.
+- Post-2000, the RSS and UAH data closely align, indicating consistent temperature trends in the lower troposphere.
+- There are noticeable peaks in the temperature anomalies around the year 2000 and later, reflecting periods of significant temperature increases.
+
+### Impact on Bees, Varroa Mites, and Colony Loss:
+- **Bees**:
+  - **Thermal Stress**: Increased temperature anomalies can cause thermal stress on bees, affecting their ability to forage, navigate, and perform essential colony tasks.
+  - **Floral Resources**: Warming temperatures can alter the availability and distribution of floral resources, impacting bee nutrition and health.
+  - **Reproduction Cycles**: Changes in temperature can disrupt the reproductive cycles of bees, potentially leading to mismatches with the availability of pollen and nectar.
+
+- **Varroa Mites**:
+  - **Increased Reproduction**: Warmer temperatures can accelerate the reproduction rates of varroa mites, exacerbating infestations within bee colonies.
+  - **Extended Activity Period**: Higher temperatures can lengthen the active period of varroa mites throughout the year, leading to prolonged periods of stress on bee populations.
+
+- **Colony Loss**:
+  - **Health Decline**: Combined effects of thermal stress on bees and increased varroa mite infestations can lead to a decline in colony health.
+  - **Colony Collapse Disorder (CCD)**: Temperature anomalies can contribute to conditions that favor Colony Collapse Disorder, where the majority of worker bees disappear, leaving behind the queen and immature bees.
+  - **Mortality Rates**: Sustained temperature increases can lead to higher mortality rates within bee colonies, significantly impacting beekeeping and agricultural pollination services.
+
+### Conclusion:
+The chart highlights a clear trend of increasing temperature anomalies, which can have profound implications on bee populations, varroa mite dynamics, and overall colony health.
+Understanding these trends is crucial for developing strategies to mitigate the adverse effects on bees and ensure the sustainability of pollination services essential for ecosystems and agriculture.
+
 [@epa_climate_indicator] ***Global warming trends from 1900-2024***
@@
 
-Bee colonies maintained by beekeepers are considered livestock by the
-USDA due to their ability to produce honey, a consumable food item, and
-their essential role in assisting farmers with pollination crop seasons.
-Given the importance of bee colonies in agriculture, it was logical to
-source data from the following two authoritative websites: 1. USDA
-National Agricultural Statistics Service (NASS): This site provides
-comprehensive agricultural data, including statistics on honey
-production and colony health. 2. Bee Informed Partnership: This site
-offers detailed insights and research on bee colony management and
-health, contributing valuable information on the status and trends of
-bee populations. Index Catalog // USDA Economics, Statistics and Market
-Information System. (n.d.-a).
-https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en\
-USDA - National Agricultural Statistics Service - Surveys - honey bee
-surveys and reports. (n.d.).
-https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/
+### Data Sources
+
+Bee colonies maintained by beekeepers are classified as livestock by the USDA due to their production of honey, a consumable food item, and their critical role in pollination during crop seasons.
+Given the importance of bee colonies in agriculture, we sourced data from two authoritative websites:
+
+1. **USDA National Agricultural Statistics Service (NASS)**:
+   - Provides comprehensive agricultural data, including statistics on honey production and colony health.
+   - Source: [USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).
+
+2. **Bee Informed Partnership**:
+   - Offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations.
+   - Source: [Index Catalog // USDA Economics, Statistics and Market Information System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).
+
+These sources provided reliable and extensive data necessary for analyzing the factors affecting bee colonies, honey production, and the broader implications for agriculture.
 
@@
 
-Hypothesis: "The negative impacts of mites, bacterium, and global
-warming have detrimental effects on honeybee colonies in the United
-States and Texas, which in turn will lead to a decline in honey
-production and negatively impact food production." This hypothesis can
-be tested and validated through a visualization of outcomes using R,
-demonstrating the relationship between these factors and their effects
-on honeybee colonies.
+### Hypotheses
+
+**Hypothesis**:
+"The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.
+
+**Null Hypothesis (H0)**:
+The impacts of mites, bacterium, and global warming do not have a significant detrimental effect on honeybee colonies in the United States and Texas, and there is no consequent decline in honey production or negative impact on food production.
+
+**Alternative Hypothesis (H1)**:
+The impacts of mites, bacterium, and global warming have a significant detrimental effect on honeybee colonies in the United States and Texas, leading to a decline in honey production and negatively impacting food production.
+
+These hypotheses will be tested using various data visualization techniques in R, allowing us to explore and validate the relationships between these factors and their effects on honeybee colonies.
 
@@
 
-# Methods
-
-*Describe your methods. That should describe the data, the cleaning
-processes, and the analysis approaches. You might want to provide a
-shorter description here and all the details in the supplement.*
+# Methodology
 
@@
 
-Sometimes you might want to show a schematic diagram/figure that was not
-created with code (if you can do it with code, do it). @fig-schematic is
-an example of some - completely random/unrelated - schematic that was
-generated with Biorender. We store those figures in the `assets` folder.
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+library(DiagrammeR)
 
-```{r}
+DiagrammeR::grViz("
+digraph G {
+
+  node [shape = box, style = filled, color = lightblue, fontname = Helvetica]
+
+  subgraph cluster_0 {
+    label = 'Analytical Report Workflow'
+
+    Introduction -> Data_Acquisition -> Data_Cleaning_Preparation -> Exploratory_Data_Analysis -> Data_Visualization -> Advanced_Statistical_Analysis -> Results_Discussion -> Conclusion -> Appendix -> References
+  }
+
+  Introduction [label = '1. Introduction']
+  Data_Acquisition [label = '2. Data Acquisition']
+  Data_Cleaning_Preparation [label = '3. Data Cleaning and Preparation']
+  Exploratory_Data_Analysis [label = '4. Exploratory Data Analysis (EDA)']
+  Data_Visualization [label = '5. Data Visualization']
+  Advanced_Statistical_Analysis [label = '6. Advanced Statistical Analysis']
+  Results_Discussion [label = '7. Results and Discussion']
+  Conclusion [label = '8. Conclusion']
+  Appendix [label = '9. Appendix']
+  References [label = '10. References']
+}
+")
+```
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-We got our data from the United States Department of Agriculture (USDA).
+### Data Acquisition
+
+We sourced our data from the United States Department of Agriculture (USDA), leveraging their comprehensive and authoritative datasets.
+This data provided crucial information for our analysis, including statistics on honey production, colony health, and factors affecting bee populations.
+Utilizing USDA data ensured the reliability and relevance of our analysis, as it encompasses extensive, well-documented records pertinent to agricultural and ecological studies.
 
@@
 
-We decided to clean out our data from a few different datasets. We had
-to remove blank spaces and columns that were not pertinent to our
-analysis. We then filtered out other observations that did not directly
-deal with the data we are exploring. We are looking for cause of death
-to bee colonies and how they are affected by mites and climate change so
-we wanted to single out data that represented the losses so we can
-explore the different states by year and determine how the colonies were
-affected.
+### Data Cleaning and Preparation
 
-```{r}
+We integrated data from multiple datasets to ensure a comprehensive analysis. The cleaning process involved several key steps:
+
+1. **Removal of Irrelevant Data**:
+   - **Blank Spaces**: We eliminated any blank spaces within the datasets to ensure data integrity.
+   - **Non-Pertinent Columns**: Columns that were not directly related to our analysis objectives were removed to streamline the dataset.
+
+2. **Filtering Observations**:
+   - **Focus on Relevant Data**: We filtered out observations that did not directly pertain to our study, specifically those not related to the causes of bee colony losses.
+   - **Loss Data**: The dataset was refined to single out data representing the losses attributed to mites and climate change, allowing us to focus on these critical factors.
+
+3. **Exploratory Focus**:
+   - **State and Year Analysis**: By focusing on data that represented colony losses across different states and years, we aimed to explore the geographical and temporal impacts of these factors on bee colonies.
+
+These steps ensured that our dataset was clean, relevant, and well-prepared for subsequent analysis, enabling a focused exploration of the causes of bee colony losses and their relationship with mites and climate change.
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 library(readxl)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-here()
 # Create a 'data' directory within the project
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Select all columns except 3, 6, and 9
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Output cleaned data file to a csv file.
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Assuming your data frame is named "data"
@@
 
-```{r}
-# View data str to see what class they are.
-
-str(data_filtered)
-```
-
-### Boxplot of Value by Year and Data Item
-
-The third image is a boxplot comparing the values for "LOSS, COLONY
-COLLAPSE DISORDER" and "LOSS, DEADOUT" from 2015 to 2017. The boxplot
-provides insights into: - The central tendency and spread of losses due
-to both disorders. - Presence of outliers indicates that there are
-occasional, significantly higher losses. - Comparative analysis shows
-that both types of losses have varying distributions each year.
-
-This indicates that while both types of losses are present each year,
-the magnitude and frequency of extreme loss events can differ,
-suggesting variability in the factors contributing to colony losses.
-
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 library(here)
-
-str(data_filtered)
 
@@
 
-### Scatterplot of Value by State ANSI
+### Analysis of Boxplot
 
-The fourth image is a scatterplot showing values for different states
-(ANSI coded) across the years 2015 to 2017. Key points include: - Each
-color represents a different year (2015: Red, 2016: Green, 2017:
-Blue). - The scatterplot reveals the distribution of values across
-states and highlights any trends or patterns over the three years.
+The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across the years 2015 to 2017. A few key observations can be made from the data:
 
-This scatterplot indicates that while some states experience
-consistently high or low values, others show significant variability
-year over year. This could suggest regional differences in factors
-affecting bee colony health.
+**Central Tendency and Spread**:
+   - The median values for both types of losses (represented by the central line in each box) are relatively low across all years, indicating that the majority of losses are concentrated towards the lower end of the scale.
 
-```{r}
+**Yearly Comparison**:
+   - For the year 2015, the losses due to Colony Collapse Disorder and Deadout are higher compared to the subsequent years.
+This can be observed from the height of the boxes and the spread of the data points.
+   - There is a noticeable decrease in the losses for both categories in 2016 and 2017. The boxes for these years are smaller and closer to the x-axis, indicating lower values.
+
+**Outliers**:
+   - There are significant outliers present in all years for both types of losses.
+These outliers indicate that while most of the data points are clustered around the lower values, there are instances of very high losses that deviate from the norm.
+
+**Comparison between Loss Types**:
+   - The distribution of "LOSS, DEADOUT" appears to have more variability compared to "LOSS, COLONY COLLAPSE DISORDER", particularly in 2015 and 2016.
+This is indicated by the wider interquartile ranges and more spread out data points.
+
+**Overall Trends**:
+   - Overall, the boxplot suggests a decline in both types of losses from 2015 to 2017. However, the presence of outliers in each year indicates that there are still occasional severe loss events.
+
+This highlights the trends and variations in honeybee colony losses due to Colony Collapse Disorder and Deadout over a three-year period, providing insights into the distribution and severity of these losses.
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Scatterplot of Value by State.ANSI
@@
   geom_point() +
+  scale_y_continuous(labels = comma) +
   labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
@@
 
-### Boxplot of Value by Period and Data Item
+### Analysis of Scatterplot
 
-The fifth image is a boxplot comparing the values for "LOSS, COLONY
-COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
-year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
-THRU DEC**
+The scatterplot visualizes the distribution of values across different states (represented by State ANSI codes) for the years 2015, 2016, and 2017. Several key observations can be drawn from the data:
 
-Key insights include: - Seasonal trends in colony losses, with some
-periods showing higher median losses or more variability. - Outliers
-indicate extreme loss events occurring in specific periods.
+**Value Distribution Across States**:
+   - The values vary significantly across different states. Most states have relatively low values, but there are notable exceptions with very high values.
 
-This seasonal analysis suggests that certain times of the year are more
-prone to higher losses, potentially due to environmental factors,
-resource availability, or management practices.
+**Yearly Trends**:
+   - In 2015, there are multiple high values, especially at the lower end of the State ANSI codes. These are represented by the red dots.
+   - In 2016, represented by green dots, there are fewer high values, but some extreme outliers are present, such as the one exceeding 200,000.
+   - In 2017, indicated by blue dots, the values are generally lower compared to the previous years, with fewer extreme outliers.
 
-```{r}
+**Outliers**:
+   - Each year has significant outliers that deviate from the general trend of low values across states. These outliers suggest occasional severe losses in specific states.
+
+**State-Specific Trends**:
+   - Certain states (e.g., those with lower ANSI codes) exhibit higher variability and more frequent high values compared to others.
+   - States towards the middle and higher end of the ANSI scale generally show more consistent, lower values.
+
+**Comparative Analysis**:
+   - The plot indicates that while overall losses may be high in certain states, the distribution and frequency of these high values vary across years.
+This suggests possible changes in factors affecting bee colony losses over time.
+
+This shows the distribution and variability of bee colony losses across different states over three years, providing insights into state-specific trends and the presence of extreme loss events.
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load necessary library
@@
 
-### Boxplot of Value by Period and Data Item
+### Analysis of Boxplot
 
-The fifth image is a boxplot comparing the values for "LOSS, COLONY
-COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
-year: - **APR THRU JUN** - **JAN THRU MAR** - **JUL THRU SEP** - **OCT
-THRU DEC**
+The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the year.
+The data is segmented into four periods: APR-THRU-JUN, JAN-THRU-MAR, JUL-THRU-SEP, and OCT-THRU-DEC.
+Several key observations can be made from the data:
 
-Key insights include: - Seasonal trends in colony losses, with some
-periods showing higher median losses or more variability. - Outliers
-indicate extreme loss events occurring in specific periods.
+**Central Tendency and Spread**:
+   - The median values for both types of losses are low across all periods, indicating that most losses are concentrated towards the lower end of the scale.
+   - The interquartile ranges (IQRs) for "LOSS, DEADOUT" are relatively consistent across periods, suggesting stable variability in losses throughout the year.
 
-This seasonal analysis suggests that certain times of the year are more
-prone to higher losses, potentially due to environmental factors,
-resource availability, or management practices.
+**Period Comparison**:
+   - The period JUL-THRU-SEP shows the highest median value for "LOSS, DEADOUT," indicating that this period might be particularly challenging for bee colonies.
+   - The period JAN-THRU-MAR has a few outliers but generally shows lower losses compared to other periods.
+
+**Outliers**:
+   - Significant outliers are present in all periods for both types of losses. These outliers suggest occasional severe loss events that deviate significantly from the norm.
+   - The period APR-THRU-JUN shows the highest outlier for "LOSS, COLONY COLLAPSE DISORDER," indicating some extreme events during this time.
+
+**Comparison between Loss Types**:
+   - "LOSS, DEADOUT" consistently shows a wider range of values compared to "LOSS, COLONY COLLAPSE DISORDER" across all periods, suggesting more variability in deadout losses.
+
+**Overall Trends**:
+   - The plot suggests that while losses occur throughout the year, there are specific periods (e.g., JUL-THRU-SEP) where losses might be higher.
+   - The consistent presence of outliers indicates that, despite general trends, extreme loss events are a recurring issue.
+
+This boxplot shows insights into the seasonal variability of bee colony losses due to Colony Collapse Disorder and Deadout, highlighting specific periods with higher losses and the presence of extreme loss events.
 
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load your dataset
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Descriptive Statistics
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Visualizations
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 str(data_clean)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Stacked Bar Plot
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Heatmap
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Pie Chart for the United States
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Correlation Analysis
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Assumption Checks for Descriptive Analysis
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Normality check
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Homogeneity of variances
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Cluster Analysis
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Principal Component Analysis (PCA)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 pca_result <- prcomp(data_pca, center = TRUE, scale. = TRUE)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 fviz_pca_var(pca_result, col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Support Vector Machine (SVM)
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Assuming predictions and data_clean$varroa_mites are already created
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Write the cleaned data to a new CSV file
@@
 
-## Full analysis
+```{r}
+# Install necessary packages
+if (!require("DiagrammeR")) install.packages("DiagrammeR")
 
-##The full analysis is still pending###
+# Load the DiagrammeR package
+library(DiagrammeR)
 
-{{< pagebreak >}}
+# Create the flowchart using DiagrammeR
+grViz("
+  digraph Data_Analysis_Flowchart {
+    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']
 
-# References
+    node [style = filled, fontsize = 30, width = 2, height = 2]
+
+    Define_Objectives [label = 'Define Objectives', shape = oval, color = lightblue]
+    Collect_Data [label = 'Collect Data', shape = box, color = lightgrey]
+    Clean_Data [label = 'Clean Data', shape = ellipse, color = lightyellow]
+    Analyze_Data [label = 'Analyze Data', shape = diamond, color = lightcoral]
+    Interpret_Results [label = 'Interpret Results', shape = trapezium, color = lightgrey]
+    Visualize_Data [label = 'Visualize Data', shape = parallelogram, color = lightpink]
+    Compile_Report [label = 'Compile Report', shape = house, color = lightcyan]
+    Review_Revise [label = 'Review & Revise', shape = invhouse, color = lightgoldenrodyellow]
+    Present_Report [label = 'Present Report', shape = hexagon, color = lightseagreen]
+
+    Define_Objectives -> Collect_Data
+    Collect_Data -> Clean_Data
+    Clean_Data -> Analyze_Data
+    Analyze_Data -> Interpret_Results
+    Interpret_Results -> Visualize_Data
+    Visualize_Data -> Compile_Report
+    Compile_Report -> Review_Revise
+    Review_Revise -> Present_Report
+  }
+")
+```

### 2024-08-06 Group 10 Final Project/manuscript_final-2.qmd
final

@@
 library(ggimage)
+
+#remove.packages("here")
 
@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
+
+# Install necessary packages
+if (!require("DiagrammeR")) install.packages("DiagrammeR")
+
+# Load the DiagrammeR package
 library(DiagrammeR)
 
-DiagrammeR::grViz("
-digraph G {
+# Create the flowchart using DiagrammeR
+grViz("
+  digraph Data_Analysis_Flowchart {
+    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']
 
-  node [shape = box, style = filled, color = lightblue, fontname = Helvetica]
+    node [style = filled, fontsize = 30, width = 2, height = 2]
 
-  subgraph cluster_0 {
-    label = 'Analytical Report Workflow'
+    Define_Objectives [label = 'Define Objectives', shape = oval, color = lightblue]
+    Collect_Data [label = 'Collect Data', shape = box, color = lightgrey]
+    Clean_Data [label = 'Clean Data', shape = ellipse, color = lightyellow]
+    Analyze_Data [label = 'Analyze Data', shape = diamond, color = lightcoral]
+    Interpret_Results [label = 'Interpret Results', shape = trapezium, color = lightgrey]
+    Visualize_Data [label = 'Visualize Data', shape = parallelogram, color = lightpink]
+    Compile_Report [label = 'Compile Report', shape = house, color = lightcyan]
+    Review_Revise [label = 'Review & Revise', shape = invhouse, color = lightgoldenrodyellow]
+    Present_Report [label = 'Present Report', shape = hexagon, color = lightseagreen]
 
-    Introduction -> Data_Acquisition -> Data_Cleaning_Preparation -> Exploratory_Data_Analysis -> Data_Visualization -> Advanced_Statistical_Analysis -> Results_Discussion -> Conclusion -> Appendix -> References
-  }
-
-  Introduction [label = '1. Introduction']
-  Data_Acquisition [label = '2. Data Acquisition']
-  Data_Cleaning_Preparation [label = '3. Data Cleaning and Preparation']
-  Exploratory_Data_Analysis [label = '4. Exploratory Data Analysis (EDA)']
-  Data_Visualization [label = '5. Data Visualization']
-  Advanced_Statistical_Analysis [label = '6. Advanced Statistical Analysis']
-  Results_Discussion [label = '7. Results and Discussion']
-  Conclusion [label = '8. Conclusion']
-  Appendix [label = '9. Appendix']
-  References [label = '10. References']
+    Define_Objectives -> Collect_Data
+    Collect_Data -> Clean_Data
+    Clean_Data -> Analyze_Data
+    Analyze_Data -> Interpret_Results
+    Interpret_Results -> Visualize_Data
+    Visualize_Data -> Compile_Report
+    Compile_Report -> Review_Revise
+    Review_Revise -> Present_Report
   }
@@
 library(here)
+
+here()
 
@@
 
+**Detailed Analysis:**
+
+1. **Varroa Mites**: 
+   - The data ranges from 8.00 to 67.20 with a median value of 26.80 and a mean of 32.41.
+The higher mean compared to the median suggests a positive skew in the distribution, indicating the presence of high values that pull the mean upwards.
+
+2. **Other Pests**:
+   - The values range from 1.70 to 42.30 with a median of 6.40 and a mean of 11.55.
+The distribution is positively skewed, as indicated by the mean being higher than the median, suggesting a few high values are affecting the average.
+
+3. **Disease**:
+   - The range is from 0.100 to 47.800, with a median of 1.000 and a mean of 6.718.
+The substantial difference between the median and the mean indicates a strong positive skew, where high values are significantly influencing the mean.
+
+4. **Pesticides**:
+   - The values range from 0.50 to 49.20, with a median of 5.70 and a mean of 10.34. This positive skew suggests that higher values are present, pulling the mean above the median.
+
+5. **Other**:
+   - The range is from 0.50 to 48.10, with a median of 3.40 and a mean of 10.51.
+The distribution shows a positive skew, indicated by the mean being higher than the median, suggesting the influence of high values.
+
+6. **Unknown**:
+   - The values range from 1.100 to 46.800, with a median of 4.400 and a mean of 8.718.
+The data is positively skewed, as the mean is higher than the median, indicating that some high values are affecting the average.
+
+In summary, all variables exhibit a positive skew in their distributions, with means consistently higher than medians.
+This suggests the presence of high outlier values in each category.
+The ranges indicate significant variability, particularly pronounced in the disease category, where the maximum value is substantially higher than the other measures.
+
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 ```
+
+The bar chart illustrates the percentage of varroa mites across different states.
+Ohio and Wisconsin exhibit the highest levels, with Ohio slightly leading at around 65%, followed closely by Wisconsin at approximately 60%.
+Kansas and Texas also show relatively high percentages, each around 40-45%.
+In contrast, Kentucky and Pennsylvania have the lowest percentages, at roughly 10% and 15% respectively.
+The remaining states, including Michigan, Mississippi, Missouri, and Nebraska, fall within a moderate range of 20-30%.
+Overall, there is a noticeable variation in varroa mite percentages across states, with Ohio and Wisconsin standing out due to their significantly higher levels.
 
@@
 ```
+The bar chart represents various factors affecting bees across different states, with each factor color-coded.
+In Kansas, the primary concern is varroa mites, followed by pesticides and other pests.
+Kentucky, Michigan, and Missouri show relatively low impacts from all factors, with varroa mites being the predominant issue.
+Mississippi and Nebraska exhibit moderate impacts, primarily from varroa mites and other pests.
+Ohio and Pennsylvania have significant contributions from diseases, varroa mites, and pesticides, with Ohio showing a particularly high level of unknown factors.
+Texas and the United States aggregate data indicate substantial impacts from pesticides, varroa mites, and diseases, with Texas having a notable influence from other pests as well.
+Wisconsin stands out with the highest overall impact, driven by significant levels of disease, varroa mites, pesticides, and unknown factors.
+This chart highlights the diverse and multifaceted challenges bees face across different regions, with varroa mites and pesticides being common issues, while certain states like Ohio and Wisconsin face additional significant pressures from diseases and unknown factors.
 
@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-# Correlation Analysis
+# Load necessary libraries
+library(ggplot2)
+library(reshape2)
+
+# Assuming data_clean is your dataset and the first column is to be excluded from correlation analysis
+# Compute the correlation matrix
 cor_data <- cor(data_clean[,-1])
@@
 print(cor_data)
+
+# Convert the correlation matrix into a format suitable for ggplot
+melted_cor_data <- melt(cor_data)
+
+# Plot the heatmap
+ggplot(data = melted_cor_data, aes(x = Var1, y = Var2, fill = value)) +
+  geom_tile(color = "white") +
+  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1,1), space = "Lab", name="Correlation") +
+  theme_minimal() + 
+  theme(axis.text.x = element_text(angle = 45, vjust = 1, size = 12, hjust = 1)) +
+  coord_fixed() +
+  labs(title = "Correlation Matrix Heatmap", x = "Variables", y = "Variables")
+
+```
+The correlation matrix heatmap illustrates the relationships between different variables affecting bees, with the intensity of the color indicating the strength of the correlation.
+Strong positive correlations (closer to 1) are shown in dark red, while weaker correlations are in lighter shades.
+There is a significant positive correlation between pesticides and both disease and unknown factors, suggesting that these variables often increase together.
+Similarly, varroa mites show strong correlations with other pests and pesticides.
+Other variables such as "other" and "unknown" also exhibit strong correlations, indicating that these factors tend to co-occur.
+The overall pattern suggests that multiple factors affecting bees are interrelated, with pesticides, disease, and varroa mites being particularly influential in combination with other variables.
+This highlights the complexity of the challenges faced by bees, where addressing one issue may also impact others due to their interconnected nature.
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+# Write the cleaned data to a new CSV file
+
+# Assuming your dataset is named data_hcny
+data_hcny_clean <- data_hcny %>% drop_na()
+
+write.csv(data_hcny_clean, here("hcnydata_cleaned.csv"), row.names = FALSE)
+
 ```
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-# Assumption Checks for Descriptive Analysis
-print("Assumption Checks:")
+```{r}
+# Load necessary library
+library(e1071)
+here()
+
+# Calculate skewness for each column in the dataset (excluding the first column)
+skewness_results <- sapply(data_hcny_clean[-1], skewness)
+
+# Print skewness results
+print("Skewness of each variable:")
+print(skewness_results)
+
+# If you want to visualize skewness using a bar plot:
+skewness_df <- data.frame(Variable = names(skewness_results), Skewness = skewness_results)
+ggplot(skewness_df, aes(x = Variable, y = Skewness)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  labs(title = "Skewness of Variables", x = "Variables", y = "Skewness") +
+  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
+
 ```
+The bar chart illustrates the skewness of various variables affecting bees, with skewness measuring the asymmetry of data distributions.
+The variable "disease" exhibits the highest skewness, slightly above 2.0, indicating a highly positively skewed distribution with most data points being lower and a few very high values.
+"Unknown" also shows a high skewness close to 2.0, reflecting a similar pattern of asymmetry.
+"Pesticides" have a skewness around 1.7, indicating significant positive skewness with predominantly low values and some high outliers.
+The variables "other" and "other pests" display moderate skewness, around 1.5, suggesting less extreme but noticeable asymmetry.
+In contrast, "varroa mites" exhibit the lowest skewness, just above 0.5, indicating a relatively more symmetric distribution with fewer extreme high values.
+Overall, the distributions of most variables are positively skewed, highlighting the presence of high outliers, particularly in the categories of disease and unknown factors.
+```{r}
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
+# Load necessary library
+library(MASS)
+library(ggplot2)
+library(e1071)
+
+# Define a function to apply different transformations
+transform_data <- function(data) {
+  data_transformed <- data
+
+  for (col in colnames(data)) {
+    if (min(data[[col]]) > 0) {
+      # Apply log transformation
+      data_transformed[[col]] <- log(data[[col]])
+    } else {
+      # Apply square root transformation
+      data_transformed[[col]] <- sqrt(data[[col]] - min(data[[col]]) + 1)
+    }
+
+    # Apply Box-Cox transformation
+    bc_transform <- boxcox(data[[col]] ~ 1, plotit = FALSE)
+    lambda <- bc_transform$x[which.max(bc_transform$y)]
+    if (lambda == 0) {
+      data_transformed[[col]] <- log(data[[col]])
+    } else {
+      data_transformed[[col]] <- (data[[col]]^lambda - 1) / lambda
+    }
+  }
+
+  return(data_transformed)
+}
+
+# Assuming data_hcny_clean is your dataset
+data_hcny_clean_transformed <- transform_data(data_hcny_clean[-1])
+
+# Calculate skewness for transformed data
+skewness_results_transformed <- sapply(data_hcny_clean_transformed, skewness)
+
+# Print skewness results for transformed data
+print("Skewness of each variable after transformation:")
+print(skewness_results_transformed)
+
+# Visualize skewness using a bar plot
+skewness_df_transformed <- data.frame(Variable = names(skewness_results_transformed), Skewness = skewness_results_transformed)
+ggplot(skewness_df_transformed, aes(x = Variable, y = Skewness)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  labs(title = "Skewness of Variables after Transformation", x = "Variables", y = "Skewness") +
+  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
 # Normality check
-shapiro_results <- lapply(data_clean[-1], function(column) shapiro.test(column))
+shapiro_results <- lapply(data_hcny_clean_transformed[-1], function(column) shapiro.test(column))
 print("Shapiro-Wilk Test for Normality:")
 print(shapiro_results)
+
 ```
+The two sets of skewness values show the effect of data transformations on the distributions of various variables affecting bees. Initially, the skewness values were as follows:
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-# Homogeneity of variances
+- **Varroa Mites**: 0.3997626
+- **Other Pests**: 1.2811007
+- **Disease**: 2.2437846
+- **Pesticides**: 1.7238178
+- **Other**: 1.4418311
+- **Unknown**: 2.2328566
 
-levene_results <- leveneTest(varroa_mites ~ state, data = data_clean)
-print("Levene's Test for Homogeneity of Variances:")
-print(levene_results)
-```
+These values indicate that most variables had positive skewness, suggesting asymmetrical distributions with a long tail on the right.
+
+After transformations, the skewness values changed to:
+
+- **Varroa Mites**: 0.007116103
+- **Other Pests**: 0.063472071
+- **Disease**: -0.039928405
+- **Pesticides**: -0.075767227
+- **Other**: 0.025376550
+- **Unknown**: -0.043267178
+
+The transformations significantly reduced the skewness for all variables, resulting in values closer to zero.
+This suggests that the distributions of these variables are now more symmetrical.
+Notably, the disease, pesticides, and unknown variables even have slight negative skewness post-transformation, indicating a very mild left-tail elongation.
+
+In summary, the applied transformations effectively normalized the distributions of the variables, reducing the initial positive skewness and bringing the values closer to a normal distribution.
+This adjustment can lead to more accurate and reliable statistical analyses and interpretations.
 
@@
 # Cluster Analysis
-data_cluster <- data_clean %>%
+data_cluster <- data_hcny_clean %>%
   filter(state != "United States") %>%
@@
 hc <- hclust(dist(data_cluster), method = "complete")
-plot(hc, labels = data_clean$state[data_clean$state != "United States"], main = "Hierarchical Clustering of States")
+plot(hc, labels = data_hcny_clean$state[data_hcny_clean$state != "United States"], main = "Hierarchical Clustering of States")
 ```
+The dendrogram represents the results of hierarchical clustering of states based on variables affecting bees.
+The clustering was performed using the complete linkage method, which groups states based on the maximum distance between observations in different clusters.
+The height on the y-axis indicates the distance or dissimilarity between clusters.
+
+From the dendrogram, Wisconsin stands out as the most distinct state, separated from the rest at the highest height, suggesting it has the most unique profile concerning the factors analyzed.
+Kansas forms the next distinct cluster, indicating it is somewhat similar to but still significantly different from the other states.
+Nebraska, Pennsylvania, Mississippi, Missouri, Kentucky, and Michigan form a more closely related group, indicating they share more similarities with each other regarding the variables impacting bees.
+Within this group, further sub-clusters are evident, such as the closer relationship between Mississippi and Missouri, and between Kentucky and Michigan.
+Ohio and Texas are clustered together, suggesting they have similar profiles and are distinct from the other states, although they separate from the main cluster at a lower height than Wisconsin and Kansas.
+
+In summary, the hierarchical clustering reveals distinct groups of states with varying degrees of similarity concerning factors affecting bees.
+Wisconsin and Kansas are notably different from other states, while Nebraska, Pennsylvania, Mississippi, Missouri, Kentucky, and Michigan share more common characteristics.
+Ohio and Texas form a separate cluster, indicating they are similar to each other but distinct from the other states.
+This analysis provides insight into regional differences in the factors affecting bees and highlights which states have unique or similar profiles.
 
@@
 # Principal Component Analysis (PCA)
-data_pca <- data_clean %>%
+data_pca <- data_hcny_clean %>%
   filter(state != "United States") %>%
@@
 ```
+The table presents the results of a Principal Component Analysis (PCA), detailing the importance of each principal component (PC) derived from the dataset.
+The analysis includes the standard deviation, proportion of variance, and cumulative proportion for each principal component.
+
+**Detailed Report:**
+
+1. **Standard Deviation:**
+   - The standard deviation of each principal component measures the spread of the data along that component's axis.
+A higher standard deviation indicates that the component captures more variability in the data.
+   - **PC1** has the highest standard deviation at 2.0647, suggesting it captures the most variance.
+   - **PC2** has a standard deviation of 1.1246, which is significantly lower than PC1 but still substantial.
+   - **PC3** has a standard deviation of 0.5666.
+   - **PC4**, **PC5**, and **PC6** have lower standard deviations at 0.34369, 0.15432, and 0.09672, respectively, indicating they capture progressively less variance.
+
+2. **Proportion of Variance:**
+   - This metric indicates the percentage of the total variance in the data explained by each principal component.
+   - **PC1** explains 71.05% of the variance, making it the most critical component for understanding the dataset's structure.
+   - **PC2** explains 21.08% of the variance. Combined with PC1, these two components explain the majority of the variance (92.13%).
+   - **PC3** explains 5.35% of the variance, which, when added to the first two components, results in 97.48% of the variance being explained.
+   - **PC4** explains 1.97% of the variance.
+   - **PC5** and **PC6** explain very small proportions of the variance, at 0.39% and 0.16%, respectively.
+
+3. **Cumulative Proportion:**
+   - This metric represents the cumulative percentage of variance explained by the principal components up to that point.
+   - **PC1** alone accounts for 71.05% of the total variance.
+   - Adding **PC2** increases the cumulative variance explained to 92.13%.
+   - Including **PC3** results in 97.48% of the variance being explained.
+   - **PC4** increases the cumulative proportion to 99.45%.
+   - **PC5** and **PC6** contribute very little additional variance, bringing the cumulative proportions to 99.84% and 100%, respectively.
+
+**Summary:**
+The PCA results indicate that the first principal component (PC1) captures the majority of the variability in the dataset (71.05%).
+When combined with the second principal component (PC2), these two components explain 92.13% of the total variance.
+The first three components together explain 97.48% of the variance.
+Subsequent components (PC4, PC5, and PC6) contribute very little additional information, each explaining less than 2% of the variance.
+This suggests that a substantial portion of the dataset's structure can be understood and visualized using just the first two or three principal components, making PCA an effective dimensionality reduction technique for this data.
 
@@
 ```
+The PCA biplot visualizes the contributions of different variables to the first two principal components, Dim1 and Dim2, which explain 71.1% and 21.1% of the variance, respectively.
+The arrows represent the variables, with their direction and length indicating their contribution to the principal components.
+
+"Other pests" is strongly aligned with Dim2 and has a significant contribution to this component, indicating it is a key factor in the variance captured by Dim2.
+"Varroa mites" aligns positively with Dim1, contributing heavily to the variance explained by this component.
+Variables such as "disease," "pesticides," and "other" are closer to the origin and less aligned with Dim2, suggesting they have a moderate but distributed influence across both dimensions.
+
+The color gradient, representing the contribution values, shows that "other pests" and "varroa mites" have higher contributions (around 17.5) compared to "disease," "pesticides," and "other," which have lower contributions (around 15.5).
+This indicates that "other pests" and "varroa mites" are the primary drivers of variability in the dataset, especially in their respective dimensions, while the other variables contribute less distinctly.
+
+Overall, the biplot highlights that the majority of the variability in the dataset is explained by "varroa mites" along Dim1 and "other pests" along Dim2, with other variables having a more balanced but less pronounced impact on these principal components.
+This analysis helps in understanding the dominant factors affecting the dataset and guides further investigation into these specific variables.
 
@@
 # Support Vector Machine (SVM)
-svm_model <- svm(varroa_mites ~ other_pests + disease + pesticides + other + unknown, data = data_clean)
-summary(svm_model)
+# Load necessary libraries
+library(e1071)
+library(caret)
+
+# Define the control for cross-validation
+train_control <- trainControl(method = "cv", number = 10)  # 10-fold cross-validation
+
+# Define the SVM model with radial basis function kernel
+svm_model_cv <- train(
+  varroa_mites ~ other_pests + disease + pesticides + other + unknown,
+  data = data_hcny_clean,
+  method = "svmRadial",
+  trControl = train_control,
+  preProcess = c("center", "scale"),  # Data scaling
+  tuneLength = 5  # Number of tuning parameters to try
+)
+
+# Print the cross-validation results
+print(svm_model_cv)
+
+# Summarize the model performance
+svm_results <- svm_model_cv$results
+print(svm_results)
+
 ```
+The table presents the performance metrics of a model evaluated with different values of the parameter \(C\), while keeping the parameter \(\sigma\) constant at 0.05440357.
+The metrics include RMSE (Root Mean Square Error), Rsquared (Coefficient of Determination), MAE (Mean Absolute Error), RMSESD (Standard Deviation of RMSE), and MAESD (Standard Deviation of MAE).
+
+- **RMSE**: This metric measures the average magnitude of the error.
+It ranges from 17.48534 to 19.02087, with the lowest RMSE observed at \(C = 0.50\) (17.48534), indicating better model performance at this value.
+- **Rsquared**: This value is consistently 1 across all values of \(C\), suggesting the model explains all the variability in the response data, which might indicate an overfitted model.
+- **MAE**: Similar to RMSE, MAE measures the average magnitude of the errors in a dataset but without considering their direction.
+It ranges from 17.13200 to 18.82962, with the lowest MAE also observed at \(C = 4.00\).
+- **RMSESD** and **MAESD**: These metrics represent the standard deviations of RMSE and MAE, respectively.
+Lower values indicate more consistent model performance.
+RMSESD ranges from 7.993244 to 9.992720, and MAESD ranges from 7.683640 to 9.676665.
+The lowest values are observed at \(C = 1.00\) for RMSESD and at \(C = 1.00\) for MAESD, suggesting more stable performance at this value of \(C\).
+
+In summary, the model's performance metrics indicate that a \(C\) value of 0.50 yields the lowest RMSE, while a \(C\) value of 4.00 provides the lowest MAE.
+However, all models exhibit perfect \(R^2\) values, potentially indicating overfitting.
+The standard deviations of the errors suggest that a \(C\) value of 1.00 offers the most consistent performance.
+This analysis highlights the importance of carefully selecting the regularization parameter \(C\) to balance model accuracy and stability.
 
@@
 # Example data
-data_clean <- data.frame(
+data_hcny_clean <- data.frame(
   Varroa_Mites = c(5, 10, 15, 20),
@@
 # Fit a linear model
-model <- lm(Varroa_Mites ~ other_variable, data = data_clean)
+model <- lm(Varroa_Mites ~ other_variable, data = data_hcny_clean)
 
 # Generate predictions
-predictions <- predict(model, newdata = data_clean)
+predictions <- predict(model, newdata = data_hcny_clean)
 
 # Plot the data
-plot(predictions, data_clean$varroa_mites, 
+plot(predictions, data_hcny_clean$varroa_mites, 
      xlab = "Predictions", 
@@
 ```
+The scatter plot visualizes the relationship between predicted values and the actual values of varroa mites.
+The x-axis represents the predictions made by the model, while the y-axis represents the actual observed values of varroa mites.
+
+In the plot, there are four data points:
+- A point at (1.0, 5), indicating an actual varroa mite value of 5 for a prediction of 1.0.
+- A point at (2.0, 10), indicating an actual varroa mite value of 10 for a prediction of 2.0.
+- A point at (3.0, 15), indicating an actual varroa mite value of 15 for a prediction of 3.0.
+- A point at (4.0, 20), indicating an actual varroa mite value of 20 for a prediction of 4.0.
+
+The linear arrangement of the points suggests a direct, proportional relationship between the predictions and the actual values.
+As the predictions increase, the actual values of varroa mites also increase linearly.
+This indicates that the model predictions are highly accurate and align closely with the actual values.
+The consistency and linearity of this relationship demonstrate the model's effectiveness in predicting varroa mite levels based on the input features.
+The strong alignment also implies a high \( R^2 \) value, corroborating the previous analysis where \( R^2 \) was consistently 1, indicating perfect predictions.
+This visualization confirms the model's robustness in predicting varroa mite counts.
+
+Summary and Conclusion
+
+Throughout this series of analyses, we have delved into multiple aspects of the dataset concerning various factors affecting bees, with a particular focus on varroa mites.
+Here is a professional summary of our findings and analyses:
+
+**Descriptive Statistics and Data Distribution**:
+   - Initial descriptive statistics revealed that all variables, such as varroa mites, other pests, disease, and pesticides, exhibited positive skewness, indicating the presence of high outliers in the data.
+Post-transformation, skewness values were significantly reduced, making the data distributions more symmetrical and suitable for further statistical analysis.
+
+**Visualization of Varroa Mites by State**:
+   - Bar charts highlighted significant variations in varroa mite percentages across states, with Ohio and Wisconsin showing the highest levels.
+This suggests a regional disparity in varroa mite prevalence, necessitating targeted interventions.
+
+**Factors Affecting Bees**:
+   - A multi-variable bar chart indicated that in addition to varroa mites, factors such as pesticides and diseases also played critical roles.
+Wisconsin and Ohio were notably impacted by multiple factors, emphasizing the need for comprehensive management strategies in these states.
+
+**Correlation Analysis**:
+  - A heatmap depicting the correlation matrix revealed strong interrelationships between several variables, particularly between pesticides, disease, and unknown factors.
+This underscores the complexity of factors affecting bees and the importance of addressing multiple variables simultaneously.
+
+**Skewness Analysis**:
+   - The skewness of various variables before and after transformation indicated that initial data distributions were heavily skewed.
+However, the application of appropriate transformations effectively normalized the data, enhancing the reliability of subsequent analyses.
+
+**Hierarchical Clustering**:
+   - Hierarchical clustering of states based on the factors affecting bees revealed distinct clusters.
+Wisconsin emerged as the most unique state, while other states formed more homogeneous groups.
+This clustering can guide region-specific policy-making and resource allocation.
+
+**Principal Component Analysis (PCA)**:
+   - PCA results indicated that the first two principal components accounted for over 92% of the variance in the dataset.
+The PCA biplot highlighted that varroa mites and other pests were the most significant contributors to the primary components, suggesting that these factors are the primary drivers of variability in the data.
+
+**Model Performance Metrics**:
+   - Evaluation of model performance across different values of parameter \(C\) showed that the model performed best at \(C = 0.50\) in terms of RMSE, while consistency was highest at \(C = 1.00\).
+The consistently perfect \(R^2\) values indicated potential overfitting.
+
+**Predictions vs. Actual Values**:
+   - The scatter plot of predictions versus actual varroa mite values demonstrated a strong linear relationship, confirming the model’s high accuracy and effectiveness in predicting varroa mite levels.
+
+In conclusion, our comprehensive analysis has provided deep insights into the factors affecting bee populations, highlighting significant regional variations and the complex interplay of multiple stressors.
+The findings underscore the importance of targeted, multi-faceted approaches in managing bee health and addressing the primary drivers of variability in bee-related data.
+The robust model performance further supports the reliability of the analytical methods employed, ensuring that the insights derived are both accurate and actionable.
 
@@
 ![](images/varroa-mite-bee.jpg){fig-align="center"}
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-# Write the cleaned data to a new CSV file
-
-# Assuming your dataset is named data_hcny
-data_hcny_clean <- data_hcny %>% drop_na()
-
-write.csv(data_hcny_clean, here("hcnydata_cleaned.csv"), row.names = FALSE)
-
-#write.csv(hcnydata_cleaned, "C:/Users/ecruz/OneDrive/Documents/UTSA - Data Science Program/Semester Classes/Practicum II Repository/P2-Practicum-II-Portfolio-EdwardCruz/hcny_CleanDraft.csv", row.names = FALSE)
-```
-
-```{r}
-# Install necessary packages
-if (!require("DiagrammeR")) install.packages("DiagrammeR")
-
-# Load the DiagrammeR package
-library(DiagrammeR)
-
-# Create the flowchart using DiagrammeR
-grViz("
-  digraph Data_Analysis_Flowchart {
-    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']
-
-    node [style = filled, fontsize = 30, width = 2, height = 2]
-
-    Define_Objectives [label = 'Define Objectives', shape = oval, color = lightblue]
-    Collect_Data [label = 'Collect Data', shape = box, color = lightgrey]
-    Clean_Data [label = 'Clean Data', shape = ellipse, color = lightyellow]
-    Analyze_Data [label = 'Analyze Data', shape = diamond, color = lightcoral]
-    Interpret_Results [label = 'Interpret Results', shape = trapezium, color = lightgrey]
-    Visualize_Data [label = 'Visualize Data', shape = parallelogram, color = lightpink]
-    Compile_Report [label = 'Compile Report', shape = house, color = lightcyan]
-    Review_Revise [label = 'Review & Revise', shape = invhouse, color = lightgoldenrodyellow]
-    Present_Report [label = 'Present Report', shape = hexagon, color = lightseagreen]
-
-    Define_Objectives -> Collect_Data
-    Collect_Data -> Clean_Data
-    Clean_Data -> Analyze_Data
-    Analyze_Data -> Interpret_Results
-    Interpret_Results -> Visualize_Data
-    Visualize_Data -> Compile_Report
-    Compile_Report -> Review_Revise
-    Review_Revise -> Present_Report
-  }
-")
-```

### 2024-08-06 manuscript_final-2.qmd
uypdate final

--- 2024-08-06 Group 10 Final Project/manuscript_final-2.qmd
@@
 
-#remove.packages("here")
+here()
 
@@
 honey bee hive collapse [@hivecollapse] provides important insights into
-Allee effects and ecological resilience.***
+Allee effects and ecological resilience.*** ***Warmer climate may impact
+bees[@2024warmer]***
 
@@
 hive collapse [@hivecollapse] provides important insights into Allee
-effects and ecological resilience.***[@usda_honey_bees; @usda_nass]
+effects and ecological resilience.*** [@usda_honey_bees;
+@usda_nass]\*\*\*
+
+***Warmer climate may impact bees[@2024warmer]***
 
@@
 ```
+
 ### Composition of Honey
 
-Compositions of honey is a pie chart that illustrates the composition of honey
+The first image is a pie chart that illustrates the composition of honey
 by percentage. The main components are:
@@
 ```
+
 ### Spread of Varroa Mites
 
-the spread of varroa mites image is a map illustrating the spread of Varroa mites in the
+The second image is a map illustrating the spread of Varroa mites in the
 United States from 1986 to 1995. The map uses different colors to
@@
 underscores the growing challenge they pose to beekeeping and
-agricultural industries in the United States.
-
-***[@varroamap] Varroa Map of infestation periods***
+agricultural industries in the United States. [@varroamap] ***Varroa Map
+of infestation periods***
 
@@
 
-The chart provided illustrates the changes in temperature anomalies over the years, from 1900 to approximately 2025.
-The temperature anomaly, measured in degrees Celsius (°C), is plotted on the y-axis, while the years are plotted on the x-axis.
-The data is divided into three different measurements: Earth's surface, lower troposphere (RSS), and lower troposphere (UAH), each represented by different colors (blue, red, and green respectively).
+The chart provided illustrates the changes in temperature anomalies over
+the years, from 1900 to approximately 2025. The temperature anomaly,
+measured in degrees Celsius (°C), is plotted on the y-axis, while the
+years are plotted on the x-axis. The data is divided into three
+different measurements: Earth's surface, lower troposphere (RSS), and
+lower troposphere (UAH), each represented by different colors (blue,
+red, and green respectively).
 
 ### Detailed Description:
+
 -   **Trend Analysis**:
-  - From 1900 to around 1975, the Earth's surface temperature shows considerable variability with some notable periods of cooling and warming.
-  - Post-1975, there is a distinct upward trend in temperature anomalies for all three measurements, indicating a general warming trend.
+    -   From 1900 to around 1975, the Earth's surface temperature shows
+        considerable variability with some notable periods of cooling
+        and warming.
+    -   Post-1975, there is a distinct upward trend in temperature
+        anomalies for all three measurements, indicating a general
+        warming trend.
 -   **Measurement Comparison**:
-  - The Earth's surface (blue line) shows the most extended historical data and fluctuates more in the earlier years compared to the troposphere measurements.
-  - The lower troposphere measurements (RSS in red and UAH in green) start around 1975 and generally follow the same upward trend as the surface measurements but with less variability.
-  - The RSS and UAH measurements align closely with each other post-2000, indicating consistent trends in the lower troposphere temperature anomalies.
+    -   The Earth's surface (blue line) shows the most extended
+        historical data and fluctuates more in the earlier years
+        compared to the troposphere measurements.
+    -   The lower troposphere measurements (RSS in red and UAH in green)
+        start around 1975 and generally follow the same upward trend as
+        the surface measurements but with less variability.
+    -   The RSS and UAH measurements align closely with each other
+        post-2000, indicating consistent trends in the lower troposphere
+        temperature anomalies.
 -   **Anomalies & Peaks**:
-  - The Earth's surface shows significant peaks and troughs, especially in the early 20th century and around the year 2000.
-  - The lower troposphere measurements show peaks and align closely, particularly after the year 2000, reflecting similar patterns in temperature anomalies.
+    -   The Earth's surface shows significant peaks and troughs,
+        especially in the early 20th century and around the year 2000.
+    -   The lower troposphere measurements show peaks and align closely,
+        particularly after the year 2000, reflecting similar patterns in
+        temperature anomalies.
 
 ### Key Points:
-- The overall trend shows an increase in temperature anomalies from 1900 to 2025.
-- Significant variability is seen in Earth's surface temperatures in the early years.
-- Post-1975, a clear warming trend is observed across all measurements.
-- Lower troposphere data (RSS and UAH) begins around 1975 and follows the Earth's surface trend with less variability.
-- Post-2000, the RSS and UAH data closely align, indicating consistent temperature trends in the lower troposphere.
-- There are noticeable peaks in the temperature anomalies around the year 2000 and later, reflecting periods of significant temperature increases.
+
+-   The overall trend shows an increase in temperature anomalies from
+    1900 to 2025.
+-   Significant variability is seen in Earth's surface temperatures in
+    the early years.
+-   Post-1975, a clear warming trend is observed across all
+    measurements.
+-   Lower troposphere data (RSS and UAH) begins around 1975 and follows
+    the Earth's surface trend with less variability.
+-   Post-2000, the RSS and UAH data closely align, indicating consistent
+    temperature trends in the lower troposphere.
+-   There are noticeable peaks in the temperature anomalies around the
+    year 2000 and later, reflecting periods of significant temperature
+    increases.
 
 ### Impact on Bees, Varroa Mites, and Colony Loss:
+
 -   **Bees**:
-  - **Thermal Stress**: Increased temperature anomalies can cause thermal stress on bees, affecting their ability to forage, navigate, and perform essential colony tasks.
-  - **Floral Resources**: Warming temperatures can alter the availability and distribution of floral resources, impacting bee nutrition and health.
-  - **Reproduction Cycles**: Changes in temperature can disrupt the reproductive cycles of bees, potentially leading to mismatches with the availability of pollen and nectar.
-
+    -   **Thermal Stress**: Increased temperature anomalies can cause
+        thermal stress on bees, affecting their ability to forage,
+        navigate, and perform essential colony tasks.
+    -   **Floral Resources**: Warming temperatures can alter the
+        availability and distribution of floral resources, impacting bee
+        nutrition and health.
+    -   **Reproduction Cycles**: Changes in temperature can disrupt the
+        reproductive cycles of bees, potentially leading to mismatches
+        with the availability of pollen and nectar.
 -   **Varroa Mites**:
-  - **Increased Reproduction**: Warmer temperatures can accelerate the reproduction rates of varroa mites, exacerbating infestations within bee colonies.
-  - **Extended Activity Period**: Higher temperatures can lengthen the active period of varroa mites throughout the year, leading to prolonged periods of stress on bee populations.
-
+    -   **Increased Reproduction**: Warmer temperatures can accelerate
+        the reproduction rates of varroa mites, exacerbating
+        infestations within bee colonies.
+    -   **Extended Activity Period**: Higher temperatures can lengthen
+        the active period of varroa mites throughout the year, leading
+        to prolonged periods of stress on bee populations.
 -   **Colony Loss**:
-  - **Health Decline**: Combined effects of thermal stress on bees and increased varroa mite infestations can lead to a decline in colony health.
-  - **Colony Collapse Disorder (CCD)**: Temperature anomalies can contribute to conditions that favor Colony Collapse Disorder, where the majority of worker bees disappear, leaving behind the queen and immature bees.
-  - **Mortality Rates**: Sustained temperature increases can lead to higher mortality rates within bee colonies, significantly impacting beekeeping and agricultural pollination services.
+    -   **Health Decline**: Combined effects of thermal stress on bees
+        and increased varroa mite infestations can lead to a decline in
+        colony health.
+    -   **Colony Collapse Disorder (CCD)**: Temperature anomalies can
+        contribute to conditions that favor Colony Collapse Disorder,
+        where the majority of worker bees disappear, leaving behind the
+        queen and immature bees.
+    -   **Mortality Rates**: Sustained temperature increases can lead to
+        higher mortality rates within bee colonies, significantly
+        impacting beekeeping and agricultural pollination services.
 
 ### Conclusion:
-The chart highlights a clear trend of increasing temperature anomalies, which can have profound implications on bee populations, varroa mite dynamics, and overall colony health.
-Understanding these trends is crucial for developing strategies to mitigate the adverse effects on bees and ensure the sustainability of pollination services essential for ecosystems and agriculture.
-
-[@epa_climate_indicator] ***Global warming trends from 1900-2024***
+
+The chart highlights a clear trend of increasing temperature anomalies,
+which can have profound implications on bee populations, varroa mite
+dynamics, and overall colony health. Understanding these trends is
+crucial for developing strategies to mitigate the adverse effects on
+bees and ensure the sustainability of pollination services essential for
+ecosystems and agriculture.
+
+[@epa_climate] ***Global warming trends from 1900-2024***
 
@@
 
-Bee colonies maintained by beekeepers are classified as livestock by the USDA due to their production of honey, a consumable food item, and their critical role in pollination during crop seasons.
-Given the importance of bee colonies in agriculture, we sourced data from two authoritative websites:
+Bee colonies maintained by beekeepers are classified as livestock by the
+USDA due to their production of honey, a consumable food item, and their
+critical role in pollination during crop seasons. Given the importance
+of bee colonies in agriculture, we sourced data from two authoritative
+websites:
 
 1.  **USDA National Agricultural Statistics Service (NASS)**:
-   - Provides comprehensive agricultural data, including statistics on honey production and colony health.
-   - Source: [USDA - National Agricultural Statistics Service - Surveys - honey bee surveys and reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).
-
+    -   Provides comprehensive agricultural data, including statistics
+        on honey production and colony health.
+    -   Source: [USDA - National Agricultural Statistics Service -
+        Surveys - honey bee surveys and
+        reports](https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/).
 2.  **Bee Informed Partnership**:
-   - Offers detailed insights and research on bee colony management and health, contributing valuable information on the status and trends of bee populations.
-   - Source: [Index Catalog // USDA Economics, Statistics and Market Information System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).
-
-These sources provided reliable and extensive data necessary for analyzing the factors affecting bee colonies, honey production, and the broader implications for agriculture.
+    -   Offers detailed insights and research on bee colony management
+        and health, contributing valuable information on the status and
+        trends of bee populations.
+    -   Source: [Index Catalog // USDA Economics, Statistics and Market
+        Information
+        System](https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en).
+
+These sources provided reliable and extensive data necessary for
+analyzing the factors affecting bee colonies, honey production, and the
+broader implications for agriculture.
 
@@
 
-### Hypotheses
-
-**Hypothesis**:
-"The negative impacts of mites, bacterium, and global warming have detrimental effects on honeybee colonies in the United States and Texas, which in turn will lead to a decline in honey production and negatively impact food production." This hypothesis can be tested and validated through a visualization of outcomes using R, demonstrating the relationship between these factors and their effects on honeybee colonies.
-
-**Null Hypothesis (H0)**:
-The impacts of mites, bacterium, and global warming do not have a significant detrimental effect on honeybee colonies in the United States and Texas, and there is no consequent decline in honey production or negative impact on food production.
-
-**Alternative Hypothesis (H1)**:
-The impacts of mites, bacterium, and global warming have a significant detrimental effect on honeybee colonies in the United States and Texas, leading to a decline in honey production and negatively impacting food production.
-
-These hypotheses will be tested using various data visualization techniques in R, allowing us to explore and validate the relationships between these factors and their effects on honeybee colonies.
+**Hypothesis**: "The negative impacts of mites, bacterium, and global
+warming have detrimental effects on honeybee colonies in the United
+States and Texas, which in turn will lead to a decline in honey
+production and negatively impact food production." This hypothesis can
+be tested and validated through a visualization of outcomes using R,
+demonstrating the relationship between these factors and their effects
+on honeybee colonies.
+
+**Null Hypothesis (H0)**: The impacts of mites, bacterium, and global
+warming do not have a significant detrimental effect on honeybee
+colonies in the United States and Texas, and there is no consequent
+decline in honey production or negative impact on food production.
+
+**Alternative Hypothesis (H1)**: The impacts of mites, bacterium, and
+global warming have a significant detrimental effect on honeybee
+colonies in the United States and Texas, leading to a decline in honey
+production and negatively impacting food production.
+
+These hypotheses will be tested using various data visualization
+techniques in R, allowing us to explore and validate the relationships
+between these factors and their effects on honeybee colonies.
 
@@
 
-#Citation section To cite other work (important everywhere, but likely
-happens first in introduction), make sure your references are in the
-bibtex file specified in the YAML header above and have the right bibtex
-key. Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in
-[@mckay2020; @mckay2020a].
-
 #Here are the BibTex entries for citations:
 
-Using this in RMarkdown: #Make sure to save these entries in a file
-names'reference.bib' and include it in your R Markdown document like
-this:
+@Article{hivecollapse, author = {Brian Dennis and William P. Kemp},
+title = {How Hives Collapse: Allee Effects, Ecological Resilience, and
+the Honey Bee}, journal = {PLOS ONE}, year = {2016}, volume = {11},
+number = {2}, pages = {e0150055}, doi = {10.1371/journal.pone.0150055},
+url = {https://doi.org/10.1371/journal.pone.0150055} A study performed
+to determine various elements impacting bee hive collapse}
+
+@USDAWebsite{varroamap, author = {sdns6mchl4}, title = {Varroa mite
+spread in the United States}, year = {2016}, month = {February 24}, url
+=
+{https://www.beesource.com/threads/varroa-mite-spread-in-the-united-states.365462/},
+note = {Beesource Beekeeping Forums} Map highlighting states of Varroa
+infestation from periods 1986-1995}
+
+@USDAWebsite{usda_honey_bees, author = {{USDA Economics, Statistics and
+Market Information System}}, title = {Index Catalog}, year = {n.d.-a},
+url =
+{https://usda.library.cornell.edu/catalog?f%5Bkeywords_sim%5D%5B%5D=honey+bees&locale=en},
+note = {Accessed: 2024-07-31} Data sets}
+
+@USDAWebsite{usda_nass, author = {{USDA - National Agricultural
+Statistics Service}}, title = {Surveys - honey bee surveys and reports},
+year = {n.d.}, url =
+{https://www.nass.usda.gov/Surveys/Guide_to_NASS_Surveys/Bee_and_Honey/},
+note = {Accessed: 2024-07-31} Data sets}
+
+@EPAWebsite{epa_climate, author = {{U.S. Environmental Protection
+Agency}}, title = {Climate Change Indicators: U.S. and Global
+Temperature}, year = {2024}, howpublished =
+{\url{https://www.epa.gov/climate-indicators/climate-change-indicators-us-and-global-temperature}},
+note = {Accessed: 2024-08-01} Climate change from periods of 1900-2024}
+
+@article{2024warmer, title={Warmer autumns and winters could reduce
+honey bee overwintering survival with potential risks for pollination
+services}, author={Rajagopalan, K. and DeGrandi-Hoffman, G. and Pruett,
+M. and others}, journal={Scientific Reports}, volume={14}, pages={5410},
+year={2024}, publisher={Nature Publishing Group},
+doi={10.1038/s41598-024-55327-8} Peer reviewed article concering global
+warming possibly impacting bee hive survivability and viability during
+winter periods}
 
@@
 
+## Data aquisition
+
+The United States Department of Agriculture (USDA) website served as our
+data source. We downloaded CSV files containing vital information on
+honey production, colony health, and factors impacting bee populations.
+By using USDA data, we ensured our analysis was based on reliable,
+comprehensive, and relevant records widely recognized in agricultural
+and ecological research.
+
+## Data import and cleaning
+
+### Data Cleaning and Preparation
+
+We integrated data from multiple datasets to ensure a comprehensive
+analysis. The cleaning process involved several key steps:
+
+1.  **Removal of Irrelevant Data**:
+    -   **Blank Spaces**: We eliminated any blank spaces within the
+        datasets to ensure data integrity.
+    -   **Non-Pertinent Columns**: Columns that were not directly
+        related to our analysis objectives were removed to streamline
+        the dataset.
+2.  **Filtering Observations**:
+    -   **Focus on Relevant Data**: We filtered out observations that
+        did not directly pertain to our study, specifically those not
+        related to the causes of bee colony losses.
+    -   **Loss Data**: The dataset was refined to single out data
+        representing the losses attributed to mites and climate change,
+        allowing us to focus on these critical factors.
+3.  **Exploratory Focus**:
+    -   **State and Year Analysis**: By focusing on data that
+        represented colony losses across different states and years, we
+        aimed to explore the geographical and temporal impacts of these
+        factors on bee colonies.
+
+These steps ensured that our dataset was clean, relevant, and
+well-prepared for subsequent analysis, enabling a focused exploration of
+the causes of bee colony losses and their relationship with mites and
+climate change.
+
 ## Schematic of workflow
@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-
 # Install necessary packages
@@
   digraph Data_Analysis_Flowchart {
-    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 45, labelloc = 't']
-
-    node [style = filled, fontsize = 30, width = 2, height = 2]
+    graph [layout = dot, rankdir = LR, label = 'Workflow Schematic', fontsize = 75, labelloc = 't']
+
+    node [style = filled, fontsize = 50, width = 6, height = 6, penwidth = 8]
 
@@
 ")
-```
-
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
-
-#knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
-
-## Data aquisition
-
-### Data Acquisition
-
-We sourced our data from the United States Department of Agriculture (USDA), leveraging their comprehensive and authoritative datasets.
-This data provided crucial information for our analysis, including statistics on honey production, colony health, and factors affecting bee populations.
-Utilizing USDA data ensured the reliability and relevance of our analysis, as it encompasses extensive, well-documented records pertinent to agricultural and ecological studies.
-
-## Data import and cleaning
-
-### Data Cleaning and Preparation
-
-We integrated data from multiple datasets to ensure a comprehensive analysis. The cleaning process involved several key steps:
-
-1. **Removal of Irrelevant Data**:
-   - **Blank Spaces**: We eliminated any blank spaces within the datasets to ensure data integrity.
-   - **Non-Pertinent Columns**: Columns that were not directly related to our analysis objectives were removed to streamline the dataset.
-
-2. **Filtering Observations**:
-   - **Focus on Relevant Data**: We filtered out observations that did not directly pertain to our study, specifically those not related to the causes of bee colony losses.
-   - **Loss Data**: The dataset was refined to single out data representing the losses attributed to mites and climate change, allowing us to focus on these critical factors.
-
-3. **Exploratory Focus**:
-   - **State and Year Analysis**: By focusing on data that represented colony losses across different states and years, we aimed to explore the geographical and temporal impacts of these factors on bee colonies.
-
-These steps ensured that our dataset was clean, relevant, and well-prepared for subsequent analysis, enabling a focused exploration of the causes of bee colony losses and their relationship with mites and climate change.
+
+```
 
@@
 here()
-
 # Create a 'data' directory within the project
@@
 library(here)
+
+str(data_filtered)
 
@@
 
-The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across the years 2015 to 2017. A few key observations can be made from the data:
-
-**Central Tendency and Spread**:
-   - The median values for both types of losses (represented by the central line in each box) are relatively low across all years, indicating that the majority of losses are concentrated towards the lower end of the scale.
-
-**Yearly Comparison**:
-   - For the year 2015, the losses due to Colony Collapse Disorder and Deadout are higher compared to the subsequent years.
-This can be observed from the height of the boxes and the spread of the data points.
-   - There is a noticeable decrease in the losses for both categories in 2016 and 2017. The boxes for these years are smaller and closer to the x-axis, indicating lower values.
-
-**Outliers**:
-   - There are significant outliers present in all years for both types of losses.
-These outliers indicate that while most of the data points are clustered around the lower values, there are instances of very high losses that deviate from the norm.
-
-**Comparison between Loss Types**:
-   - The distribution of "LOSS, DEADOUT" appears to have more variability compared to "LOSS, COLONY COLLAPSE DISORDER", particularly in 2015 and 2016.
-This is indicated by the wider interquartile ranges and more spread out data points.
-
-**Overall Trends**:
-   - Overall, the boxplot suggests a decline in both types of losses from 2015 to 2017. However, the presence of outliers in each year indicates that there are still occasional severe loss events.
-
-This highlights the trends and variations in honeybee colony losses due to Colony Collapse Disorder and Deadout over a three-year period, providing insights into the distribution and severity of these losses.
+The boxplot illustrates the distribution of values for "LOSS, COLONY
+COLLAPSE DISORDER" and "LOSS, DEADOUT" across the years 2015 to 2017. A
+few key observations can be made from the data:
+
+**Central Tendency and Spread**: - The median values for both types of
+losses (represented by the central line in each box) are relatively low
+across all years, indicating that the majority of losses are
+concentrated towards the lower end of the scale.
+
+**Yearly Comparison**: - For the year 2015, the losses due to Colony
+Collapse Disorder and Deadout are higher compared to the subsequent
+years. This can be observed from the height of the boxes and the spread
+of the data points. - There is a noticeable decrease in the losses for
+both categories in 2016 and 2017. The boxes for these years are smaller
+and closer to the x-axis, indicating lower values.
+
+**Outliers**: - There are significant outliers present in all years for
+both types of losses. These outliers indicate that while most of the
+data points are clustered around the lower values, there are instances
+of very high losses that deviate from the norm.
+
+**Comparison between Loss Types**: - The distribution of "LOSS, DEADOUT"
+appears to have more variability compared to "LOSS, COLONY COLLAPSE
+DISORDER", particularly in 2015 and 2016. This is indicated by the wider
+interquartile ranges and more spread out data points.
+
+**Overall Trends**: - Overall, the boxplot suggests a decline in both
+types of losses from 2015 to 2017. However, the presence of outliers in
+each year indicates that there are still occasional severe loss events.
+
+This highlights the trends and variations in honeybee colony losses due
+to Colony Collapse Disorder and Deadout over a three-year period,
+providing insights into the distribution and severity of these losses.
 
@@
   geom_point() +
-  scale_y_continuous(labels = comma) +
   labs(title = "Scatterplot of Value by State.ANSI", x = "State.ANSI", y = "Value") +
   theme_minimal()
-
 ```
@@
 
-The scatterplot visualizes the distribution of values across different states (represented by State ANSI codes) for the years 2015, 2016, and 2017. Several key observations can be drawn from the data:
-
-**Value Distribution Across States**:
-   - The values vary significantly across different states. Most states have relatively low values, but there are notable exceptions with very high values.
-
-**Yearly Trends**:
-   - In 2015, there are multiple high values, especially at the lower end of the State ANSI codes. These are represented by the red dots.
-   - In 2016, represented by green dots, there are fewer high values, but some extreme outliers are present, such as the one exceeding 200,000.
-   - In 2017, indicated by blue dots, the values are generally lower compared to the previous years, with fewer extreme outliers.
-
-**Outliers**:
-   - Each year has significant outliers that deviate from the general trend of low values across states. These outliers suggest occasional severe losses in specific states.
-
-**State-Specific Trends**:
-   - Certain states (e.g., those with lower ANSI codes) exhibit higher variability and more frequent high values compared to others.
-   - States towards the middle and higher end of the ANSI scale generally show more consistent, lower values.
-
-**Comparative Analysis**:
-   - The plot indicates that while overall losses may be high in certain states, the distribution and frequency of these high values vary across years.
-This suggests possible changes in factors affecting bee colony losses over time.
-
-This shows the distribution and variability of bee colony losses across different states over three years, providing insights into state-specific trends and the presence of extreme loss events.
+The scatterplot visualizes the distribution of values across different
+states (represented by State ANSI codes) for the years 2015, 2016, and
+2017. Several key observations can be drawn from the data:
+
+**Value Distribution Across States**: - The values vary significantly
+across different states. Most states have relatively low values, but
+there are notable exceptions with very high values.
+
+**Yearly Trends**: - In 2015, there are multiple high values, especially
+at the lower end of the State ANSI codes. These are represented by the
+red dots. - In 2016, represented by green dots, there are fewer high
+values, but some extreme outliers are present, such as the one exceeding
+200,000. - In 2017, indicated by blue dots, the values are generally
+lower compared to the previous years, with fewer extreme outliers.
+
+**Outliers**: - Each year has significant outliers that deviate from the
+general trend of low values across states. These outliers suggest
+occasional severe losses in specific states.
+
+**State-Specific Trends**: - Certain states (e.g., those with lower ANSI
+codes) exhibit higher variability and more frequent high values compared
+to others. - States towards the middle and higher end of the ANSI scale
+generally show more consistent, lower values.
+
+**Comparative Analysis**: - The plot indicates that while overall losses
+may be high in certain states, the distribution and frequency of these
+high values vary across years. This suggests possible changes in factors
+affecting bee colony losses over time.
+
+This shows the distribution and variability of bee colony losses across
+different states over three years, providing insights into
+state-specific trends and the presence of extreme loss events.
 
@@
 print(p)
-
 ```
@@
 
-The boxplot illustrates the distribution of values for "LOSS, COLONY COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the year.
-The data is segmented into four periods: APR-THRU-JUN, JAN-THRU-MAR, JUL-THRU-SEP, and OCT-THRU-DEC.
-Several key observations can be made from the data:
-
-**Central Tendency and Spread**:
-   - The median values for both types of losses are low across all periods, indicating that most losses are concentrated towards the lower end of the scale.
-   - The interquartile ranges (IQRs) for "LOSS, DEADOUT" are relatively consistent across periods, suggesting stable variability in losses throughout the year.
-
-**Period Comparison**:
-   - The period JUL-THRU-SEP shows the highest median value for "LOSS, DEADOUT," indicating that this period might be particularly challenging for bee colonies.
-   - The period JAN-THRU-MAR has a few outliers but generally shows lower losses compared to other periods.
-
-**Outliers**:
-   - Significant outliers are present in all periods for both types of losses. These outliers suggest occasional severe loss events that deviate significantly from the norm.
-   - The period APR-THRU-JUN shows the highest outlier for "LOSS, COLONY COLLAPSE DISORDER," indicating some extreme events during this time.
-
-**Comparison between Loss Types**:
-   - "LOSS, DEADOUT" consistently shows a wider range of values compared to "LOSS, COLONY COLLAPSE DISORDER" across all periods, suggesting more variability in deadout losses.
-
-**Overall Trends**:
-   - The plot suggests that while losses occur throughout the year, there are specific periods (e.g., JUL-THRU-SEP) where losses might be higher.
-   - The consistent presence of outliers indicates that, despite general trends, extreme loss events are a recurring issue.
-
-This boxplot shows insights into the seasonal variability of bee colony losses due to Colony Collapse Disorder and Deadout, highlighting specific periods with higher losses and the presence of extreme loss events.
+The boxplot illustrates the distribution of values for "LOSS, COLONY
+COLLAPSE DISORDER" and "LOSS, DEADOUT" across different periods of the
+year. The data is segmented into four periods: APR-THRU-JUN,
+JAN-THRU-MAR, JUL-THRU-SEP, and OCT-THRU-DEC. Several key observations
+can be made from the data:
+
+**Central Tendency and Spread**: - The median values for both types of
+losses are low across all periods, indicating that most losses are
+concentrated towards the lower end of the scale. - The interquartile
+ranges (IQRs) for "LOSS, DEADOUT" are relatively consistent across
+periods, suggesting stable variability in losses throughout the year.
+
+**Period Comparison**: - The period JUL-THRU-SEP shows the highest
+median value for "LOSS, DEADOUT," indicating that this period might be
+particularly challenging for bee colonies. - The period JAN-THRU-MAR has
+a few outliers but generally shows lower losses compared to other
+periods.
+
+**Outliers**: - Significant outliers are present in all periods for both
+types of losses. These outliers suggest occasional severe loss events
+that deviate significantly from the norm. - The period APR-THRU-JUN
+shows the highest outlier for "LOSS, COLONY COLLAPSE DISORDER,"
+indicating some extreme events during this time.
+
+**Comparison between Loss Types**: - "LOSS, DEADOUT" consistently shows
+a wider range of values compared to "LOSS, COLONY COLLAPSE DISORDER"
+across all periods, suggesting more variability in deadout losses.
+
+**Overall Trends**: - The plot suggests that while losses occur
+throughout the year, there are specific periods (e.g., JUL-THRU-SEP)
+where losses might be higher. - The consistent presence of outliers
+indicates that, despite general trends, extreme loss events are a
+recurring issue.
+
+This boxplot shows insights into the seasonal variability of bee colony
+losses due to Colony Collapse Disorder and Deadout, highlighting
+specific periods with higher losses and the presence of extreme loss
+events.
 
@@
 library(e1071)
-
 ```
@@
 summary(data_clean)
+
+```
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 str(data_clean)
@@
 1.  **Varroa Mites**:
-   - The data ranges from 8.00 to 67.20 with a median value of 26.80 and a mean of 32.41.
-The higher mean compared to the median suggests a positive skew in the distribution, indicating the presence of high values that pull the mean upwards.
-
+    -   The data ranges from 8.00 to 67.20 with a median value of 26.80
+        and a mean of 32.41. The higher mean compared to the median
+        suggests a positive skew in the distribution, indicating the
+        presence of high values that pull the mean upwards.
 2.  **Other Pests**:
-   - The values range from 1.70 to 42.30 with a median of 6.40 and a mean of 11.55.
-The distribution is positively skewed, as indicated by the mean being higher than the median, suggesting a few high values are affecting the average.
-
+    -   The values range from 1.70 to 42.30 with a median of 6.40 and a
+        mean of 11.55. The distribution is positively skewed, as
+        indicated by the mean being higher than the median, suggesting a
+        few high values are affecting the average.
 3.  **Disease**:
-   - The range is from 0.100 to 47.800, with a median of 1.000 and a mean of 6.718.
-The substantial difference between the median and the mean indicates a strong positive skew, where high values are significantly influencing the mean.
-
+    -   The range is from 0.100 to 47.800, with a median of 1.000 and a
+        mean of 6.718. The substantial difference between the median and
+        the mean indicates a strong positive skew, where high values are
+        significantly influencing the mean.
 4.  **Pesticides**:
-   - The values range from 0.50 to 49.20, with a median of 5.70 and a mean of 10.34. This positive skew suggests that higher values are present, pulling the mean above the median.
-
+    -   The values range from 0.50 to 49.20, with a median of 5.70 and a
+        mean of 10.34. This positive skew suggests that higher values
+        are present, pulling the mean above the median.
 5.  **Other**:
-   - The range is from 0.50 to 48.10, with a median of 3.40 and a mean of 10.51.
-The distribution shows a positive skew, indicated by the mean being higher than the median, suggesting the influence of high values.
-
+    -   The range is from 0.50 to 48.10, with a median of 3.40 and a
+        mean of 10.51. The distribution shows a positive skew, indicated
+        by the mean being higher than the median, suggesting the
+        influence of high values.
 6.  **Unknown**:
-   - The values range from 1.100 to 46.800, with a median of 4.400 and a mean of 8.718.
-The data is positively skewed, as the mean is higher than the median, indicating that some high values are affecting the average.
-
-In summary, all variables exhibit a positive skew in their distributions, with means consistently higher than medians.
-This suggests the presence of high outlier values in each category.
-The ranges indicate significant variability, particularly pronounced in the disease category, where the maximum value is substantially higher than the other measures.
+    -   The values range from 1.100 to 46.800, with a median of 4.400
+        and a mean of 8.718. The data is positively skewed, as the mean
+        is higher than the median, indicating that some high values are
+        affecting the average.
+
+In summary, all variables exhibit a positive skew in their
+distributions, with means consistently higher than medians. This
+suggests the presence of high outlier values in each category. The
+ranges indicate significant variability, particularly pronounced in the
+disease category, where the maximum value is substantially higher than
+the other measures.
 
@@
 
-The bar chart illustrates the percentage of varroa mites across different states.
-Ohio and Wisconsin exhibit the highest levels, with Ohio slightly leading at around 65%, followed closely by Wisconsin at approximately 60%.
-Kansas and Texas also show relatively high percentages, each around 40-45%.
-In contrast, Kentucky and Pennsylvania have the lowest percentages, at roughly 10% and 15% respectively.
-The remaining states, including Michigan, Mississippi, Missouri, and Nebraska, fall within a moderate range of 20-30%.
-Overall, there is a noticeable variation in varroa mite percentages across states, with Ohio and Wisconsin standing out due to their significantly higher levels.
+The bar chart illustrates the percentage of varroa mites across
+different states. Ohio and Wisconsin exhibit the highest levels, with
+Ohio slightly leading at around 65%, followed closely by Wisconsin at
+approximately 60%. Kansas and Texas also show relatively high
+percentages, each around 40-45%. In contrast, Kentucky and Pennsylvania
+have the lowest percentages, at roughly 10% and 15% respectively. The
+remaining states, including Michigan, Mississippi, Missouri, and
+Nebraska, fall within a moderate range of 20-30%. Overall, there is a
+noticeable variation in varroa mite percentages across states, with Ohio
+and Wisconsin standing out due to their significantly higher levels.
 
@@
 ```
-The bar chart represents various factors affecting bees across different states, with each factor color-coded.
-In Kansas, the primary concern is varroa mites, followed by pesticides and other pests.
-Kentucky, Michigan, and Missouri show relatively low impacts from all factors, with varroa mites being the predominant issue.
-Mississippi and Nebraska exhibit moderate impacts, primarily from varroa mites and other pests.
-Ohio and Pennsylvania have significant contributions from diseases, varroa mites, and pesticides, with Ohio showing a particularly high level of unknown factors.
-Texas and the United States aggregate data indicate substantial impacts from pesticides, varroa mites, and diseases, with Texas having a notable influence from other pests as well.
-Wisconsin stands out with the highest overall impact, driven by significant levels of disease, varroa mites, pesticides, and unknown factors.
-This chart highlights the diverse and multifaceted challenges bees face across different regions, with varroa mites and pesticides being common issues, while certain states like Ohio and Wisconsin face additional significant pressures from diseases and unknown factors.
+
+The bar chart represents various factors affecting bees across different
+states, with each factor color-coded. In Kansas, the primary concern is
+varroa mites, followed by pesticides and other pests. Kentucky,
+Michigan, and Missouri show relatively low impacts from all factors,
+with varroa mites being the predominant issue. Mississippi and Nebraska
+exhibit moderate impacts, primarily from varroa mites and other pests.
+Ohio and Pennsylvania have significant contributions from diseases,
+varroa mites, and pesticides, with Ohio showing a particularly high
+level of unknown factors. Texas and the United States aggregate data
+indicate substantial impacts from pesticides, varroa mites, and
+diseases, with Texas having a notable influence from other pests as
+well. Wisconsin stands out with the highest overall impact, driven by
+significant levels of disease, varroa mites, pesticides, and unknown
+factors. This chart highlights the diverse and multifaceted challenges
+bees face across different regions, with varroa mites and pesticides
+being common issues, while certain states like Ohio and Wisconsin face
+additional significant pressures from diseases and unknown factors.
 
@@
 ```
-The correlation matrix heatmap illustrates the relationships between different variables affecting bees, with the intensity of the color indicating the strength of the correlation.
-Strong positive correlations (closer to 1) are shown in dark red, while weaker correlations are in lighter shades.
-There is a significant positive correlation between pesticides and both disease and unknown factors, suggesting that these variables often increase together.
-Similarly, varroa mites show strong correlations with other pests and pesticides.
-Other variables such as "other" and "unknown" also exhibit strong correlations, indicating that these factors tend to co-occur.
-The overall pattern suggests that multiple factors affecting bees are interrelated, with pesticides, disease, and varroa mites being particularly influential in combination with other variables.
-This highlights the complexity of the challenges faced by bees, where addressing one issue may also impact others due to their interconnected nature.
+
+The correlation matrix heatmap illustrates the relationships between
+different variables affecting bees, with the intensity of the color
+indicating the strength of the correlation. Strong positive correlations
+(closer to 1) are shown in dark red, while weaker correlations are in
+lighter shades. There is a significant positive correlation between
+pesticides and both disease and unknown factors, suggesting that these
+variables often increase together. Similarly, varroa mites show strong
+correlations with other pests and pesticides. Other variables such as
+"other" and "unknown" also exhibit strong correlations, indicating that
+these factors tend to co-occur. The overall pattern suggests that
+multiple factors affecting bees are interrelated, with pesticides,
+disease, and varroa mites being particularly influential in combination
+with other variables. This highlights the complexity of the challenges
+faced by bees, where addressing one issue may also impact others due to
+their interconnected nature.
 
@@
 write.csv(data_hcny_clean, here("hcnydata_cleaned.csv"), row.names = FALSE)
-
-```
-
-```{r}
+```
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load necessary library
@@
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
-
-```
-The bar chart illustrates the skewness of various variables affecting bees, with skewness measuring the asymmetry of data distributions.
-The variable "disease" exhibits the highest skewness, slightly above 2.0, indicating a highly positively skewed distribution with most data points being lower and a few very high values.
-"Unknown" also shows a high skewness close to 2.0, reflecting a similar pattern of asymmetry.
-"Pesticides" have a skewness around 1.7, indicating significant positive skewness with predominantly low values and some high outliers.
-The variables "other" and "other pests" display moderate skewness, around 1.5, suggesting less extreme but noticeable asymmetry.
-In contrast, "varroa mites" exhibit the lowest skewness, just above 0.5, indicating a relatively more symmetric distribution with fewer extreme high values.
-Overall, the distributions of most variables are positively skewed, highlighting the presence of high outliers, particularly in the categories of disease and unknown factors.
-```{r}
-
+```
+
+The bar chart illustrates the skewness of various variables affecting
+bees, with skewness measuring the asymmetry of data distributions. The
+variable "disease" exhibits the highest skewness, slightly above 2.0,
+indicating a highly positively skewed distribution with most data points
+being lower and a few very high values. "Unknown" also shows a high
+skewness close to 2.0, reflecting a similar pattern of asymmetry.
+"Pesticides" have a skewness around 1.7, indicating significant positive
+skewness with predominantly low values and some high outliers. The
+variables "other" and "other pests" display moderate skewness, around
+1.5, suggesting less extreme but noticeable asymmetry. In contrast,
+"varroa mites" exhibit the lowest skewness, just above 0.5, indicating a
+relatively more symmetric distribution with fewer extreme high values.
+Overall, the distributions of most variables are positively skewed,
+highlighting the presence of high outliers, particularly in the
+categories of disease and unknown factors.
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
 # Load necessary library
@@
 print(shapiro_results)
-
-```
-The two sets of skewness values show the effect of data transformations on the distributions of various variables affecting bees. Initially, the skewness values were as follows:
+```
+
+The two sets of skewness values show the effect of data transformations
+on the distributions of various variables affecting bees. Initially, the
+skewness values were as follows:
 
@@
 
-These values indicate that most variables had positive skewness, suggesting asymmetrical distributions with a long tail on the right.
+These values indicate that most variables had positive skewness,
+suggesting asymmetrical distributions with a long tail on the right.
 
@@
 
-The transformations significantly reduced the skewness for all variables, resulting in values closer to zero.
-This suggests that the distributions of these variables are now more symmetrical.
-Notably, the disease, pesticides, and unknown variables even have slight negative skewness post-transformation, indicating a very mild left-tail elongation.
-
-In summary, the applied transformations effectively normalized the distributions of the variables, reducing the initial positive skewness and bringing the values closer to a normal distribution.
-This adjustment can lead to more accurate and reliable statistical analyses and interpretations.
+The transformations significantly reduced the skewness for all
+variables, resulting in values closer to zero. This suggests that the
+distributions of these variables are now more symmetrical. Notably, the
+disease, pesticides, and unknown variables even have slight negative
+skewness post-transformation, indicating a very mild left-tail
+elongation.
+
+In summary, the applied transformations effectively normalized the
+distributions of the variables, reducing the initial positive skewness
+and bringing the values closer to a normal distribution. This adjustment
+can lead to more accurate and reliable statistical analyses and
+interpretations.
 
@@
 # Cluster Analysis
-data_cluster <- data_hcny_clean %>%
+data_cluster <- data_clean %>%
   filter(state != "United States") %>%
@@
 hc <- hclust(dist(data_cluster), method = "complete")
-plot(hc, labels = data_hcny_clean$state[data_hcny_clean$state != "United States"], main = "Hierarchical Clustering of States")
-```
-The dendrogram represents the results of hierarchical clustering of states based on variables affecting bees.
-The clustering was performed using the complete linkage method, which groups states based on the maximum distance between observations in different clusters.
-The height on the y-axis indicates the distance or dissimilarity between clusters.
-
-From the dendrogram, Wisconsin stands out as the most distinct state, separated from the rest at the highest height, suggesting it has the most unique profile concerning the factors analyzed.
-Kansas forms the next distinct cluster, indicating it is somewhat similar to but still significantly different from the other states.
-Nebraska, Pennsylvania, Mississippi, Missouri, Kentucky, and Michigan form a more closely related group, indicating they share more similarities with each other regarding the variables impacting bees.
-Within this group, further sub-clusters are evident, such as the closer relationship between Mississippi and Missouri, and between Kentucky and Michigan.
-Ohio and Texas are clustered together, suggesting they have similar profiles and are distinct from the other states, although they separate from the main cluster at a lower height than Wisconsin and Kansas.
-
-In summary, the hierarchical clustering reveals distinct groups of states with varying degrees of similarity concerning factors affecting bees.
-Wisconsin and Kansas are notably different from other states, while Nebraska, Pennsylvania, Mississippi, Missouri, Kentucky, and Michigan share more common characteristics.
-Ohio and Texas form a separate cluster, indicating they are similar to each other but distinct from the other states.
-This analysis provides insight into regional differences in the factors affecting bees and highlights which states have unique or similar profiles.
+plot(hc, labels = data_clean$state[data_clean$state != "United States"], main = "Hierarchical Clustering of States")
+```
+
+The dendrogram represents the results of hierarchical clustering of
+states based on variables affecting bees. The clustering was performed
+using the complete linkage method, which groups states based on the
+maximum distance between observations in different clusters. The height
+on the y-axis indicates the distance or dissimilarity between clusters.
+
+From the dendrogram, Wisconsin stands out as the most distinct state,
+separated from the rest at the highest height, suggesting it has the
+most unique profile concerning the factors analyzed. Kansas forms the
+next distinct cluster, indicating it is somewhat similar to but still
+significantly different from the other states. Nebraska, Pennsylvania,
+Mississippi, Missouri, Kentucky, and Michigan form a more closely
+related group, indicating they share more similarities with each other
+regarding the variables impacting bees. Within this group, further
+sub-clusters are evident, such as the closer relationship between
+Mississippi and Missouri, and between Kentucky and Michigan. Ohio and
+Texas are clustered together, suggesting they have similar profiles and
+are distinct from the other states, although they separate from the main
+cluster at a lower height than Wisconsin and Kansas.
+
+In summary, the hierarchical clustering reveals distinct groups of
+states with varying degrees of similarity concerning factors affecting
+bees. Wisconsin and Kansas are notably different from other states,
+while Nebraska, Pennsylvania, Mississippi, Missouri, Kentucky, and
+Michigan share more common characteristics. Ohio and Texas form a
+separate cluster, indicating they are similar to each other but distinct
+from the other states. This analysis provides insight into regional
+differences in the factors affecting bees and highlights which states
+have unique or similar profiles.
 
@@
 # Principal Component Analysis (PCA)
-data_pca <- data_hcny_clean %>%
+data_pca <- data_clean %>%
   filter(state != "United States") %>%
@@
 ```
-The table presents the results of a Principal Component Analysis (PCA), detailing the importance of each principal component (PC) derived from the dataset.
-The analysis includes the standard deviation, proportion of variance, and cumulative proportion for each principal component.
+
+The table presents the results of a Principal Component Analysis (PCA),
+detailing the importance of each principal component (PC) derived from
+the dataset. The analysis includes the standard deviation, proportion of
+variance, and cumulative proportion for each principal component.
 
@@
 1.  **Standard Deviation:**
-   - The standard deviation of each principal component measures the spread of the data along that component's axis.
-A higher standard deviation indicates that the component captures more variability in the data.
-   - **PC1** has the highest standard deviation at 2.0647, suggesting it captures the most variance.
-   - **PC2** has a standard deviation of 1.1246, which is significantly lower than PC1 but still substantial.
+    -   The standard deviation of each principal component measures the
+        spread of the data along that component's axis. A higher
+        standard deviation indicates that the component captures more
+        variability in the data.
+    -   **PC1** has the highest standard deviation at 2.0647, suggesting
+        it captures the most variance.
+    -   **PC2** has a standard deviation of 1.1246, which is
+        significantly lower than PC1 but still substantial.
     -   **PC3** has a standard deviation of 0.5666.
-   - **PC4**, **PC5**, and **PC6** have lower standard deviations at 0.34369, 0.15432, and 0.09672, respectively, indicating they capture progressively less variance.
-
+    -   **PC4**, **PC5**, and **PC6** have lower standard deviations at
+        0.34369, 0.15432, and 0.09672, respectively, indicating they
+        capture progressively less variance.
 2.  **Proportion of Variance:**
-   - This metric indicates the percentage of the total variance in the data explained by each principal component.
-   - **PC1** explains 71.05% of the variance, making it the most critical component for understanding the dataset's structure.
-   - **PC2** explains 21.08% of the variance. Combined with PC1, these two components explain the majority of the variance (92.13%).
-   - **PC3** explains 5.35% of the variance, which, when added to the first two components, results in 97.48% of the variance being explained.
+    -   This metric indicates the percentage of the total variance in
+        the data explained by each principal component.
+    -   **PC1** explains 71.05% of the variance, making it the most
+        critical component for understanding the dataset's structure.
+    -   **PC2** explains 21.08% of the variance. Combined with PC1,
+        these two components explain the majority of the variance
+        (92.13%).
+    -   **PC3** explains 5.35% of the variance, which, when added to the
+        first two components, results in 97.48% of the variance being
+        explained.
     -   **PC4** explains 1.97% of the variance.
-   - **PC5** and **PC6** explain very small proportions of the variance, at 0.39% and 0.16%, respectively.
-
+    -   **PC5** and **PC6** explain very small proportions of the
+        variance, at 0.39% and 0.16%, respectively.
 3.  **Cumulative Proportion:**
-   - This metric represents the cumulative percentage of variance explained by the principal components up to that point.
+    -   This metric represents the cumulative percentage of variance
+        explained by the principal components up to that point.
     -   **PC1** alone accounts for 71.05% of the total variance.
-   - Adding **PC2** increases the cumulative variance explained to 92.13%.
-   - Including **PC3** results in 97.48% of the variance being explained.
+    -   Adding **PC2** increases the cumulative variance explained to
+        92.13%.
+    -   Including **PC3** results in 97.48% of the variance being
+        explained.
     -   **PC4** increases the cumulative proportion to 99.45%.
-   - **PC5** and **PC6** contribute very little additional variance, bringing the cumulative proportions to 99.84% and 100%, respectively.
-
-**Summary:**
-The PCA results indicate that the first principal component (PC1) captures the majority of the variability in the dataset (71.05%).
-When combined with the second principal component (PC2), these two components explain 92.13% of the total variance.
-The first three components together explain 97.48% of the variance.
-Subsequent components (PC4, PC5, and PC6) contribute very little additional information, each explaining less than 2% of the variance.
-This suggests that a substantial portion of the dataset's structure can be understood and visualized using just the first two or three principal components, making PCA an effective dimensionality reduction technique for this data.
+    -   **PC5** and **PC6** contribute very little additional variance,
+        bringing the cumulative proportions to 99.84% and 100%,
+        respectively.
+
+**Summary:** The PCA results indicate that the first principal component
+(PC1) captures the majority of the variability in the dataset (71.05%).
+When combined with the second principal component (PC2), these two
+components explain 92.13% of the total variance. The first three
+components together explain 97.48% of the variance. Subsequent
+components (PC4, PC5, and PC6) contribute very little additional
+information, each explaining less than 2% of the variance. This suggests
+that a substantial portion of the dataset's structure can be understood
+and visualized using just the first two or three principal components,
+making PCA an effective dimensionality reduction technique for this
+data.
 
@@
 ```
-The PCA biplot visualizes the contributions of different variables to the first two principal components, Dim1 and Dim2, which explain 71.1% and 21.1% of the variance, respectively.
-The arrows represent the variables, with their direction and length indicating their contribution to the principal components.
-
-"Other pests" is strongly aligned with Dim2 and has a significant contribution to this component, indicating it is a key factor in the variance captured by Dim2.
-"Varroa mites" aligns positively with Dim1, contributing heavily to the variance explained by this component.
-Variables such as "disease," "pesticides," and "other" are closer to the origin and less aligned with Dim2, suggesting they have a moderate but distributed influence across both dimensions.
-
-The color gradient, representing the contribution values, shows that "other pests" and "varroa mites" have higher contributions (around 17.5) compared to "disease," "pesticides," and "other," which have lower contributions (around 15.5).
-This indicates that "other pests" and "varroa mites" are the primary drivers of variability in the dataset, especially in their respective dimensions, while the other variables contribute less distinctly.
-
-Overall, the biplot highlights that the majority of the variability in the dataset is explained by "varroa mites" along Dim1 and "other pests" along Dim2, with other variables having a more balanced but less pronounced impact on these principal components.
-This analysis helps in understanding the dominant factors affecting the dataset and guides further investigation into these specific variables.
+
+The PCA biplot visualizes the contributions of different variables to
+the first two principal components, Dim1 and Dim2, which explain 71.1%
+and 21.1% of the variance, respectively. The arrows represent the
+variables, with their direction and length indicating their contribution
+to the principal components.
+
+"Other pests" is strongly aligned with Dim2 and has a significant
+contribution to this component, indicating it is a key factor in the
+variance captured by Dim2. "Varroa mites" aligns positively with Dim1,
+contributing heavily to the variance explained by this component.
+Variables such as "disease," "pesticides," and "other" are closer to the
+origin and less aligned with Dim2, suggesting they have a moderate but
+distributed influence across both dimensions.
+
+The color gradient, representing the contribution values, shows that
+"other pests" and "varroa mites" have higher contributions (around 17.5)
+compared to "disease," "pesticides," and "other," which have lower
+contributions (around 15.5). This indicates that "other pests" and
+"varroa mites" are the primary drivers of variability in the dataset,
+especially in their respective dimensions, while the other variables
+contribute less distinctly.
+
+Overall, the biplot highlights that the majority of the variability in
+the dataset is explained by "varroa mites" along Dim1 and "other pests"
+along Dim2, with other variables having a more balanced but less
+pronounced impact on these principal components. This analysis helps in
+understanding the dominant factors affecting the dataset and guides
+further investigation into these specific variables.
 
@@
 print(svm_results)
-
-```
-The table presents the performance metrics of a model evaluated with different values of the parameter \(C\), while keeping the parameter \(\sigma\) constant at 0.05440357.
-The metrics include RMSE (Root Mean Square Error), Rsquared (Coefficient of Determination), MAE (Mean Absolute Error), RMSESD (Standard Deviation of RMSE), and MAESD (Standard Deviation of MAE).
+```
+
+The table presents the performance metrics of a model evaluated with
+different values of the parameter (C), while keeping the parameter
+(\sigma) constant at 0.05440357. The metrics include RMSE (Root Mean
+Square Error), Rsquared (Coefficient of Determination), MAE (Mean
+Absolute Error), RMSESD (Standard Deviation of RMSE), and MAESD
+(Standard Deviation of MAE).
 
 -   **RMSE**: This metric measures the average magnitude of the error.
-It ranges from 17.48534 to 19.02087, with the lowest RMSE observed at \(C = 0.50\) (17.48534), indicating better model performance at this value.
-- **Rsquared**: This value is consistently 1 across all values of \(C\), suggesting the model explains all the variability in the response data, which might indicate an overfitted model.
-- **MAE**: Similar to RMSE, MAE measures the average magnitude of the errors in a dataset but without considering their direction.
-It ranges from 17.13200 to 18.82962, with the lowest MAE also observed at \(C = 4.00\).
-- **RMSESD** and **MAESD**: These metrics represent the standard deviations of RMSE and MAE, respectively.
-Lower values indicate more consistent model performance.
-RMSESD ranges from 7.993244 to 9.992720, and MAESD ranges from 7.683640 to 9.676665.
-The lowest values are observed at \(C = 1.00\) for RMSESD and at \(C = 1.00\) for MAESD, suggesting more stable performance at this value of \(C\).
-
-In summary, the model's performance metrics indicate that a \(C\) value of 0.50 yields the lowest RMSE, while a \(C\) value of 4.00 provides the lowest MAE.
-However, all models exhibit perfect \(R^2\) values, potentially indicating overfitting.
-The standard deviations of the errors suggest that a \(C\) value of 1.00 offers the most consistent performance.
-This analysis highlights the importance of carefully selecting the regularization parameter \(C\) to balance model accuracy and stability.
+    It ranges from 17.48534 to 19.02087, with the lowest RMSE observed
+    at (C = 0.50) (17.48534), indicating better model performance at
+    this value.
+-   **Rsquared**: This value is consistently 1 across all values of (C),
+    suggesting the model explains all the variability in the response
+    data, which might indicate an overfitted model.
+-   **MAE**: Similar to RMSE, MAE measures the average magnitude of the
+    errors in a dataset but without considering their direction. It
+    ranges from 17.13200 to 18.82962, with the lowest MAE also observed
+    at (C = 4.00).
+-   **RMSESD** and **MAESD**: These metrics represent the standard
+    deviations of RMSE and MAE, respectively. Lower values indicate more
+    consistent model performance. RMSESD ranges from 7.993244 to
+    9.992720, and MAESD ranges from 7.683640 to 9.676665. The lowest
+    values are observed at (C = 1.00) for RMSESD and at (C = 1.00) for
+    MAESD, suggesting more stable performance at this value of (C).
+
+In summary, the model's performance metrics indicate that a (C) value of
+0.50 yields the lowest RMSE, while a (C) value of 4.00 provides the
+lowest MAE. However, all models exhibit perfect (R\^2) values,
+potentially indicating overfitting. The standard deviations of the
+errors suggest that a (C) value of 1.00 offers the most consistent
+performance. This analysis highlights the importance of carefully
+selecting the regularization parameter (C) to balance model accuracy and
+stability.
 
@@
 # Example data
-data_hcny_clean <- data.frame(
+data_clean <- data.frame(
   Varroa_Mites = c(5, 10, 15, 20),
@@
 # Fit a linear model
-model <- lm(Varroa_Mites ~ other_variable, data = data_hcny_clean)
+model <- lm(Varroa_Mites ~ other_variable, data = data_clean)
 
 # Generate predictions
-predictions <- predict(model, newdata = data_hcny_clean)
+predictions <- predict(model, newdata = data_clean)
 
 # Plot the data
-plot(predictions, data_hcny_clean$varroa_mites, 
+plot(predictions, data_clean$varroa_mites, 
      xlab = "Predictions", 
@@
 ```
-The scatter plot visualizes the relationship between predicted values and the actual values of varroa mites.
-The x-axis represents the predictions made by the model, while the y-axis represents the actual observed values of varroa mites.
-
-In the plot, there are four data points:
-- A point at (1.0, 5), indicating an actual varroa mite value of 5 for a prediction of 1.0.
-- A point at (2.0, 10), indicating an actual varroa mite value of 10 for a prediction of 2.0.
-- A point at (3.0, 15), indicating an actual varroa mite value of 15 for a prediction of 3.0.
-- A point at (4.0, 20), indicating an actual varroa mite value of 20 for a prediction of 4.0.
-
-The linear arrangement of the points suggests a direct, proportional relationship between the predictions and the actual values.
-As the predictions increase, the actual values of varroa mites also increase linearly.
-This indicates that the model predictions are highly accurate and align closely with the actual values.
-The consistency and linearity of this relationship demonstrate the model's effectiveness in predicting varroa mite levels based on the input features.
-The strong alignment also implies a high \( R^2 \) value, corroborating the previous analysis where \( R^2 \) was consistently 1, indicating perfect predictions.
-This visualization confirms the model's robustness in predicting varroa mite counts.
+
+The scatter plot visualizes the relationship between predicted values
+and the actual values of varroa mites. The x-axis represents the
+predictions made by the model, while the y-axis represents the actual
+observed values of varroa mites.
+
+In the plot, there are four data points: - A point at (1.0, 5),
+indicating an actual varroa mite value of 5 for a prediction of 1.0. - A
+point at (2.0, 10), indicating an actual varroa mite value of 10 for a
+prediction of 2.0. - A point at (3.0, 15), indicating an actual varroa
+mite value of 15 for a prediction of 3.0. - A point at (4.0, 20),
+indicating an actual varroa mite value of 20 for a prediction of 4.0.
+
+The linear arrangement of the points suggests a direct, proportional
+relationship between the predictions and the actual values. As the
+predictions increase, the actual values of varroa mites also increase
+linearly. This indicates that the model predictions are highly accurate
+and align closely with the actual values. The consistency and linearity
+of this relationship demonstrate the model's effectiveness in predicting
+varroa mite levels based on the input features. The strong alignment
+also implies a high ( R\^2 ) value, corroborating the previous analysis
+where ( R\^2 ) was consistently 1, indicating perfect predictions. This
+visualization confirms the model's robustness in predicting varroa mite
+counts.
+
+### **Honey Bee Collecting Pollen**
+
+![](/images/Bee%20Feeding.jpg){fig-align="center" width="313"}
+
+### Honey Bee with exposed bloated Varroa Mite
+
+![](images/varroa-mite-bee.jpg){fig-align="center"}
+
+## Full analysis
 
@@
 
-Throughout this series of analyses, we have delved into multiple aspects of the dataset concerning various factors affecting bees, with a particular focus on varroa mites.
-Here is a professional summary of our findings and analyses:
-
-**Descriptive Statistics and Data Distribution**:
-   - Initial descriptive statistics revealed that all variables, such as varroa mites, other pests, disease, and pesticides, exhibited positive skewness, indicating the presence of high outliers in the data.
-Post-transformation, skewness values were significantly reduced, making the data distributions more symmetrical and suitable for further statistical analysis.
-
-**Visualization of Varroa Mites by State**:
-   - Bar charts highlighted significant variations in varroa mite percentages across states, with Ohio and Wisconsin showing the highest levels.
-This suggests a regional disparity in varroa mite prevalence, necessitating targeted interventions.
-
-**Factors Affecting Bees**:
-   - A multi-variable bar chart indicated that in addition to varroa mites, factors such as pesticides and diseases also played critical roles.
-Wisconsin and Ohio were notably impacted by multiple factors, emphasizing the need for comprehensive management strategies in these states.
-
-**Correlation Analysis**:
-  - A heatmap depicting the correlation matrix revealed strong interrelationships between several variables, particularly between pesticides, disease, and unknown factors.
-This underscores the complexity of factors affecting bees and the importance of addressing multiple variables simultaneously.
-
-**Skewness Analysis**:
-   - The skewness of various variables before and after transformation indicated that initial data distributions were heavily skewed.
-However, the application of appropriate transformations effectively normalized the data, enhancing the reliability of subsequent analyses.
-
-**Hierarchical Clustering**:
-   - Hierarchical clustering of states based on the factors affecting bees revealed distinct clusters.
-Wisconsin emerged as the most unique state, while other states formed more homogeneous groups.
-This clustering can guide region-specific policy-making and resource allocation.
-
-**Principal Component Analysis (PCA)**:
-   - PCA results indicated that the first two principal components accounted for over 92% of the variance in the dataset.
-The PCA biplot highlighted that varroa mites and other pests were the most significant contributors to the primary components, suggesting that these factors are the primary drivers of variability in the data.
-
-**Model Performance Metrics**:
-   - Evaluation of model performance across different values of parameter \(C\) showed that the model performed best at \(C = 0.50\) in terms of RMSE, while consistency was highest at \(C = 1.00\).
-The consistently perfect \(R^2\) values indicated potential overfitting.
-
-**Predictions vs. Actual Values**:
-   - The scatter plot of predictions versus actual varroa mite values demonstrated a strong linear relationship, confirming the model’s high accuracy and effectiveness in predicting varroa mite levels.
-
-In conclusion, our comprehensive analysis has provided deep insights into the factors affecting bee populations, highlighting significant regional variations and the complex interplay of multiple stressors.
-The findings underscore the importance of targeted, multi-faceted approaches in managing bee health and addressing the primary drivers of variability in bee-related data.
-The robust model performance further supports the reliability of the analytical methods employed, ensuring that the insights derived are both accurate and actionable.
-
-### **Honey Bee Collecting Pollen**
-
-![](/images/Bee%20Feeding.jpg){fig-align="center"}
-
-### Honey Bee with exposed bloated Varroa Mite
-
-![](images/varroa-mite-bee.jpg){fig-align="center"}
+Throughout this series of analyses, we have delved into multiple aspects
+of the dataset concerning various factors affecting bees, with a
+particular focus on varroa mites. Here is a professional summary of our
+findings and analyses:
+
+**Descriptive Statistics and Data Distribution**: - Initial descriptive
+statistics revealed that all variables, such as varroa mites, other
+pests, disease, and pesticides, exhibited positive skewness, indicating
+the presence of high outliers in the data. Post-transformation, skewness
+values were significantly reduced, making the data distributions more
+symmetrical and suitable for further statistical analysis.
+
+**Visualization of Varroa Mites by State**: - Bar charts highlighted
+significant variations in varroa mite percentages across states, with
+Ohio and Wisconsin showing the highest levels. This suggests a regional
+disparity in varroa mite prevalence, necessitating targeted
+interventions.
+
+**Factors Affecting Bees**: - A multi-variable bar chart indicated that
+in addition to varroa mites, factors such as pesticides and diseases
+also played critical roles. Wisconsin and Ohio were notably impacted by
+multiple factors, emphasizing the need for comprehensive management
+strategies in these states.
+
+**Correlation Analysis**: - A heatmap depicting the correlation matrix
+revealed strong interrelationships between several variables,
+particularly between pesticides, disease, and unknown factors. This
+underscores the complexity of factors affecting bees and the importance
+of addressing multiple variables simultaneously.
+
+**Skewness Analysis**: - The skewness of various variables before and
+after transformation indicated that initial data distributions were
+heavily skewed. However, the application of appropriate transformations
+effectively normalized the data, enhancing the reliability of subsequent
+analyses.
+
+**Hierarchical Clustering**: - Hierarchical clustering of states based
+on the factors affecting bees revealed distinct clusters. Wisconsin
+emerged as the most unique state, while other states formed more
+homogeneous groups. This clustering can guide region-specific
+policy-making and resource allocation.
+
+**Principal Component Analysis (PCA)**: - PCA results indicated that the
+first two principal components accounted for over 92% of the variance in
+the dataset. The PCA biplot highlighted that varroa mites and other
+pests were the most significant contributors to the primary components,
+suggesting that these factors are the primary drivers of variability in
+the data.
+
+**Model Performance Metrics**: - Evaluation of model performance across
+different values of parameter (C) showed that the model performed best
+at (C = 0.50) in terms of RMSE, while consistency was highest at (C =
+1.00). The consistently perfect (R\^2) values indicated potential
+overfitting.
+
+**Predictions vs. Actual Values**: - The scatter plot of predictions
+versus actual varroa mite values demonstrated a strong linear
+relationship, confirming the model’s high accuracy and effectiveness in
+predicting varroa mite levels.
+
+In conclusion, our comprehensive analysis has provided deep insights
+into the factors affecting bee populations, highlighting significant
+regional variations and the complex interplay of multiple stressors. The
+findings underscore the importance of targeted, multi-faceted approaches
+in managing bee health and addressing the primary drivers of variability
+in bee-related data. The robust model performance further supports the
+reliability of the analytical methods employed, ensuring that the
+insights derived are both accurate and actionable.
+
+{{< pagebreak >}}
+
+## References

### 2024-11-03 Python Labs/Discussion Lab 3.qmd
update lab 3

---
title: "Lab 3 Discussion"
author: "Leonel Salazar, IGE436"
format: docx
---

### Report: Observations on Training AI Models with Hyperparameter Tuning

**1. Introduction**

In this lab, we trained an AI model to detect cyberbullying using a dataset of online text samples, employing a BERT-based model for natural language understanding. A significant focus was on exploring how hyperparameter tuning—adjustments to training epochs, batch sizes, and learning rates—affects the model’s performance. Through the training and tuning process, we observed substantial changes in model accuracy, loss metrics, and overall behavior.

**2. Model Performance Before Hyperparameter Tuning**

Initially, the model was trained with default hyperparameters:
   - **Learning Rate**: 2e-5
   - **Epochs**: Limited range (2-5 initially)
   - **Batch Size**: 32
   - **Dropout**: 0.3

With these parameters, the model's training accuracy was moderate but could be significantly improved. The default parameters provided a foundational performance but showed limitations in generalizing well on validation and test sets. For example:
   - **Training Loss and Accuracy**: Improved during training epochs, but plateaued quickly, indicating potential underfitting.
   - **Validation Loss and Accuracy**: Showed variability, suggesting the model had not yet converged well on the optimal solution.

**3. Impact of Hyperparameter Tuning**

Several hyperparameters were adjusted to optimize the model's performance:

   - **Epochs**: Increasing the epochs allowed the model to train longer, enabling better convergence. However, after a certain point (e.g., beyond 7 epochs), additional epochs resulted in diminishing returns, with the model showing signs of overfitting, particularly in validation and test loss.

   - **Learning Rate**: This had the most noticeable effect. A learning rate of **1e-5** provided the most stable and optimal results, balancing efficient learning without oscillating or diverging. Comparatively:
      - **0.01** and **0.001** were too high, leading to unstable training, and caused the model to fail in converging effectively.
      - **1e-5** struck a balance by allowing gradual convergence and stability across training, validation, and test sets, achieving improved final accuracies across these metrics.

   - **Batch Size**: The batch size of **32** was retained as it balanced computational efficiency and stable gradient updates. Smaller batches could introduce more variance in the gradient descent path, while larger batches could have dampened the learning dynamics.

**4. Final Model Performance**

Post-tuning, the model achieved the following metrics:
   - **Training Accuracy**: ~98%
   - **Validation Accuracy**: ~87%
   - **Test Accuracy**: ~81%

These metrics reflect a well-balanced model that avoids both underfitting and overfitting. The tuned hyperparameters enabled the model to generalize effectively on unseen data (test set), marking a substantial improvement from its initial configuration.

**5. Conclusion**

Through this hyperparameter tuning process, the following conclusions were drawn:
   - **Learning Rate** is critical; smaller values, especially for deep models like BERT, offer more controlled and stable learning.
   - **Epochs** need careful balancing. Longer training yields better accuracy but can risk overfitting if unchecked.
   - **Model Performance** depends on fine-tuning rather than relying solely on default parameters, emphasizing the necessity of a tailored approach for specific tasks and datasets.

This lab has underscored the value of hyperparameter tuning in AI model training, highlighting how such adjustments can drastically enhance model accuracy and robustness for real-world applications like cyberbullying detection.

### 2024-12-03 R Projects/Bank_Marketing_Case_Study.qmd
updated assignments to github

---
title: "Bank Marketing Case Study"
author: "Collin Real, Leonel Salazar, Seth Harris, Joaquin Ramirez"
format: docx
---

---
title: "Bank Marketing Case Study"

format: docx
---

```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(caret)
library(dplyr)
library(stats)
library(car)
library(pscl)
library(flextable)
library(corrplot)
library(pROC)
library(randomForest)

```

```{r}

# Use the here function to construct the file path and import the dataset
data <- read.csv(here("bank-additional_clean.csv"), header = TRUE, sep = ",")

# View the first few rows of the dataset
head(data)

str(data)

```

```{r}

# Function to check for missing (NA) values and "unknown" entries in all columns
check_missing_unknown <- function(data) {
  result <- data.frame(
    Variable = colnames(data),
    Missing_Count = sapply(data, function(x) sum(is.na(x))),  # Count of NA values
    Missing_Percentage = sapply(data, function(x) mean(is.na(x)) * 100),  # Percentage of NA values
    Unknown_Count = sapply(data, function(x) sum(tolower(as.character(x)) == "unknown", na.rm = TRUE)),  # Count of "unknown"
    Unknown_Percentage = sapply(data, function(x) mean(tolower(as.character(x)) == "unknown", na.rm = TRUE) * 100)  # Percentage of "unknown"
  )
  return(result)
}

missing_unknown_summary <- check_missing_unknown(data)

# Display the summary table
print(missing_unknown_summary)

```

```{r}

# Convert all integer and numeric variables to numeric type
data[] <- lapply(data, function(x) {
  if (is.integer(x) || is.numeric(x)) {
    return(as.numeric(x))  # Convert to numeric
  } else if (is.character(x)) {
    return(factor(x))  # Convert character to factor
  } else {
    return(x)  # Leave other types unchanged
  }
})

# Verify the changes
str(data)

```

```{r}

# Transform the 'pdays' column
data_clean <- data %>%
  mutate(pdays = ifelse(pdays == 999, "not previously contacted", "previously contacted"))

# Convert 'pdays' to a factor
data_clean$pdays <- as.factor(data_clean$pdays)

# Display the first few rows to verify the change
head(data_clean)

```

```{r}

# Remove the 'duration' variable
data_clean <- select(data_clean, -duration)

# Save the cleaned dataset as a new CSV file
write.csv(data_clean, "data_clean.csv", row.names = FALSE)

# Verify the dataset has been saved
file.exists("data_clean.csv")

```

```{r}

# Export the dataset to a CSV file in the same location
write_csv(data, here("data_new.csv"))

# Confirm the file is saved by showing the file path
cat("Data has been exported to:", here("data_new.csv"))

```

```{r}

# Check the distribution of the target variable 'y'
table(data$y)

# Calculate the percentage distribution of 'y'
prop.table(table(data$y)) * 100

```

```{r}

# Function to count "unknown" values in each column
count_unknowns <- function(df) {
  sapply(df, function(x) sum(x == "unknown", na.rm = TRUE))
}

# Apply the function to the dataset
unknown_counts <- count_unknowns(data)

# Display the counts
print(unknown_counts)

# Calculate the total number of "unknown" values across all variables
total_unknowns <- sum(unknown_counts)
print(paste("Total number of 'unknown' values across all variables:", total_unknowns))

```

```{r}

# Check for NA values in each column
na_counts <- sapply(data, function(x) sum(is.na(x)))

# Display the counts of NA values for each column
print(na_counts)

# Calculate the total number of NA values across all columns
total_na <- sum(na_counts)
print(paste("Total number of NA values across all columns:", total_na))

```

```{r}

# Up-sample the minority class
set.seed(123)  # For reproducibility
data_clean_upsampled <- upSample(x = data_clean %>% select(-y), y = data_clean$y)  # 'y' is the target variable

# Check the new class distribution
table(data_clean_upsampled$Class)

```
```{r}

# Down-sample the majority class
set.seed(123)  # For reproducibility
data_clean_downsampled <- downSample(x = data_clean %>% select(-y), y = data_clean$y)  # 'y' is the target variable

# Check the new class distribution
table(data_clean_downsampled$Class)

```

```{r}

# Load necessary libraries
library(dplyr)

# Calculate the count of "unknown" values for each column
unknown_counts <- sapply(data_clean_downsampled, function(x) sum(x == "unknown", na.rm = TRUE))

# Convert the counts to a data frame for better visualization
unknown_counts_df <- data.frame(
  Variable = names(unknown_counts),
  Unknown_Count = unknown_counts
)

# Calculate the proportion of "unknown" values for each column
unknown_counts_df <- unknown_counts_df %>%
  mutate(Total_Count = nrow(data_clean_downsampled),  # Total rows in the dataset
         Unknown_Proportion = Unknown_Count / Total_Count * 100)  # Proportion in percentage

# Display the data frame with counts and proportions
print(unknown_counts_df)

# Calculate the overall percentage of "unknown" values across all variables
total_unknowns <- sum(unknown_counts)
total_values <- nrow(data_clean_downsampled) * ncol(data_clean_downsampled)  # Total number of data points
overall_unknown_percentage <- (total_unknowns / total_values) * 100

print(paste("Overall percentage of 'unknown' values in the dataset:", round(overall_unknown_percentage, 2), "%"))

```

```{r}

# Remove rows with 'unknown' values
data_clean_downsampled_no_unknown <- data_clean_downsampled %>%
  filter_all(~ . != "unknown")

# Check the new size of the dataset
print(dim(data_clean_downsampled_no_unknown))

```

```{r}

# Load necessary libraries
library(dplyr)

# Remove rows with 'unknown' values from the downsampled data
data_clean_downsampled_no_unknown <- data_clean_downsampled %>%
  filter_all(~ . != "unknown")

# Check the distribution of the target variable 'y' after removing unknowns
balance_after_cleaning <- table(data_clean_downsampled_no_unknown$Class)  # Assuming 'Class' is the name of the target variable column

# Print the class balance
print(balance_after_cleaning)

# Calculate and display the proportion of each class
balance_proportion <- prop.table(balance_after_cleaning) * 100
print(balance_proportion)

```

```{r}

# Export the dataset to a CSV file in the same location
write_csv(data, here("data_clean_downsampled_no_unknown .csv"))

# Confirm the file is saved by showing the file path
cat("Data has been exported to:", here("data_clean_downsampled_no_unknown .csv"))
```

```{r}

df <- data_clean_downsampled_no_unknown

# Rename the column 'Class' to 'y'
df <- df %>% rename(y = Class)

# Identify categorical and numeric variables
variables <- names(df)
var_types <- ifelse(sapply(df, is.numeric), "Numeric", "Categorical")

# Create a data frame to store variable names and their types
var_table <- data.frame(Variable = variables, Type = var_types)

# Create a flextable
flex_table <- flextable(var_table)

# Apply custom formatting:
# Highlight "Categorical" types with light blue, and "Numeric" types with light pink
flex_table <- flextable::bg(flex_table, j = "Type", i = ~ Type == "Categorical", bg = "lightblue")  # Highlight categorical
flex_table <- flextable::bg(flex_table, j = "Type", i = ~ Type == "Numeric", bg = "lightpink")      # Highlight numeric

# Adjust column widths for better readability
flex_table <- autofit(flex_table)

# Display the flextable
flex_table

```

```{r}

# Set up a 3x3 plotting area
par(mfrow=c(3,3))

# Identify categorical variables (factor or character type)
categorical_vars <- sapply(df, function(x) is.factor(x) | is.character(x))  
categorical_data <- df[, categorical_vars]  # Subset the dataframe for categorical variables

# Loop through all categorical variables and plot bar plots
for (var in names(categorical_data)) {
  barplot(table(categorical_data[[var]]), main=paste("Bar Plot of", var), xlab=var, col="lightblue")
}

# Reset the plotting layout to 1x1 after plotting
par(mfrow=c(1,1))

```

```{r}

# Set up a 3x3 plotting area
par(mfrow=c(3,3))

# Loop through all the columns and plot histograms for continuous (numeric) variables
numeric_vars <- sapply(df, is.numeric)  # Identify numeric variables
continuous_vars <- df[, numeric_vars]   # Subset the dataframe for numeric variables

# Plot histograms for each numeric variable
for (var in names(continuous_vars)) {
  hist(continuous_vars[[var]], main=paste("Histogram of", var), xlab=var, col="lightblue", breaks=20)
}

# Reset the plotting layout to 1x1 after plotting
par(mfrow=c(1,1))

```

```{r}

# Set up a 3x3 plotting area
par(mfrow=c(1,2))

# Identify numeric variables in the dataset
numeric_vars <- sapply(df, is.numeric)

# Subset the dataset to include only the numeric variables
numeric_data <- df[, numeric_vars]

# Loop through the numeric variables and create boxplots
for (var in names(numeric_data)) {
  boxplot(numeric_data[[var]], main=paste("Boxplot of", var), xlab=var, col="lightblue")
}

# Reset the plotting layout to 1x1 after plotting

```

```{r}

library(corrplot)

# Select only the numeric variables from the dataset
numeric_vars <- df[, sapply(df, is.numeric)]

# Create the correlation matrix
cor_matrix <- cor(numeric_vars, use="complete.obs")

# Visualize the correlation matrix
corrplot(cor_matrix, method="circle", type="lower", 
         tl.col="black", tl.cex=0.8, title="Correlation Matrix of Numeric Variables",
         mar=c(0,0,1,0))

```
### Key Assumptions of Logistic Regression:

1. **Binary Outcome Variable**: The dependent variable should be binary.
2. **Independence of Observations**: Observations should be independent of each other.
3. **No Multicollinearity**: Predictor variables should not be highly correlated with each other.
4. **Linearity of Independent Variables and Log-Odds**: There should be a linear relationship between continuous predictors and the log-odds of the outcome.
5. **Sufficient Sample Size**: Logistic regression requires a large sample size to provide reliable results.

### Steps to Test the Assumptions

### 1. **Check for Binary Outcome Variable**
Ensure the dependent variable (`y`) is binary.

```{r}
# Check the levels of the outcome variable
table(df$y)
```

```{r}

# Check for variables with only one level
lapply(df, function(x) length(unique(x)))

```

```{r}

# Set a seed for reproducibility
set.seed(123)

# Split the data into training (70%) and testing (30%) sets
trainIndex <- createDataPartition(df$y, p = 0.7, list = FALSE)
train_data <- df[trainIndex, ]  # 70% training data
test_data  <- df[-trainIndex, ] # 30% test data

```

```{r}

# Fit the logistic regression model on the training set
logit_model_train <- glm(y ~ age + job + marital + education + housing + loan + contact + month + 
                         day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                         cons.price.idx + cons.conf.idx, 
                         family = binomial, data = train_data)

# Display the summary of the model
summary(logit_model_train)

```

- **Interpretation**: Variables with VIF > 5 may have multicollinearity issues.

```{r}

# Calculate VIF values for the logistic regression model
vif_values_train <- vif(logit_model_train)

# Print the VIF values
print(vif_values_train)

```

```{r}

# Example: Remove a variable with high VIF and refit the model
logit_model_train_refit <- glm(y ~ age + job + marital + education + housing + loan + contact + month + 
                               day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                               cons.price.idx, 
                               family = binomial, data = train_data)

# Recheck VIF values for the refit model
vif_values_refit <- vif(logit_model_train_refit)
print(vif_values_refit)

```

```{r}
# Check the number of events (e.g., 1's and 0's in the outcome variable)
table(train_data$y)

# Ensure the number of events is at least 10 times the number of predictors
```

```{r}

# Predict probabilities for the training data using the logistic regression model
predicted_probabilities_train <- predict(logit_model_train, newdata = train_data, type = "response")

# Convert probabilities to binary outcome (using 0.5 as the cutoff)
predicted_classes_train <- ifelse(predicted_probabilities_train > 0.5, 1, 0)

# Create confusion matrix for training data
confusion_matrix_train <- table(predicted_classes_train, train_data$y)
print(confusion_matrix_train)

# Extract the values from the confusion matrix
TN <- confusion_matrix_train[1,1]  # True Negatives
FP <- confusion_matrix_train[1,2]  # False Positives
FN <- confusion_matrix_train[2,1]  # False Negatives
TP <- confusion_matrix_train[2,2]  # True Positives

# Calculate accuracy
accuracy_train <- (TP + TN) / sum(confusion_matrix_train)

# Calculate precision, recall, and F1 score
precision_train <- ifelse((TP + FP) > 0, TP / (TP + FP), 0)  # TP / (TP + FP)
recall_train <- ifelse((TP + FN) > 0, TP / (TP + FN), 0)     # TP / (TP + FN)
f1_score_train <- ifelse((precision_train + recall_train) > 0, 
                         2 * ((precision_train * recall_train) / (precision_train + recall_train)), 
                         0)

# Print the performance metrics for the training data
print(paste("Accuracy:", accuracy_train))
print(paste("Precision:", precision_train))
print(paste("Recall:", recall_train))
print(paste("F1 Score:", f1_score_train))

```

```{r}

accuracy <- 0.7722   
precision <- 0.7259  
recall <- 0.8174     
f1_score <- 0.7689

# Create a dataframe with the performance metrics
metrics_data <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score"),
  Value = c(accuracy, precision, recall, f1_score)
)

# Create the flextable
performance_table <- flextable(metrics_data)

# Apply consistent styling to the flextable with alternating colors
performance_table <- performance_table %>%
  color(j = 1, color = "black") %>%                        # Text color for Metric column
  color(j = 2, color = "darkblue") %>%                     # Text color for Value column
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Accuracy") %>% # Background color for Accuracy
  bg(part = "body", bg = "lightblue", i = ~ Metric == "Precision") %>% # Background color for Precision
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Recall") %>%   # Background color for Recall
  bg(part = "body", bg = "lightblue", i = ~ Metric == "F1 Score") %>% # Background color for F1 Score
  align(j = 2, align = "center", part = "body") %>%        # Center-align the values
  autofit()                                                # Adjust column widths

# Print the flextable
performance_table

```

```{r}

# Predict probabilities for the training set using the logistic regression model
predicted_probabilities_train <- predict(logit_model_train, newdata = train_data, type = "response")

# Create the ROC curve using the training data
roc_curve_train <- roc(train_data$y, predicted_probabilities_train)

# Plot the ROC curve for the training data
plot(roc_curve_train, main = "ROC Curve for Logistic Regression Model (Training Data)", col = "blue", lwd = 2)

# Calculate the AUC for the training data
auc_train <- auc(roc_curve_train)

# Print the AUC value for the training data
print(paste("AUC for Training Data:", auc_train))

```

```{r}

# Predict probabilities for the test data using the logistic regression model
predicted_probabilities_test <- predict(logit_model_train, newdata = test_data, type = "response")

# Convert probabilities to binary outcomes (using 0.5 as the cutoff)
predicted_classes_test <- ifelse(predicted_probabilities_test > 0.5, 1, 0)

# Create confusion matrix for the test data
confusion_matrix_test <- table(predicted_classes_test, test_data$y)
print(confusion_matrix_test)

# Extract values from the confusion matrix for test data
TN_test <- confusion_matrix_test[1,1]  # True Negatives
FP_test <- confusion_matrix_test[1,2]  # False Positives
FN_test <- confusion_matrix_test[2,1]  # False Negatives
TP_test <- confusion_matrix_test[2,2]  # True Positives

# Calculate accuracy for the test data
accuracy_test <- (TP_test + TN_test) / sum(confusion_matrix_test)

# Calculate precision, recall, and F1 score for the test data
precision_test <- ifelse((TP_test + FP_test) > 0, TP_test / (TP_test + FP_test), 0)
recall_test <- ifelse((TP_test + FN_test) > 0, TP_test / (TP_test + FN_test), 0)
f1_score_test <- ifelse((precision_test + recall_test) > 0, 
                        2 * ((precision_test * recall_test) / (precision_test + recall_test)), 
                        0)

# Print the performance metrics for the test data
print(paste("Accuracy (Test):", accuracy_test))
print(paste("Precision (Test):", precision_test))
print(paste("Recall (Test):", recall_test))
print(paste("F1 Score (Test):", f1_score_test))

```

```{r}

# Checking for multicollinearity (VIF values) - same as training data, but for the model
vif_values_test <- vif(logit_model_train)  # VIF is the same as in training
print(vif_values_test)

```

```{r}

# Create the ROC curve for the test data
roc_curve_test <- roc(test_data$y, predicted_probabilities_test)

# Plot the ROC curve for the test data
plot(roc_curve_test, main = "ROC Curve for Logistic Regression Model (Test Data)", col = "blue", lwd = 2)

# Calculate the AUC for the test data
auc_test <- auc(roc_curve_test)

# Print the AUC value for the test data
print(paste("AUC for Test Data:", auc_test))

```
```{r}

# Create a dataframe for comparison
metrics_comparison <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "AUC"),
  Training = c(0.7721, 0.7259, 0.8174, 0.7689, 0.8531),
  Test = c(0.7406, 0.6846, 0.7917, 0.7343, 0.8183)
)

# Create the flextable
comparison_table <- flextable(metrics_comparison)

# Format the flextable with some custom styles
comparison_table <- comparison_table %>%
  color(j = 1, color = "black") %>%                        # Text color for Metric column
  color(j = 2:3, color = "darkblue") %>%                   # Text color for values in Training and Test columns
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Accuracy") %>%
  bg(part = "body", bg = "lightblue", i = ~ Metric == "Precision") %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Recall") %>%
  bg(part = "body", bg = "lightblue", i = ~ Metric == "F1 Score") %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "AUC") %>%
  align(j = 2:3, align = "center", part = "body") %>%      # Center-align the values
  autofit()                                                # Adjust column widths

# Print the flextable
comparison_table

```

```{r}

# Fit the Random Forest model
set.seed(123)  # Set a seed for reproducibility

rf_model <- randomForest(y ~ age + job + marital + education + housing + loan + contact + month + 
                         day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                         cons.price.idx + cons.conf.idx, 
                         data = df, ntree = 500, mtry = 3, importance = TRUE)

# Print the model summary
print(rf_model)

```

```{r}

# Predict the class (0/1) for the test data
predicted_rf <- predict(rf_model, df)

# Create confusion matrix
conf_matrix_rf <- table(predicted_rf, df$y)

# Print confusion matrix
print(conf_matrix_rf)

# Calculate accuracy
accuracy_rf <- mean(predicted_rf == df$y)
print(paste("Random Forest Accuracy:", accuracy_rf))

```

```{r}

# Get predicted probabilities from the Random Forest model
predicted_probabilities_rf <- predict(rf_model, df, type = "prob")[,2]

# Plot the ROC curve
roc_curve_rf <- roc(df$y, predicted_probabilities_rf)
plot(roc_curve_rf)

# Calculate the AUC
auc_rf <- auc(roc_curve_rf)
print(paste("AUC for Random Forest:", auc_rf))

```
```{r}

# Get variable importance
importance_rf <- importance(rf_model)

# Plot variable importance
varImpPlot(rf_model)

```

```{r}

# Set a seed for reproducibility
set.seed(123)

# Train the Random Forest model on the training data
rf_model_train <- randomForest(y ~ age + job + marital + education + housing + loan + contact + month +
                               day_of_week + campaign + pdays + previous + poutcome + emp.var.rate +
                               cons.price.idx + cons.conf.idx,
                               data = train_data, ntree = 500, mtry = 3, importance = TRUE)

# Print the summary of the Random Forest model (training data)
print(rf_model_train)

```

```{r}

# Manually input the confusion matrix values
TN_rf_train <- 177  # True Negatives
FP_rf_train <- 60   # False Positives
FN_rf_train <- 85   # False Negatives
TP_rf_train <- 174  # True Positives

# Calculate accuracy
accuracy_rf_train <- (TP_rf_train + TN_rf_train) / (TP_rf_train + TN_rf_train + FP_rf_train + FN_rf_train)
print(paste("Accuracy (Training):", accuracy_rf_train))

# Calculate precision, recall, and F1 score, handling division by zero
precision_rf_train <- ifelse((TP_rf_train + FP_rf_train) > 0, TP_rf_train / (TP_rf_train + FP_rf_train), 0)
recall_rf_train <- ifelse((TP_rf_train + FN_rf_train) > 0, TP_rf_train / (TP_rf_train + FN_rf_train), 0)
f1_score_rf_train <- ifelse((precision_rf_train + recall_rf_train) > 0,
                            2 * ((precision_rf_train * recall_rf_train) / (precision_rf_train + recall_rf_train)),
                            0)

# Print the metrics for the training data
print(paste("Precision (Training):", precision_rf_train))
print(paste("Recall (Training):", recall_rf_train))
print(paste("F1 Score (Training):", f1_score_rf_train))

```
```{r}

# Manually input the confusion matrix values for test data
TN_rf_test <- 82  # True Negatives
FP_rf_test <- 40  # False Positives
FN_rf_test <- 19  # False Negatives
TP_rf_test <- 71  # True Positives

# Calculate accuracy for test data
accuracy_rf_test <- (TP_rf_test + TN_rf_test) / (TP_rf_test + TN_rf_test + FP_rf_test + FN_rf_test)
print(paste("Accuracy (Test):", accuracy_rf_test))

# Calculate precision, recall, and F1 score, handling division by zero
precision_rf_test <- ifelse((TP_rf_test + FP_rf_test) > 0, TP_rf_test / (TP_rf_test + FP_rf_test), 0)
recall_rf_test <- ifelse((TP_rf_test + FN_rf_test) > 0, TP_rf_test / (TP_rf_test + FN_rf_test), 0)
f1_score_rf_test <- ifelse((precision_rf_test + recall_rf_test) > 0,
                           2 * ((precision_rf_test * recall_rf_test) / (precision_rf_test + recall_rf_test)),
                           0)

# Print the metrics for the test data
print(paste("Precision (Test):", precision_rf_test))
print(paste("Recall (Test):", recall_rf_test))
print(paste("F1 Score (Test):", f1_score_rf_test))

```

```{r}

# Predict probabilities for the ROC and AUC on the training data
predicted_probabilities_rf_train <- predict(rf_model_train, newdata = train_data, type = "prob")[, 2]
roc_curve_rf_train <- roc(train_data$y, predicted_probabilities_rf_train)
auc_rf_train <- auc(roc_curve_rf_train)

# Predict probabilities for the ROC and AUC on the test data
predicted_probabilities_rf_test <- predict(rf_model_train, newdata = test_data, type = "prob")[, 2]
roc_curve_rf_test <- roc(test_data$y, predicted_probabilities_rf_test)
auc_rf_test <- auc(roc_curve_rf_test)

# Print AUC for both datasets
print(paste("AUC (Training):", auc_rf_train))
print(paste("AUC (Test):", auc_rf_test))

```

```{r}

# Create a dataframe for Random Forest comparison
rf_metrics_comparison <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "AUC"),
  Training = c(accuracy_rf_train, precision_rf_train, recall_rf_train, f1_score_rf_train, auc_rf_train),
  Test = c(accuracy_rf_test, precision_rf_test, recall_rf_test, f1_score_rf_test, auc_rf_test)
)

# Create the flextable for Random Forest model comparison
rf_comparison_table <- flextable(rf_metrics_comparison)

# Format the flextable with some custom styles
rf_comparison_table <- rf_comparison_table %>%
  color(j = 1, color = "black") %>%
  color(j = 2:3, color = "darkblue") %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Accuracy") %>%
  bg(part = "body", bg = "lightblue", i = ~ Metric == "Precision") %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Recall") %>%
  bg(part = "body", bg = "lightblue", i = ~ Metric == "F1 Score") %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "AUC") %>%
  align(j = 2:3, align = "center", part = "body") %>%
  autofit()

# Print the flextable
rf_comparison_table

```

```{r}

# Set up 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Train the logistic regression model using 10-fold cross-validation
logit_model_cv <- train(y ~ age + job + marital + education + housing + loan + contact + month + 
                        day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                        cons.price.idx + cons.conf.idx,
                        data = df, method = "glm", family = binomial,
                        trControl = train_control)

# Print the results of the cross-validation
print(logit_model_cv)

```

```{r}

# Set up 10-fold cross-validation
train_control_rf <- trainControl(method = "cv", number = 10)

# Train the Random Forest model using 10-fold cross-validation on the training data
rf_model_cv <- train(y ~ age + job + marital + education + housing + loan + contact + month + 
                     day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                     cons.price.idx + cons.conf.idx,
                     data = train_data, method = "rf", 
                     ntree = 500,  # Set the number of trees
                     trControl = train_control_rf, importance = TRUE)

# Print the results of the cross-validation
print(rf_model_cv)

```
```{r}

# Set up 10-fold cross-validation for the test data
train_control_test <- trainControl(method = "cv", number = 10)

# Perform logistic regression with 10-fold cross-validation on the test data
logit_model_cv_test <- train(y ~ age + job + marital + education + housing + loan + contact + month + 
                             day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                             cons.price.idx + cons.conf.idx,
                             data = test_data, method = "glm", family = binomial,
                             trControl = train_control_test)

# Print the cross-validation results for the test data
print(logit_model_cv_test)

# Predict the probabilities on the test data
predicted_probabilities_test_cv <- predict(logit_model_cv_test, newdata = test_data, type = "prob")[, 2]

# Convert probabilities to binary outcome (using 0.5 as the cutoff)
predicted_classes_test_cv <- ifelse(predicted_probabilities_test_cv > 0.5, 1, 0)

# Create a confusion matrix for the test data
confusion_matrix_test_cv <- table(predicted_classes_test_cv, test_data$y)
print(confusion_matrix_test_cv)

# Manually input the confusion matrix values
TN_test_cv <- 82  # True Negatives
FP_test_cv <- 24  # False Positives
FN_test_cv <- 19  # False Negatives
TP_test_cv <- 87  # True Positives

# Calculate accuracy for the test data
accuracy_test_cv <- (TP_test_cv + TN_test_cv) / (TP_test_cv + TN_test_cv + FP_test_cv + FN_test_cv)
print(paste("Accuracy (Test with CV):", accuracy_test_cv))

# Calculate precision, recall, and F1 score, handling division by zero
precision_test_cv <- ifelse((TP_test_cv + FP_test_cv) > 0, TP_test_cv / (TP_test_cv + FP_test_cv), 0)
recall_test_cv <- ifelse((TP_test_cv + FN_test_cv) > 0, TP_test_cv / (TP_test_cv + FN_test_cv), 0)
f1_score_test_cv <- ifelse((precision_test_cv + recall_test_cv) > 0,
                           2 * ((precision_test_cv * recall_test_cv) / (precision_test_cv + recall_test_cv)),
                           0)

# Print the metrics for the test data
print(paste("Precision (Test with CV):", precision_test_cv))
print(paste("Recall (Test with CV):", recall_test_cv))
print(paste("F1 Score (Test with CV):", f1_score_test_cv))

```

```{r}

# Set up 10-fold cross-validation for the test data
train_control_rf_test <- trainControl(method = "cv", number = 10)

# Perform Random Forest with 10-fold cross-validation on the test data
rf_model_cv_test <- train(y ~ age + job + marital + education + housing + loan + contact + month + 
                          day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                          cons.price.idx + cons.conf.idx,
                          data = test_data, method = "rf", 
                          ntree = 500,  # Number of trees in the Random Forest
                          trControl = train_control_rf_test)

# Print the cross-validation results for the test data
print(rf_model_cv_test)

# Predict the classes on the test data using the cross-validated Random Forest model
predicted_rf_test_cv <- predict(rf_model_cv_test, newdata = test_data)

# Create a confusion matrix for the test data
confusion_matrix_rf_test_cv <- table(predicted_rf_test_cv, test_data$y)
print(confusion_matrix_rf_test_cv)

# Manually input the confusion matrix values
TN_rf_test_cv <- 91  # True Negatives
FP_rf_test_cv <- 26  # False Positives
FN_rf_test_cv <- 10  # False Negatives
TP_rf_test_cv <- 85  # True Positives

# Calculate accuracy for the test data
accuracy_rf_test_cv <- (TP_rf_test_cv + TN_rf_test_cv) / (TP_rf_test_cv + TN_rf_test_cv + FP_rf_test_cv + FN_rf_test_cv)
print(paste("Accuracy (Test with CV - Random Forest):", accuracy_rf_test_cv))

# Calculate precision, recall, and F1 score, handling division by zero
precision_rf_test_cv <- ifelse((TP_rf_test_cv + FP_rf_test_cv) > 0, TP_rf_test_cv / (TP_rf_test_cv + FP_rf_test_cv), 0)
recall_rf_test_cv <- ifelse((TP_rf_test_cv + FN_rf_test_cv) > 0, TP_rf_test_cv / (TP_rf_test_cv + FN_rf_test_cv), 0)
f1_score_rf_test_cv <- ifelse((precision_rf_test_cv + recall_rf_test_cv) > 0,
                              2 * ((precision_rf_test_cv * recall_rf_test_cv) / (precision_rf_test_cv + recall_rf_test_cv)),
                              0)

# Print the metrics for the test data
print(paste("Precision (Test with CV - Random Forest):", precision_rf_test_cv))
print(paste("Recall (Test with CV - Random Forest):", recall_rf_test_cv))
print(paste("F1 Score (Test with CV - Random Forest):", f1_score_rf_test_cv))

```
```{r}

# ---------------------- Logistic Regression with CV ----------------------

# Set up 10-fold cross-validation for the logistic regression on the test data
train_control_logit_test <- trainControl(method = "cv", number = 10)

# Perform logistic regression with 10-fold cross-validation on the test data
logit_model_cv_test <- train(y ~ age + job + marital + education + housing + loan + contact + month + 
                             day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                             cons.price.idx + cons.conf.idx,
                             data = test_data, method = "glm", family = binomial,
                             trControl = train_control_logit_test)

# Predict probabilities for ROC curve on the test data
predicted_probabilities_logit_cv_test <- predict(logit_model_cv_test, newdata = test_data, type = "prob")[, 2]

# Calculate the ROC curve and AUC for logistic regression on the test data
roc_logit_test <- roc(test_data$y, predicted_probabilities_logit_cv_test)
auc_logit_test <- auc(roc_logit_test)

# Print the AUC for logistic regression on the test data
print(paste("AUC (Logistic Regression with CV - Test Data):", auc_logit_test))

# Plot the ROC curve for logistic regression on the test data
plot(roc_logit_test, main = "ROC Curve - Logistic Regression (Test Data)", col = "blue")

# ---------------------- Random Forest with CV ----------------------

# Set up 10-fold cross-validation for the random forest on the test data
train_control_rf_test <- trainControl(method = "cv", number = 10)

# Perform Random Forest with 10-fold cross-validation on the test data
rf_model_cv_test <- train(y ~ age + job + marital + education + housing + loan + contact + month + 
                          day_of_week + campaign + pdays + previous + poutcome + emp.var.rate + 
                          cons.price.idx + cons.conf.idx,
                          data = test_data, method = "rf", 
                          ntree = 500,  # Number of trees in the Random Forest
                          trControl = train_control_rf_test)

# Predict probabilities for ROC curve on the test data
predicted_probabilities_rf_cv_test <- predict(rf_model_cv_test, newdata = test_data, type = "prob")[, 2]

# Calculate the ROC curve and AUC for Random Forest on the test data
roc_rf_test <- roc(test_data$y, predicted_probabilities_rf_cv_test)
auc_rf_test <- auc(roc_rf_test)

# Print the AUC for Random Forest on the test data
print(paste("AUC (Random Forest with CV - Test Data):", auc_rf_test))

# Plot the ROC curve for Random Forest on the test data
plot(roc_rf_test, main = "ROC Curve - Random Forest (Test Data)", col = "red")

```
```{r}

# Logistic Regression Cross-Validation Results (updated from the images)
accuracy_logit <- 0.7972  # Accuracy
precision_logit <- 0.7384  # Precision
recall_logit <- 0.8208  # Recall
f1_score_logit <- 0.8018  # F1 Score
auc_logit_test <- 0.8907  # AUC for logistic regression

# Random Forest Cross-Validation Results (updated from the images)
accuracy_rf <- 0.8302  # Accuracy
precision_rf <- 0.7658  # Precision
recall_rf <- 0.8947  # Recall
f1_score_rf <- 0.8254  # F1 Score
auc_rf_test <- 0.9395  # AUC for random forest

# Create a dataframe for comparison
metrics_comparison <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1 Score", "AUC"),
  Logistic_Regression = c(accuracy_logit, precision_logit, recall_logit, f1_score_logit, auc_logit_test),
  Random_Forest = c(accuracy_rf, precision_rf, recall_rf, f1_score_rf, auc_rf_test)
)

# Create the flextable for model comparison
comparison_table <- flextable(metrics_comparison)

# Format the flextable with custom styles
comparison_table <- comparison_table %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Accuracy") %>%
  bg(part = "body", bg = "lightblue", i = ~ Metric == "Precision") %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "Recall") %>%
  bg(part = "body", bg = "lightblue", i = ~ Metric == "F1 Score") %>%
  bg(part = "body", bg = "lightgray", i = ~ Metric == "AUC") %>%
  align(j = 2:3, align = "center", part = "body") %>%
  autofit()

# Print the flextable
comparison_table

# ---------------------- Bar Graph Comparison ----------------------

# Reshape the data for plotting
metrics_long <- reshape2::melt(metrics_comparison, id.vars = "Metric", variable.name = "Model", value.name = "Value")

# Create the bar graph
ggplot(metrics_long, aes(x = Metric, y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Logistic_Regression" = "blue", "Random_Forest" = "red")) +
  labs(title = "Model Comparison - Logistic Regression vs Random Forest", 
       x = "Metric", y = "Value", fill = "Model") +
  theme_minimal()

```
### Analytic Report: Logistic Regression vs Random Forest (Cross-Validation Results)

This report compares the performance of two models: **Logistic Regression** and **Random Forest**, using cross-validation on the test data. We evaluated the models based on the following metrics: **Accuracy**, **Precision**, **Recall**, **F1 Score**, and **AUC (Area Under the Curve)**.

#### Overall Findings:

- The **Random Forest model** consistently outperformed **Logistic Regression** in almost all metrics.
- **Random Forest** showed better generalization performance, particularly in recall and AUC, which are key indicators of a model's ability to distinguish between classes and identify positive cases.
- While **Logistic Regression** had slightly lower scores, it remained competitive, especially considering its simplicity compared to Random Forest.

#### Metric Comparisons:

1. **Accuracy**:
   - Logistic Regression: **0.7972**
   - Random Forest: **0.8302**
   - **Analysis**: Random Forest had a higher accuracy, indicating it made fewer overall errors in classifying the test data compared to Logistic Regression.

2. **Precision**:
   - Logistic Regression: **0.7384**
   - Random Forest: **0.7658**
   - **Analysis**: Precision measures the proportion of correctly predicted positive observations. Random Forest performed slightly better, meaning it had fewer false positives.

3. **Recall**:
   - Logistic Regression: **0.8208**
   - Random Forest: **0.8947**
   - **Analysis**: Random Forest had a much higher recall, indicating it correctly identified a larger proportion of actual positive cases. This makes Random Forest more reliable for detecting positive instances.

4. **F1 Score**:
   - Logistic Regression: **0.8018**
   - Random Forest: **0.8254**
   - **Analysis**: The F1 score, which balances precision and recall, shows that Random Forest had better overall performance in balancing the two metrics.

5. **AUC (Area Under the Curve)**:
   - Logistic Regression: **0.8907**
   - Random Forest: **0.9395**
   - **Analysis**: The AUC measures the model's ability to distinguish between positive and negative classes. A higher AUC means better discriminatory power. Random Forest had a significantly higher AUC, suggesting it is better at separating the classes.

#### Key Observations:

- **Random Forest excels in recall**: Its ability to detect more positive cases (higher recall) makes it suitable for applications where false negatives are costly.
- **Logistic Regression remains competitive**: Despite being outperformed by Random Forest, Logistic Regression achieved reasonably good results. Its simplicity and interpretability make it a solid choice for applications where model transparency is important.
- **Precision vs Recall Tradeoff**: Random Forest showed a stronger recall, which may indicate that it is more aggressive in predicting positives, even at the cost of some false positives. This is useful in scenarios where missing positive cases (false negatives) are more critical than misclassifying negatives.

#### Conclusion:
- **Random Forest** is the preferred model based on its superior performance across all metrics, particularly in recall and AUC.
- **Logistic Regression** still offers good performance and can be chosen in situations where model interpretability or simplicity is more important than slight gains in accuracy or recall.

### 2024-12-03 R Projects/CAPM CASE STUDY 3.qmd
updated assignments to github

---
title: "Case Study Down Jones"
author: "Leonel Salazar, Seth Harris, Joaquin Ramirez, Collin Real"
format: docx
---

```{r echo=FALSE}

library(tidyverse)
library(here)
library(ggplot2)
library(gridExtra)
library(tseries)
library(dplyr)
library(quantmod)

here()

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Use the here function to construct the file path and import the dataset
data <- read.csv(here("dow_jones_index.data.csv"), header = TRUE, sep = ",")

head(data)

str(data)

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Check for missing data or NA values in the dataset

missing_data_summary <- colSums(is.na(data))
missing_data_summary
```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# removed.

data_clean <- data

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Convert 'quarter' variable to a factor
data_clean$quarter <- as.factor(data_clean$quarter)

# Convert 'stock' variable to a factor
data_clean$stock <- as.factor(data_clean$stock)

# Convert 'date' variable to Date class format
data_clean$date <- as.Date(data_clean$date, format = "%m/%d/%Y")

# Remove non-numeric characters from 'open', 'high', 'low', and 'close' before converting to numeric
data_clean$open <- as.numeric(gsub("[^0-9.]", "", data_clean$open))
data_clean$high <- as.numeric(gsub("[^0-9.]", "", data_clean$high))
data_clean$low <- as.numeric(gsub("[^0-9.]", "", data_clean$low))
data_clean$close <- as.numeric(gsub("[^0-9.]", "", data_clean$close))

# Convert 'volume' variable to numeric
data_clean$volume <- as.numeric(gsub("[^0-9.]", "", data_clean$volume))

# # Convert 'percent_change_price' to numeric by removing '%' and dividing by 100
# data_clean$percent_change_price <- as.numeric(gsub("%", "", data_clean$percent_change_price)) / 100
# 
# Convert 'percent_change_volume_over_last_wk' to numeric by removing '%'
data_clean$percent_change_volume_over_last_wk <- as.numeric(gsub("%", "", data_clean$percent_change_volume_over_last_wk))

# Remove non-numeric characters from 'previous_weeks_volume', 'next_weeks_open', and 'next_weeks_close' before converting to numeric
data_clean$previous_weeks_volume <- as.numeric(gsub("[^0-9.]", "", data_clean$previous_weeks_volume))
data_clean$next_weeks_open <- as.numeric(gsub("[^0-9.]", "", data_clean$next_weeks_open))
data_clean$next_weeks_close <- as.numeric(gsub("[^0-9.]", "", data_clean$next_weeks_close))

# # Convert 'percent_change_next_weeks_price' to numeric by removing '%' and dividing by 100
# data_clean$percent_change_next_weeks_price <- as.numeric(gsub("%", "", data_clean$percent_change_next_weeks_price)) / 100

# Convert 'days_to_next_dividend' to numeric
data_clean$days_to_next_dividend <- as.numeric(gsub("[^0-9.]", "", data_clean$days_to_next_dividend))

str(data_clean)

```

```{r echo=FALSE, warning=FALSE, error=FALSE}
# Step 1: Inspect the unique stock symbols
unique_stocks <- unique(data_clean$stock) # Adjust column name if different

# Step 2: Randomly select 5 unique stocks
set.seed(123) # Set seed for reproducibility
selected_stocks <- sample(unique_stocks, 30)

# Step 3: Create a directory for each stock and save its data
for (stock in selected_stocks) {
  # Filter data for the current stock
  stock_data <- data_clean %>%
    filter(stock == !!stock) %>%
    select(stock,date, open, high, low, close, volume) # Adjust column names to match your dataset

   # Create file path for each stock
  file_name <- here( paste0("stock_", stock, ".csv"))

  # Save the stock data as a CSV file within its folder
  file_name <- here( paste0("stock_", stock, ".csv"))
  write.csv(stock_data, file_name, row.names = FALSE)
}
```

```{r echo=FALSE, warning=FALSE, error=FALSE}
# Step 1: Extract unique stock names from the main dataset
stock_symbols <- unique(data_clean$stock) # Ensure 'data_clean' is already processed

# Step 2: Load data for each stock into individual variables
for (symbol in stock_symbols) {
  # Construct file name dynamically
  file_path <- here(paste0("stock_", symbol, ".csv"))

  # Check if the file exists
  if (file.exists(file_path)) {
    # Read the stock data
    assign(symbol, read.csv(file_path, header = TRUE, sep = ","))
    cat("Loaded data for stock:", symbol, "\n")
  } else {
    warning(paste("File not found for stock:", symbol))
  }
}

# Example: Check the structure of one dataset (e.g., AAPL)
if (exists("AAPL")) {
  str(AAPL)
}

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Step 2: Select only the relevant variables
clean_stock <- data_clean %>%
  select(stock,date, open, high, low, close, volume) # Adjust column names to match your dataset

# Step 3: Save the cleaned dataset as dowjones.csv
output_path <- here("clean_stock.csv")
write.csv(data_clean, output_path, row.names = FALSE)

clean_stock <- read.csv(here("clean_stock.csv"), header = TRUE, sep = ",")

str(clean_stock)
```

```{r echo=FALSE, warning=FALSE, error=FALSE}

str(GE)

str(clean_stock)

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Convert 'date' column to Date format and numeric conversion for the other columns
clean_stock <- data %>%
  mutate(
    date = as.Date(date, format = "%m/%d/%Y"), # Convert 'date' to Date format
    open = as.numeric(gsub("\\$", "", open)),  # Remove dollar sign and convert to numeric
    high = as.numeric(gsub("\\$", "", high)),  # Remove dollar sign and convert to numeric
    low = as.numeric(gsub("\\$", "", low)),    # Remove dollar sign and convert to numeric
    close = as.numeric(gsub("\\$", "", close)),# Remove dollar sign and convert to numeric
    volume = as.numeric(volume)                # Ensure 'volume' is numeric
  )

# Remove the 'stock' column
clean_stock <- clean_stock %>%
  select(date, open, high, low, close, volume)

str(clean_stock)
```

```{r}

library(dplyr)

# Ensure columns are properly formatted
clean_stock <- clean_stock %>%
  mutate(
    date = as.Date(date, format = "%m/%d/%Y"),
    open = as.numeric(gsub("\\$", "", open)),
    high = as.numeric(gsub("\\$", "", high)),
    low = as.numeric(gsub("\\$", "", low)),
    close = as.numeric(gsub("\\$", "", close)),
    volume = as.numeric(volume)
  )

# Check for issues in data conversion
print(summary(data))  # Provides an overview of the data

# Group by 'date' and calculate averages, handling NAs
averaged_data <- clean_stock %>%
  group_by(date) %>%
  summarize(
    avg_open = mean(open, na.rm = TRUE),
    avg_high = mean(high, na.rm = TRUE),
    avg_low = mean(low, na.rm = TRUE),
    avg_close = mean(close, na.rm = TRUE),
    avg_volume = mean(volume, na.rm = TRUE)
  )

# View the cleaned data
print(head(averaged_data))
```
```{r echo=FALSE, warning=FALSE, error=FALSE}

# Step 1: Define stock symbols
stock_symbols <- unique(data_clean$stock)

# Step 2: Remove the 'stock' column for each dataset dynamically
for (symbol in stock_symbols) {
  # Check if the dataset exists in the environment
  if (exists(symbol)) {
    # Get the dataset
    stock_data <- get(symbol)

    # Remove the 'stock' column
    stock_data <- stock_data %>% select(-stock)

    # Save the modified dataset back to its original variable
    assign(symbol, stock_data)

    # Print confirmation
    cat("Removed 'stock' column for:", symbol, "\n")
  } else {
    warning(paste("Dataset not found for stock:", symbol))
  }
}

```

```{r echo=FALSE, warning=FALSE, error=FALSE}
here()

# Load the data files for all stocks
stock_AA <- read.csv(here("Week 6", "stock_AA.csv"), stringsAsFactors = FALSE)
stock_AXP <- read.csv(here("Week 6", "stock_AXP.csv"), stringsAsFactors = FALSE)
stock_BA <- read.csv(here("Week 6", "stock_BA.csv"), stringsAsFactors = FALSE)
stock_BAC <- read.csv(here("Week 6", "stock_BAC.csv"), stringsAsFactors = FALSE)
stock_CAT <- read.csv(here("Week 6", "stock_CAT.csv"), stringsAsFactors = FALSE)
stock_CSCO <- read.csv(here("Week 6", "stock_CSCO.csv"), stringsAsFactors = FALSE)
stock_CVX <- read.csv(here("Week 6", "stock_CVX.csv"), stringsAsFactors = FALSE)
stock_DIS <- read.csv(here("Week 6", "stock_DIS.csv"), stringsAsFactors = FALSE)
stock_DD <- read.csv(here("Week 6", "stock_DD.csv"), stringsAsFactors = FALSE)
stock_GE <- read.csv(here("Week 6", "stock_GE.csv"), stringsAsFactors = FALSE)
stock_HD <- read.csv(here("Week 6", "stock_HD.csv"), stringsAsFactors = FALSE)
stock_HPQ <- read.csv(here("Week 6", "stock_HPQ.csv"), stringsAsFactors = FALSE)
stock_IBM <- read.csv(here("Week 6", "stock_IBM.csv"), stringsAsFactors = FALSE)
stock_INTC <- read.csv(here("Week 6", "stock_INTC.csv"), stringsAsFactors = FALSE)
stock_JNJ <- read.csv(here("Week 6", "stock_JNJ.csv"), stringsAsFactors = FALSE)
stock_JPM <- read.csv(here("Week 6", "stock_JPM.csv"), stringsAsFactors = FALSE)
stock_KO <- read.csv(here("Week 6", "stock_KO.csv"), stringsAsFactors = FALSE)
stock_KRFT <- read.csv(here("Week 6", "stock_KRFT.csv"), stringsAsFactors = FALSE)
stock_MCD <- read.csv(here("Week 6", "stock_MCD.csv"), stringsAsFactors = FALSE)
stock_MMM <- read.csv(here("Week 6", "stock_MMM.csv"), stringsAsFactors = FALSE)
stock_MRK <- read.csv(here("Week 6", "stock_MRK.csv"), stringsAsFactors = FALSE)
stock_MSFT <- read.csv(here("Week 6", "stock_MSFT.csv"), stringsAsFactors = FALSE)
stock_PFE <- read.csv(here("Week 6", "stock_PFE.csv"), stringsAsFactors = FALSE)
stock_PG <- read.csv(here("Week 6", "stock_PG.csv"), stringsAsFactors = FALSE)
stock_TRV <- read.csv(here("Week 6", "stock_TRV.csv"), stringsAsFactors = FALSE)
stock_UTX <- read.csv(here("Week 6", "stock_UTX.csv"), stringsAsFactors = FALSE)
stock_VZ <- read.csv(here("Week 6", "stock_VZ.csv"), stringsAsFactors = FALSE)
stock_WMT <- read.csv(here("Week 6", "stock_WMT.csv"), stringsAsFactors = FALSE)
stock_XOM <- read.csv(here("Week 6", "stock_XOM.csv"), stringsAsFactors = FALSE)

# Save the cleaned data back to their respective files
write.csv(stock_AA, here("Week 6", "stock_AA.csv"), row.names = FALSE)
write.csv(stock_AXP, here("Week 6", "stock_AXP.csv"), row.names = FALSE)
write.csv(stock_BA, here("Week 6", "stock_BA.csv"), row.names = FALSE)
write.csv(stock_BAC, here("Week 6", "stock_BAC.csv"), row.names = FALSE)
write.csv(stock_CAT, here("Week 6", "stock_CAT.csv"), row.names = FALSE)
write.csv(stock_CSCO, here("Week 6", "stock_CSCO.csv"), row.names = FALSE)
write.csv(stock_CVX, here("Week 6", "stock_CVX.csv"), row.names = FALSE)
write.csv(stock_DIS, here("Week 6", "stock_DIS.csv"), row.names = FALSE)
write.csv(stock_DD, here("Week 6", "stock_DD.csv"), row.names = FALSE)
write.csv(stock_GE, here("Week 6", "stock_GE.csv"), row.names = FALSE)
write.csv(stock_HD, here("Week 6", "stock_HD.csv"), row.names = FALSE)
write.csv(stock_HPQ, here("Week 6", "stock_HPQ.csv"), row.names = FALSE)
write.csv(stock_IBM, here("Week 6", "stock_IBM.csv"), row.names = FALSE)
write.csv(stock_INTC, here("Week 6", "stock_INTC.csv"), row.names = FALSE)
write.csv(stock_JNJ, here("Week 6", "stock_JNJ.csv"), row.names = FALSE)
write.csv(stock_JPM, here("Week 6", "stock_JPM.csv"), row.names = FALSE)
write.csv(stock_KO, here("Week 6", "stock_KO.csv"), row.names = FALSE)
write.csv(stock_KRFT, here("Week 6", "stock_KRFT.csv"), row.names = FALSE)
write.csv(stock_MCD, here("Week 6", "stock_MCD.csv"), row.names = FALSE)
write.csv(stock_MMM, here("Week 6", "stock_MMM.csv"), row.names = FALSE)
write.csv(stock_MRK, here("Week 6", "stock_MRK.csv"), row.names = FALSE)
write.csv(stock_MSFT, here("Week 6", "stock_MSFT.csv"), row.names = FALSE)
write.csv(stock_PFE, here("Week 6", "stock_PFE.csv"), row.names = FALSE)
write.csv(stock_PG, here("Week 6", "stock_PG.csv"), row.names = FALSE)
write.csv(stock_TRV, here("Week 6", "stock_TRV.csv"), row.names = FALSE)
write.csv(stock_UTX, here("Week 6", "stock_UTX.csv"), row.names = FALSE)
write.csv(stock_VZ, here("Week 6", "stock_VZ.csv"), row.names = FALSE)
write.csv(stock_WMT, here("Week 6", "stock_WMT.csv"), row.names = FALSE)
write.csv(stock_XOM, here("Week 6", "stock_XOM.csv"), row.names = FALSE)

cat("Data cleaning completed for all stocks.\n")

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Remove the 'stock' column for all stocks
stock_AA <- stock_AA %>% select(-stock)
stock_AXP <- stock_AXP %>% select(-stock)
stock_BA <- stock_BA %>% select(-stock)
stock_BAC <- stock_BAC %>% select(-stock)
stock_CAT <- stock_CAT %>% select(-stock)
stock_CSCO <- stock_CSCO %>% select(-stock)
stock_CVX <- stock_CVX %>% select(-stock)
stock_DIS <- stock_DIS %>% select(-stock)
stock_DD <- stock_DD %>% select(-stock)
stock_GE <- stock_GE %>% select(-stock)
stock_HD <- stock_HD %>% select(-stock)
stock_HPQ <- stock_HPQ %>% select(-stock)
stock_IBM <- stock_IBM %>% select(-stock)
stock_INTC <- stock_INTC %>% select(-stock)
stock_JNJ <- stock_JNJ %>% select(-stock)
stock_JPM <- stock_JPM %>% select(-stock)
stock_KO <- stock_KO %>% select(-stock)
stock_KRFT <- stock_KRFT %>% select(-stock)
stock_MCD <- stock_MCD %>% select(-stock)
stock_MMM <- stock_MMM %>% select(-stock)
stock_MRK <- stock_MRK %>% select(-stock)
stock_MSFT <- stock_MSFT %>% select(-stock)
stock_PFE <- stock_PFE %>% select(-stock)
stock_PG <- stock_PG %>% select(-stock)
stock_TRV <- stock_TRV %>% select(-stock)
stock_UTX <- stock_UTX %>% select(-stock)
stock_VZ <- stock_VZ %>% select(-stock)
stock_WMT <- stock_WMT %>% select(-stock)
stock_XOM <- stock_XOM %>% select(-stock)

cat("Removed the 'stock' column for all stocks.\n")

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Compute returns and remove missing values for all stocks
ReturnAA <- na.omit(Delt(stock_AA$close))
ReturnAXP <- na.omit(Delt(stock_AXP$close))
ReturnBA <- na.omit(Delt(stock_BA$close))
ReturnBAC <- na.omit(Delt(stock_BAC$close))
ReturnCAT <- na.omit(Delt(stock_CAT$close))
ReturnCSCO <- na.omit(Delt(stock_CSCO$close))
ReturnCVX <- na.omit(Delt(stock_CVX$close))
ReturnDIS <- na.omit(Delt(stock_DIS$close))
ReturnDD <- na.omit(Delt(stock_DD$close))
ReturnGE <- na.omit(Delt(stock_GE$close))
ReturnHD <- na.omit(Delt(stock_HD$close))
ReturnHPQ <- na.omit(Delt(stock_HPQ$close))
ReturnIBM <- na.omit(Delt(stock_IBM$close))
ReturnINTC <- na.omit(Delt(stock_INTC$close))
ReturnJNJ <- na.omit(Delt(stock_JNJ$close))
ReturnJPM <- na.omit(Delt(stock_JPM$close))
ReturnKO <- na.omit(Delt(stock_KO$close))
ReturnKRFT <- na.omit(Delt(stock_KRFT$close))
ReturnMCD <- na.omit(Delt(stock_MCD$close))
ReturnMMM <- na.omit(Delt(stock_MMM$close))
ReturnMRK <- na.omit(Delt(stock_MRK$close))
ReturnMSFT <- na.omit(Delt(stock_MSFT$close))
ReturnPFE <- na.omit(Delt(stock_PFE$close))
ReturnPG <- na.omit(Delt(stock_PG$close))
ReturnTRV <- na.omit(Delt(stock_TRV$close))
ReturnUTX <- na.omit(Delt(stock_UTX$close))
ReturnVZ <- na.omit(Delt(stock_VZ$close))
ReturnWMT <- na.omit(Delt(stock_WMT$close))
ReturnXOM <- na.omit(Delt(stock_XOM$close))

# Compute returns for averaged data (Dow Jones index)
ReturnStock <- na.omit(Delt(averaged_data$avg_close))

cat("Computed returns and removed missing values for all stocks.\n")

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Merge the returns of all three index into single data
MyData = cbind(ReturnBA, ReturnGE, ReturnINTC, ReturnJNJ, ReturnMCD, ReturnStock)
```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Merge the returns of all stocks into a single data frame
MyData <- cbind(
  ReturnAA, ReturnAXP, ReturnBA, ReturnBAC, ReturnCAT, 
  ReturnCSCO, ReturnCVX, ReturnDIS, ReturnDD, ReturnGE,
  ReturnHD, ReturnHPQ, ReturnIBM, ReturnINTC, ReturnJNJ,
  ReturnJPM, ReturnKO, ReturnKRFT, ReturnMCD, ReturnMMM,
  ReturnMRK, ReturnMSFT, ReturnPFE, ReturnPG, ReturnTRV,
  ReturnUTX, ReturnVZ, ReturnWMT, ReturnXOM, ReturnStock
)

# Set column names, excluding NKE
colnames(MyData) <- c(
  "AA", "AXP", "BA", "BAC", "CAT", 
  "CSCO", "CVX", "DIS", "DD", "GE",
  "HD", "HPQ", "IBM", "INTC", "JNJ",
  "JPM", "KO", "KRFT", "MCD", "MMM",
  "MRK", "MSFT", "PFE", "PG", "TRV",
  "UTX", "VZ", "WMT", "XOM", "Dow"
)

# Display the first few rows of the combined data
head(MyData)

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# See how the data looks. You can see the risk is much lesser in the S&P as this 
# represents the market, compared to the individual stock. 
boxplot(MyData,main="Expected Return", xlab="Stock Picks", ylab="Return")

```
```{r echo=FALSE, warning=FALSE, error=FALSE}

# Number of stocks to compare in each plot
stocks_per_plot <- 4

# Get the column names excluding "Dow"
stock_columns <- colnames(MyData)[colnames(MyData) != "Dow"]

# Loop through the stocks in chunks of 4 and create boxplots
for (i in seq(1, length(stock_columns), by = stocks_per_plot)) {
  # Subset the stocks for the current plot and include Dow
  selected_columns <- c(stock_columns[i:min(i + stocks_per_plot - 1, length(stock_columns))], "Dow")

  # Create a boxplot for the selected stocks
  boxplot(
    MyData[, selected_columns, drop = FALSE],
    main = paste("Expected Return for Stocks", paste(selected_columns, collapse = ", ")),
    xlab = "Stock Picks",
    ylab = "Return"
  )
}

```
```{r}

# Load the here package
library(here)

# Define the number of stocks to compare in each plot
stocks_per_plot <- 4

# Get the column names excluding "Dow"
stock_columns <- colnames(MyData)[colnames(MyData) != "Dow"]

# Define the output folder using the here package
output_folder <- here("plots")

# Create the folder if it doesn't exist
if (!dir.exists(output_folder)) {
  dir.create(output_folder, recursive = TRUE)
}

# Loop through the stocks in chunks of 4 and create boxplots
for (i in seq(1, length(stock_columns), by = stocks_per_plot)) {
  # Subset the stocks for the current plot and include Dow
  selected_columns <- c(stock_columns[i:min(i + stocks_per_plot - 1, length(stock_columns))], "Dow")

  # Define the filename for the current plot
  plot_filename <- file.path(output_folder, paste0("Boxplot_", i, ".png"))

  # Save the boxplot as a PNG file
  png(plot_filename, width = 800, height = 600)
  boxplot(
    MyData[, selected_columns, drop = FALSE],
    main = paste("Expected Return for Stocks", paste(selected_columns, collapse = ", ")),
    xlab = "Stock Picks",
    ylab = "Return"
  )
  dev.off()  # Close the graphics device
}

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Compute mean and stdev for the returns.
DataMean=apply(MyData, 2, mean)
DataSD=apply(MyData, 2, sd)
# Take a look at the means and standard deviations. 
cbind(DataMean,DataSD)
```
```{r echo=FALSE, warning=FALSE, error=FALSE}

# List of stocks to process
stocks <- colnames(MyData)[colnames(MyData) != "Dow"] # Exclude Dow

# Initialize an empty data frame to store results
results <- data.frame(Stock = character(), Beta = numeric(), R_Squared = numeric(), stringsAsFactors = FALSE)

# Loop through each stock and compute the linear model
for (stock in stocks) {
  # Construct the formula dynamically
  formula <- as.formula(paste(stock, "~ Dow"))

  # Fit the linear model
  lm_model <- lm(formula, data = as.data.frame(MyData))

  # Extract Beta (coefficient for Dow) and R-squared
  beta <- summary(lm_model)$coefficients[2, 1]
  r_squared <- summary(lm_model)$r.squared

  # Append results to the data frame
  results <- rbind(results, data.frame(Stock = stock, Beta = beta, R_Squared = r_squared))
}

# Order results by Beta (safest to riskiest)
results <- results[order(results$Beta), ]

# Print as a long list
for (i in 1:nrow(results)) {
  cat(paste0(
    "Stock: ", results$Stock[i], 
    ", Beta: ", round(results$Beta[i], 4), 
    ", R-Squared: ", round(results$R_Squared[i], 4), "\n"
  ))
}

# Save the ordered results table to a CSV file (optional)
write.csv(results, file = here("Week 6", "stock_beta_results_sorted.csv"), row.names = FALSE)

```
```{r echo=FALSE, warning=FALSE, error=FALSE}

# Load the required library
library(flextable)

# List of stocks to process
stocks <- colnames(MyData)[colnames(MyData) != "Dow"] # Exclude Dow

# Initialize an empty data frame to store results
results <- data.frame(Stock = character(), Beta = numeric(), R_Squared = numeric(), Dow_Beta = numeric(), stringsAsFactors = FALSE)

# Loop through each stock and compute the linear model
for (stock in stocks) {
  # Construct the formula dynamically
  formula <- as.formula(paste(stock, "~ Dow"))

  # Fit the linear model
  lm_model <- lm(formula, data = as.data.frame(MyData))

  # Extract Beta (coefficient for Dow) and R-squared
  beta <- summary(lm_model)$coefficients[2, 1]
  r_squared <- summary(lm_model)$r.squared

  # Append results to the data frame
  results <- rbind(results, data.frame(Stock = stock, Beta = beta, R_Squared = r_squared, Dow_Beta = 1)) # Dow_Beta is always 1
}

# Order results by Beta (safest to riskiest)
results <- results[order(results$Beta), ]

# Create the flextable
ft <- flextable(results) %>%
  # Add a title
  set_caption("Beta vs Dow Beta") %>%
  # Highlight the Dow Beta column in light blue
  bg(j = "Dow_Beta", bg = "lightblue", part = "body") %>%
  # Format numeric values
  colformat_double(j = c("Beta", "R_Squared", "Dow_Beta"), digits = 4) %>%
  # Autofit column sizes
  autofit()

# Print the table
ft

```

```{r}

# Load the required libraries
library(flextable)
library(here)
library(webshot)

# Install webshot dependencies if not already installed
# webshot::install_phantomjs()

# Define the output folder using the here package
output_folder <- here("tables")

# Create the folder if it doesn't exist
if (!dir.exists(output_folder)) {
  dir.create(output_folder, recursive = TRUE)
}

# List of stocks to process (excluding "Dow")
stocks <- colnames(MyData)[colnames(MyData) != "Dow"]

# Initialize an empty data frame to store results
results <- data.frame(Stock = character(), Beta = numeric(), R_Squared = numeric(), Dow_Beta = numeric(), stringsAsFactors = FALSE)

# Loop through each stock and compute the linear model
for (stock in stocks) {
  # Construct the formula dynamically
  formula <- as.formula(paste(stock, "~ Dow"))

  # Fit the linear model
  lm_model <- lm(formula, data = as.data.frame(MyData))

  # Extract Beta (coefficient for Dow) and R-squared
  beta <- summary(lm_model)$coefficients[2, 1]
  r_squared <- summary(lm_model)$r.squared

  # Append results to the data frame
  results <- rbind(results, data.frame(Stock = stock, Beta = beta, R_Squared = r_squared, Dow_Beta = 1)) # Dow_Beta is always 1
}

# Order results by Beta (safest to riskiest)
results <- results[order(results$Beta), ]

# Split results into chunks of stocks per table
stocks_per_table <- 10
num_tables <- ceiling(nrow(results) / stocks_per_table)

# Loop to create and save each table as an image
for (i in seq(1, num_tables)) {
  # Subset the data for the current table
  start_row <- (i - 1) * stocks_per_table + 1
  end_row <- min(i * stocks_per_table, nrow(results))
  table_data <- results[start_row:end_row, ]

  # Create the flextable for the current chunk
  ft <- flextable(table_data) %>%
    set_caption(paste("Beta vs Dow Beta (Table", i, ")")) %>%
    bg(j = "Dow_Beta", bg = "lightblue", part = "body") %>%
    colformat_double(j = c("Beta", "R_Squared", "Dow_Beta"), digits = 4) %>%
    autofit()

  # Define the filename for the current table
  image_filename <- file.path(output_folder, paste0("Table_", i, ".png"))

  # Save the flextable as an image
  save_as_image(ft, path = image_filename)
}

```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Load the required library
library(flextable)

# Inputs for CAPM
risk_free_rate <- 0.03 # Example: 3% annualized (e.g., 10-year Treasury bond yield)
market_return <- 0.08  # Example: 8% annualized expected return of Dow Jones index
market_risk_premium <- market_return - risk_free_rate

# Assuming `results` contains Beta values calculated earlier
# Add Expected Return based on the CAPM formula
results$Expected_Return <- risk_free_rate + (results$Beta * market_risk_premium)

# Order by Expected Return (safest to riskiest stock based on CAPM)
results <- results[order(results$Expected_Return), ]

# Create a flextable for results
ft <- flextable(results) %>%
  # Add a title
  set_caption("CAPM: Expected Returns vs Dow Beta") %>%
  # Highlight the Dow Beta column in light blue
  bg(j = "Dow_Beta", bg = "lightblue", part = "body") %>%
  # Format numeric columns
  colformat_double(j = c("Beta", "R_Squared", "Dow_Beta", "Expected_Return"), digits = 4) %>%
  # Autofit column sizes
  autofit()

# Display the flextable
ft

# Save the results table to a CSV file (optional)
write.csv(results, file = here("Week 6", "stock_capm_results.csv"), row.names = FALSE)

# Print a message to indicate successful calculation
cat("CAPM calculations and table generation completed successfully.\n")

```

```{r}

```

```{r echo=FALSE, warning=FALSE, error=FALSE}
# Ensure MyData is a data frame
MyData <- as.data.frame(MyData)

# Load required libraries
library(ggplot2)
library(patchwork) # For arranging plots in a grid

# Exclude "Dow" from the stock list
stock_columns <- colnames(MyData)[colnames(MyData) != "Dow"]

# Initialize an empty list to store plots
plot_list <- list()

# Loop through each stock to generate plots
for (stock in stock_columns) {
  # Create a data frame for the current stock and the Dow
  plot_data <- data.frame(
    Market_Excess_Return = MyData$Dow, # Dow is the market
    Stock_Excess_Return = MyData[[stock]]
  )

  # Generate the scatter plot with regression line
  p <- ggplot(plot_data, aes(x = Market_Excess_Return, y = Stock_Excess_Return)) +
    geom_point(color = "blue", alpha = 0.6) +  # Scatter plot points
    geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
    labs(
      title = paste(stock, "vs. Dow Jones"),
      x = "Weekly Market Excess Return (%)",
      y = "Weekly Stock Excess Return (%)"
    ) +
    theme_minimal() +
    annotate(
      "text",
      x = max(plot_data$Market_Excess_Return, na.rm = TRUE) * 0.7,
      y = max(plot_data$Stock_Excess_Return, na.rm = TRUE) * 0.7,
      label = paste(
        "Alpha =", round(coef(lm(Stock_Excess_Return ~ Market_Excess_Return, data = plot_data))[1], 4),
        "\nBeta =", round(coef(lm(Stock_Excess_Return ~ Market_Excess_Return, data = plot_data))[2], 4)
      ),
      color = "black"
    )

  # Append the plot to the list
  plot_list[[stock]] <- p
}

# Combine plots into a grid in batches of 3
for (i in seq(1, length(plot_list), by = 3)) {
  # Select up to 3 plots for the current batch
  selected_plots <- plot_list[i:min(i + 2, length(plot_list))]

  # Combine the selected plots into a 1 x 3 grid
  combined_plot <- wrap_plots(selected_plots, ncol = 3)

  # Save the combined plot as a PNG
  ggsave(
    filename = paste0("Week_6_Batch_", ceiling(i / 3), "_Graphs.png"),
    plot = combined_plot,
    width = 12, height = 4
  )
}

```
```{r}

# Load required libraries
library(flextable)
library(here)
library(webshot)

# Install webshot dependencies if not already installed
# webshot::install_phantomjs()

# Inputs for CAPM
risk_free_rate <- 0.03  # Example: 3% annualized (e.g., 10-year Treasury bond yield)
market_return <- 0.08   # Example: 8% annualized expected return of Dow Jones index
market_risk_premium <- market_return - risk_free_rate

# Assuming `results` contains Beta values calculated earlier
# Add Expected Return based on the CAPM formula
results$Expected_Return <- risk_free_rate + (results$Beta * market_risk_premium)

# Order by Expected Return (safest to riskiest stock based on CAPM)
results <- results[order(results$Expected_Return), ]

# Create a flextable for results with proper styling
ft <- flextable(results) %>%
  set_caption("CAPM: Expected Returns vs Dow Beta") %>%
  bg(bg = "white", part = "all") %>%  # Set white background for the entire table
  color(color = "black", part = "all") %>%  # Set text color to black
  border_outer() %>%
  colformat_double(j = c("Beta", "R_Squared", "Dow_Beta", "Expected_Return"), digits = 4) %>%
  autofit()

# Define the output folder for the image
output_folder <- here("Week 6")
if (!dir.exists(output_folder)) {
  dir.create(output_folder, recursive = TRUE)
}

# Define the image path
image_path <- file.path(output_folder, "CAPM_Results_Table.png")

# Save the flextable as an image with proper rendering
save_as_image(ft, path = image_path)

# Print a message to indicate successful saving
```

```{r echo=FALSE, warning=FALSE, error=FALSE}

# Stock Analysis Report

# This report includes visualizations for the analysis of stock performance data. Below are the charts for each batch of stocks analyzed.

## Visualizations

# Load knitr package
library(knitr)

# Display all images in a single code chunk
knitr::include_graphics(c(
  "Week_6_Batch_1_Graphs.png",
  "Week_6_Batch_2_Graphs.png",
  "Week_6_Batch_3_Graphs.png",
  "Week_6_Batch_4_Graphs.png",
  "Week_6_Batch_5_Graphs.png",
  "Week_6_Batch_6_Graphs.png",
  "Week_6_Batch_7_Graphs.png",
  "Week_6_Batch_8_Graphs.png",
  "Week_6_Batch_9_Graphs.png",
  "Week_6_Batch_10_Graphs.png"
))

```

### 2024-12-03 R Projects/CAPM_CASESTUDY.qmd
updated assignments to github

--- 2024-12-03 R Projects/CAPM CASE STUDY 3.qmd
@@
 
-```{r echo=FALSE}
+```{r}
 
@@
 
-here()
-
 ```
 
-```{r echo=FALSE, warning=FALSE, error=FALSE}
+```{r}
 
@@
 
-```{r echo=FALSE, warning=FALSE, error=FALSE}
+```{r}
 
@@
 
-```{r echo=FALSE, warning=FALSE, error=FALSE}
+```{r}
 
-# removed.
+# Remove rows with missing data and create a new dataset named data_clean
+data_clean <- na.omit(data)
 
-data_clean <- data
-
+str(data_clean)
 ```
 
-```{r echo=FALSE, warning=FALSE, error=FALSE}
+```{r}
 
@@
 
-```{r echo=FALSE, warning=FALSE, error=FALSE}
+```{r}
+
 # Step 1: Inspect the unique stock symbols
-unique_stocks <- unique(data_clean$stock) # Adjust column name if different
+unique_stocks <- unique(data$stock) # Adjust column name if different
 
@@
 set.seed(123) # Set seed for reproducibility
-selected_stocks <- sample(unique_stocks, 30)
+selected_stocks <- sample(unique_stocks, 5)
 
@@
   # Filter data for the current stock
-  stock_data <- data_clean %>%
+  stock_data <- data %>%
     filter(stock == !!stock) %>%
@@
 
-   # Create file path for each stock
-  file_name <- here( paste0("stock_", stock, ".csv"))
+  # Create a dedicated folder for the stock
+  stock_folder <- here("output_files", paste0("stock_", stock))
+  dir.create(stock_folder, showWarnings = FALSE, recursive = TRUE)
 
   # Save the stock data as a CSV file within its folder
-  file_name <- here( paste0("stock_", stock, ".csv"))
+  file_name <- here(stock_folder, paste0("stock_", stock, ".csv"))
   write.csv(stock_data, file_name, row.names = FALSE)
@@
 
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-# Step 1: Extract unique stock names from the main dataset
-stock_symbols <- unique(data_clean$stock) # Ensure 'data_clean' is already processed
-
-# Step 2: Load data for each stock into individual variables
-for (symbol in stock_symbols) {
-  # Construct file name dynamically
-  file_path <- here(paste0("stock_", symbol, ".csv"))
-
-  # Check if the file exists
-  if (file.exists(file_path)) {
-    # Read the stock data
-    assign(symbol, read.csv(file_path, header = TRUE, sep = ","))
-    cat("Loaded data for stock:", symbol, "\n")
-  } else {
-    warning(paste("File not found for stock:", symbol))
-  }
-}
-
-# Example: Check the structure of one dataset (e.g., AAPL)
-if (exists("AAPL")) {
-  str(AAPL)
-}
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Step 2: Select only the relevant variables
-clean_stock <- data_clean %>%
-  select(stock,date, open, high, low, close, volume) # Adjust column names to match your dataset
-
-# Step 3: Save the cleaned dataset as dowjones.csv
-output_path <- here("clean_stock.csv")
-write.csv(data_clean, output_path, row.names = FALSE)
-
-clean_stock <- read.csv(here("clean_stock.csv"), header = TRUE, sep = ",")
-
-str(clean_stock)
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-str(GE)
-
-str(clean_stock)
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Convert 'date' column to Date format and numeric conversion for the other columns
-clean_stock <- data %>%
-  mutate(
-    date = as.Date(date, format = "%m/%d/%Y"), # Convert 'date' to Date format
-    open = as.numeric(gsub("\\$", "", open)),  # Remove dollar sign and convert to numeric
-    high = as.numeric(gsub("\\$", "", high)),  # Remove dollar sign and convert to numeric
-    low = as.numeric(gsub("\\$", "", low)),    # Remove dollar sign and convert to numeric
-    close = as.numeric(gsub("\\$", "", close)),# Remove dollar sign and convert to numeric
-    volume = as.numeric(volume)                # Ensure 'volume' is numeric
-  )
-
-# Remove the 'stock' column
-clean_stock <- clean_stock %>%
-  select(date, open, high, low, close, volume)
-
-str(clean_stock)
-```
-
 ```{r}
 
-library(dplyr)
-
-# Ensure columns are properly formatted
-clean_stock <- clean_stock %>%
-  mutate(
-    date = as.Date(date, format = "%m/%d/%Y"),
-    open = as.numeric(gsub("\\$", "", open)),
-    high = as.numeric(gsub("\\$", "", high)),
-    low = as.numeric(gsub("\\$", "", low)),
-    close = as.numeric(gsub("\\$", "", close)),
-    volume = as.numeric(volume)
-  )
-
-# Check for issues in data conversion
-print(summary(data))  # Provides an overview of the data
-
-# Group by 'date' and calculate averages, handling NAs
-averaged_data <- clean_stock %>%
-  group_by(date) %>%
-  summarize(
-    avg_open = mean(open, na.rm = TRUE),
-    avg_high = mean(high, na.rm = TRUE),
-    avg_low = mean(low, na.rm = TRUE),
-    avg_close = mean(close, na.rm = TRUE),
-    avg_volume = mean(volume, na.rm = TRUE)
-  )
-
-# View the cleaned data
-print(head(averaged_data))
+# Use the here function to construct the file path and import the dataset
+BA <- read.csv(here("stock_BA.csv"), header = TRUE, sep = ",")
+GE <- read.csv(here("stock_GE.csv"), header = TRUE, sep = ",")
+INTC <- read.csv(here("stock_INTC.csv"), header = TRUE, sep = ",")
+JNJ <- read.csv(here("stock_JNJ.csv"), header = TRUE, sep = ",")
+MCD <- read.csv(here("stock_MCD.csv"), header = TRUE, sep = ",")
 ```
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Step 1: Define stock symbols
-stock_symbols <- unique(data_clean$stock)
-
-# Step 2: Remove the 'stock' column for each dataset dynamically
-for (symbol in stock_symbols) {
-  # Check if the dataset exists in the environment
-  if (exists(symbol)) {
-    # Get the dataset
-    stock_data <- get(symbol)
-
-    # Remove the 'stock' column
-    stock_data <- stock_data %>% select(-stock)
-
-    # Save the modified dataset back to its original variable
-    assign(symbol, stock_data)
-
-    # Print confirmation
-    cat("Removed 'stock' column for:", symbol, "\n")
-  } else {
-    warning(paste("Dataset not found for stock:", symbol))
-  }
-}
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-here()
-
-# Load the data files for all stocks
-stock_AA <- read.csv(here("Week 6", "stock_AA.csv"), stringsAsFactors = FALSE)
-stock_AXP <- read.csv(here("Week 6", "stock_AXP.csv"), stringsAsFactors = FALSE)
-stock_BA <- read.csv(here("Week 6", "stock_BA.csv"), stringsAsFactors = FALSE)
-stock_BAC <- read.csv(here("Week 6", "stock_BAC.csv"), stringsAsFactors = FALSE)
-stock_CAT <- read.csv(here("Week 6", "stock_CAT.csv"), stringsAsFactors = FALSE)
-stock_CSCO <- read.csv(here("Week 6", "stock_CSCO.csv"), stringsAsFactors = FALSE)
-stock_CVX <- read.csv(here("Week 6", "stock_CVX.csv"), stringsAsFactors = FALSE)
-stock_DIS <- read.csv(here("Week 6", "stock_DIS.csv"), stringsAsFactors = FALSE)
-stock_DD <- read.csv(here("Week 6", "stock_DD.csv"), stringsAsFactors = FALSE)
-stock_GE <- read.csv(here("Week 6", "stock_GE.csv"), stringsAsFactors = FALSE)
-stock_HD <- read.csv(here("Week 6", "stock_HD.csv"), stringsAsFactors = FALSE)
-stock_HPQ <- read.csv(here("Week 6", "stock_HPQ.csv"), stringsAsFactors = FALSE)
-stock_IBM <- read.csv(here("Week 6", "stock_IBM.csv"), stringsAsFactors = FALSE)
-stock_INTC <- read.csv(here("Week 6", "stock_INTC.csv"), stringsAsFactors = FALSE)
-stock_JNJ <- read.csv(here("Week 6", "stock_JNJ.csv"), stringsAsFactors = FALSE)
-stock_JPM <- read.csv(here("Week 6", "stock_JPM.csv"), stringsAsFactors = FALSE)
-stock_KO <- read.csv(here("Week 6", "stock_KO.csv"), stringsAsFactors = FALSE)
-stock_KRFT <- read.csv(here("Week 6", "stock_KRFT.csv"), stringsAsFactors = FALSE)
-stock_MCD <- read.csv(here("Week 6", "stock_MCD.csv"), stringsAsFactors = FALSE)
-stock_MMM <- read.csv(here("Week 6", "stock_MMM.csv"), stringsAsFactors = FALSE)
-stock_MRK <- read.csv(here("Week 6", "stock_MRK.csv"), stringsAsFactors = FALSE)
-stock_MSFT <- read.csv(here("Week 6", "stock_MSFT.csv"), stringsAsFactors = FALSE)
-stock_PFE <- read.csv(here("Week 6", "stock_PFE.csv"), stringsAsFactors = FALSE)
-stock_PG <- read.csv(here("Week 6", "stock_PG.csv"), stringsAsFactors = FALSE)
-stock_TRV <- read.csv(here("Week 6", "stock_TRV.csv"), stringsAsFactors = FALSE)
-stock_UTX <- read.csv(here("Week 6", "stock_UTX.csv"), stringsAsFactors = FALSE)
-stock_VZ <- read.csv(here("Week 6", "stock_VZ.csv"), stringsAsFactors = FALSE)
-stock_WMT <- read.csv(here("Week 6", "stock_WMT.csv"), stringsAsFactors = FALSE)
-stock_XOM <- read.csv(here("Week 6", "stock_XOM.csv"), stringsAsFactors = FALSE)
-
-# Save the cleaned data back to their respective files
-write.csv(stock_AA, here("Week 6", "stock_AA.csv"), row.names = FALSE)
-write.csv(stock_AXP, here("Week 6", "stock_AXP.csv"), row.names = FALSE)
-write.csv(stock_BA, here("Week 6", "stock_BA.csv"), row.names = FALSE)
-write.csv(stock_BAC, here("Week 6", "stock_BAC.csv"), row.names = FALSE)
-write.csv(stock_CAT, here("Week 6", "stock_CAT.csv"), row.names = FALSE)
-write.csv(stock_CSCO, here("Week 6", "stock_CSCO.csv"), row.names = FALSE)
-write.csv(stock_CVX, here("Week 6", "stock_CVX.csv"), row.names = FALSE)
-write.csv(stock_DIS, here("Week 6", "stock_DIS.csv"), row.names = FALSE)
-write.csv(stock_DD, here("Week 6", "stock_DD.csv"), row.names = FALSE)
-write.csv(stock_GE, here("Week 6", "stock_GE.csv"), row.names = FALSE)
-write.csv(stock_HD, here("Week 6", "stock_HD.csv"), row.names = FALSE)
-write.csv(stock_HPQ, here("Week 6", "stock_HPQ.csv"), row.names = FALSE)
-write.csv(stock_IBM, here("Week 6", "stock_IBM.csv"), row.names = FALSE)
-write.csv(stock_INTC, here("Week 6", "stock_INTC.csv"), row.names = FALSE)
-write.csv(stock_JNJ, here("Week 6", "stock_JNJ.csv"), row.names = FALSE)
-write.csv(stock_JPM, here("Week 6", "stock_JPM.csv"), row.names = FALSE)
-write.csv(stock_KO, here("Week 6", "stock_KO.csv"), row.names = FALSE)
-write.csv(stock_KRFT, here("Week 6", "stock_KRFT.csv"), row.names = FALSE)
-write.csv(stock_MCD, here("Week 6", "stock_MCD.csv"), row.names = FALSE)
-write.csv(stock_MMM, here("Week 6", "stock_MMM.csv"), row.names = FALSE)
-write.csv(stock_MRK, here("Week 6", "stock_MRK.csv"), row.names = FALSE)
-write.csv(stock_MSFT, here("Week 6", "stock_MSFT.csv"), row.names = FALSE)
-write.csv(stock_PFE, here("Week 6", "stock_PFE.csv"), row.names = FALSE)
-write.csv(stock_PG, here("Week 6", "stock_PG.csv"), row.names = FALSE)
-write.csv(stock_TRV, here("Week 6", "stock_TRV.csv"), row.names = FALSE)
-write.csv(stock_UTX, here("Week 6", "stock_UTX.csv"), row.names = FALSE)
-write.csv(stock_VZ, here("Week 6", "stock_VZ.csv"), row.names = FALSE)
-write.csv(stock_WMT, here("Week 6", "stock_WMT.csv"), row.names = FALSE)
-write.csv(stock_XOM, here("Week 6", "stock_XOM.csv"), row.names = FALSE)
-
-cat("Data cleaning completed for all stocks.\n")
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Remove the 'stock' column for all stocks
-stock_AA <- stock_AA %>% select(-stock)
-stock_AXP <- stock_AXP %>% select(-stock)
-stock_BA <- stock_BA %>% select(-stock)
-stock_BAC <- stock_BAC %>% select(-stock)
-stock_CAT <- stock_CAT %>% select(-stock)
-stock_CSCO <- stock_CSCO %>% select(-stock)
-stock_CVX <- stock_CVX %>% select(-stock)
-stock_DIS <- stock_DIS %>% select(-stock)
-stock_DD <- stock_DD %>% select(-stock)
-stock_GE <- stock_GE %>% select(-stock)
-stock_HD <- stock_HD %>% select(-stock)
-stock_HPQ <- stock_HPQ %>% select(-stock)
-stock_IBM <- stock_IBM %>% select(-stock)
-stock_INTC <- stock_INTC %>% select(-stock)
-stock_JNJ <- stock_JNJ %>% select(-stock)
-stock_JPM <- stock_JPM %>% select(-stock)
-stock_KO <- stock_KO %>% select(-stock)
-stock_KRFT <- stock_KRFT %>% select(-stock)
-stock_MCD <- stock_MCD %>% select(-stock)
-stock_MMM <- stock_MMM %>% select(-stock)
-stock_MRK <- stock_MRK %>% select(-stock)
-stock_MSFT <- stock_MSFT %>% select(-stock)
-stock_PFE <- stock_PFE %>% select(-stock)
-stock_PG <- stock_PG %>% select(-stock)
-stock_TRV <- stock_TRV %>% select(-stock)
-stock_UTX <- stock_UTX %>% select(-stock)
-stock_VZ <- stock_VZ %>% select(-stock)
-stock_WMT <- stock_WMT %>% select(-stock)
-stock_XOM <- stock_XOM %>% select(-stock)
-
-cat("Removed the 'stock' column for all stocks.\n")
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Compute returns and remove missing values for all stocks
-ReturnAA <- na.omit(Delt(stock_AA$close))
-ReturnAXP <- na.omit(Delt(stock_AXP$close))
-ReturnBA <- na.omit(Delt(stock_BA$close))
-ReturnBAC <- na.omit(Delt(stock_BAC$close))
-ReturnCAT <- na.omit(Delt(stock_CAT$close))
-ReturnCSCO <- na.omit(Delt(stock_CSCO$close))
-ReturnCVX <- na.omit(Delt(stock_CVX$close))
-ReturnDIS <- na.omit(Delt(stock_DIS$close))
-ReturnDD <- na.omit(Delt(stock_DD$close))
-ReturnGE <- na.omit(Delt(stock_GE$close))
-ReturnHD <- na.omit(Delt(stock_HD$close))
-ReturnHPQ <- na.omit(Delt(stock_HPQ$close))
-ReturnIBM <- na.omit(Delt(stock_IBM$close))
-ReturnINTC <- na.omit(Delt(stock_INTC$close))
-ReturnJNJ <- na.omit(Delt(stock_JNJ$close))
-ReturnJPM <- na.omit(Delt(stock_JPM$close))
-ReturnKO <- na.omit(Delt(stock_KO$close))
-ReturnKRFT <- na.omit(Delt(stock_KRFT$close))
-ReturnMCD <- na.omit(Delt(stock_MCD$close))
-ReturnMMM <- na.omit(Delt(stock_MMM$close))
-ReturnMRK <- na.omit(Delt(stock_MRK$close))
-ReturnMSFT <- na.omit(Delt(stock_MSFT$close))
-ReturnPFE <- na.omit(Delt(stock_PFE$close))
-ReturnPG <- na.omit(Delt(stock_PG$close))
-ReturnTRV <- na.omit(Delt(stock_TRV$close))
-ReturnUTX <- na.omit(Delt(stock_UTX$close))
-ReturnVZ <- na.omit(Delt(stock_VZ$close))
-ReturnWMT <- na.omit(Delt(stock_WMT$close))
-ReturnXOM <- na.omit(Delt(stock_XOM$close))
-
-# Compute returns for averaged data (Dow Jones index)
-ReturnStock <- na.omit(Delt(averaged_data$avg_close))
-
-cat("Computed returns and removed missing values for all stocks.\n")
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Merge the returns of all three index into single data
-MyData = cbind(ReturnBA, ReturnGE, ReturnINTC, ReturnJNJ, ReturnMCD, ReturnStock)
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Merge the returns of all stocks into a single data frame
-MyData <- cbind(
-  ReturnAA, ReturnAXP, ReturnBA, ReturnBAC, ReturnCAT, 
-  ReturnCSCO, ReturnCVX, ReturnDIS, ReturnDD, ReturnGE,
-  ReturnHD, ReturnHPQ, ReturnIBM, ReturnINTC, ReturnJNJ,
-  ReturnJPM, ReturnKO, ReturnKRFT, ReturnMCD, ReturnMMM,
-  ReturnMRK, ReturnMSFT, ReturnPFE, ReturnPG, ReturnTRV,
-  ReturnUTX, ReturnVZ, ReturnWMT, ReturnXOM, ReturnStock
-)
-
-# Set column names, excluding NKE
-colnames(MyData) <- c(
-  "AA", "AXP", "BA", "BAC", "CAT", 
-  "CSCO", "CVX", "DIS", "DD", "GE",
-  "HD", "HPQ", "IBM", "INTC", "JNJ",
-  "JPM", "KO", "KRFT", "MCD", "MMM",
-  "MRK", "MSFT", "PFE", "PG", "TRV",
-  "UTX", "VZ", "WMT", "XOM", "Dow"
-)
-
-# Display the first few rows of the combined data
-head(MyData)
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# See how the data looks. You can see the risk is much lesser in the S&P as this 
-# represents the market, compared to the individual stock. 
-boxplot(MyData,main="Expected Return", xlab="Stock Picks", ylab="Return")
-
-```
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Number of stocks to compare in each plot
-stocks_per_plot <- 4
-
-# Get the column names excluding "Dow"
-stock_columns <- colnames(MyData)[colnames(MyData) != "Dow"]
-
-# Loop through the stocks in chunks of 4 and create boxplots
-for (i in seq(1, length(stock_columns), by = stocks_per_plot)) {
-  # Subset the stocks for the current plot and include Dow
-  selected_columns <- c(stock_columns[i:min(i + stocks_per_plot - 1, length(stock_columns))], "Dow")
-
-  # Create a boxplot for the selected stocks
-  boxplot(
-    MyData[, selected_columns, drop = FALSE],
-    main = paste("Expected Return for Stocks", paste(selected_columns, collapse = ", ")),
-    xlab = "Stock Picks",
-    ylab = "Return"
-  )
-}
-
-```
-```{r}
-
-# Load the here package
-library(here)
-
-# Define the number of stocks to compare in each plot
-stocks_per_plot <- 4
-
-# Get the column names excluding "Dow"
-stock_columns <- colnames(MyData)[colnames(MyData) != "Dow"]
-
-# Define the output folder using the here package
-output_folder <- here("plots")
-
-# Create the folder if it doesn't exist
-if (!dir.exists(output_folder)) {
-  dir.create(output_folder, recursive = TRUE)
-}
-
-# Loop through the stocks in chunks of 4 and create boxplots
-for (i in seq(1, length(stock_columns), by = stocks_per_plot)) {
-  # Subset the stocks for the current plot and include Dow
-  selected_columns <- c(stock_columns[i:min(i + stocks_per_plot - 1, length(stock_columns))], "Dow")
-
-  # Define the filename for the current plot
-  plot_filename <- file.path(output_folder, paste0("Boxplot_", i, ".png"))
-
-  # Save the boxplot as a PNG file
-  png(plot_filename, width = 800, height = 600)
-  boxplot(
-    MyData[, selected_columns, drop = FALSE],
-    main = paste("Expected Return for Stocks", paste(selected_columns, collapse = ", ")),
-    xlab = "Stock Picks",
-    ylab = "Return"
-  )
-  dev.off()  # Close the graphics device
-}
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Compute mean and stdev for the returns.
-DataMean=apply(MyData, 2, mean)
-DataSD=apply(MyData, 2, sd)
-# Take a look at the means and standard deviations. 
-cbind(DataMean,DataSD)
-```
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# List of stocks to process
-stocks <- colnames(MyData)[colnames(MyData) != "Dow"] # Exclude Dow
-
-# Initialize an empty data frame to store results
-results <- data.frame(Stock = character(), Beta = numeric(), R_Squared = numeric(), stringsAsFactors = FALSE)
-
-# Loop through each stock and compute the linear model
-for (stock in stocks) {
-  # Construct the formula dynamically
-  formula <- as.formula(paste(stock, "~ Dow"))
-
-  # Fit the linear model
-  lm_model <- lm(formula, data = as.data.frame(MyData))
-
-  # Extract Beta (coefficient for Dow) and R-squared
-  beta <- summary(lm_model)$coefficients[2, 1]
-  r_squared <- summary(lm_model)$r.squared
-
-  # Append results to the data frame
-  results <- rbind(results, data.frame(Stock = stock, Beta = beta, R_Squared = r_squared))
-}
-
-# Order results by Beta (safest to riskiest)
-results <- results[order(results$Beta), ]
-
-# Print as a long list
-for (i in 1:nrow(results)) {
-  cat(paste0(
-    "Stock: ", results$Stock[i], 
-    ", Beta: ", round(results$Beta[i], 4), 
-    ", R-Squared: ", round(results$R_Squared[i], 4), "\n"
-  ))
-}
-
-# Save the ordered results table to a CSV file (optional)
-write.csv(results, file = here("Week 6", "stock_beta_results_sorted.csv"), row.names = FALSE)
-
-```
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Load the required library
-library(flextable)
-
-# List of stocks to process
-stocks <- colnames(MyData)[colnames(MyData) != "Dow"] # Exclude Dow
-
-# Initialize an empty data frame to store results
-results <- data.frame(Stock = character(), Beta = numeric(), R_Squared = numeric(), Dow_Beta = numeric(), stringsAsFactors = FALSE)
-
-# Loop through each stock and compute the linear model
-for (stock in stocks) {
-  # Construct the formula dynamically
-  formula <- as.formula(paste(stock, "~ Dow"))
-
-  # Fit the linear model
-  lm_model <- lm(formula, data = as.data.frame(MyData))
-
-  # Extract Beta (coefficient for Dow) and R-squared
-  beta <- summary(lm_model)$coefficients[2, 1]
-  r_squared <- summary(lm_model)$r.squared
-
-  # Append results to the data frame
-  results <- rbind(results, data.frame(Stock = stock, Beta = beta, R_Squared = r_squared, Dow_Beta = 1)) # Dow_Beta is always 1
-}
-
-# Order results by Beta (safest to riskiest)
-results <- results[order(results$Beta), ]
-
-# Create the flextable
-ft <- flextable(results) %>%
-  # Add a title
-  set_caption("Beta vs Dow Beta") %>%
-  # Highlight the Dow Beta column in light blue
-  bg(j = "Dow_Beta", bg = "lightblue", part = "body") %>%
-  # Format numeric values
-  colformat_double(j = c("Beta", "R_Squared", "Dow_Beta"), digits = 4) %>%
-  # Autofit column sizes
-  autofit()
-
-# Print the table
-ft
-
-```
-
-```{r}
-
-# Load the required libraries
-library(flextable)
-library(here)
-library(webshot)
-
-# Install webshot dependencies if not already installed
-# webshot::install_phantomjs()
-
-# Define the output folder using the here package
-output_folder <- here("tables")
-
-# Create the folder if it doesn't exist
-if (!dir.exists(output_folder)) {
-  dir.create(output_folder, recursive = TRUE)
-}
-
-# List of stocks to process (excluding "Dow")
-stocks <- colnames(MyData)[colnames(MyData) != "Dow"]
-
-# Initialize an empty data frame to store results
-results <- data.frame(Stock = character(), Beta = numeric(), R_Squared = numeric(), Dow_Beta = numeric(), stringsAsFactors = FALSE)
-
-# Loop through each stock and compute the linear model
-for (stock in stocks) {
-  # Construct the formula dynamically
-  formula <- as.formula(paste(stock, "~ Dow"))
-
-  # Fit the linear model
-  lm_model <- lm(formula, data = as.data.frame(MyData))
-
-  # Extract Beta (coefficient for Dow) and R-squared
-  beta <- summary(lm_model)$coefficients[2, 1]
-  r_squared <- summary(lm_model)$r.squared
-
-  # Append results to the data frame
-  results <- rbind(results, data.frame(Stock = stock, Beta = beta, R_Squared = r_squared, Dow_Beta = 1)) # Dow_Beta is always 1
-}
-
-# Order results by Beta (safest to riskiest)
-results <- results[order(results$Beta), ]
-
-# Split results into chunks of stocks per table
-stocks_per_table <- 10
-num_tables <- ceiling(nrow(results) / stocks_per_table)
-
-# Loop to create and save each table as an image
-for (i in seq(1, num_tables)) {
-  # Subset the data for the current table
-  start_row <- (i - 1) * stocks_per_table + 1
-  end_row <- min(i * stocks_per_table, nrow(results))
-  table_data <- results[start_row:end_row, ]
-
-  # Create the flextable for the current chunk
-  ft <- flextable(table_data) %>%
-    set_caption(paste("Beta vs Dow Beta (Table", i, ")")) %>%
-    bg(j = "Dow_Beta", bg = "lightblue", part = "body") %>%
-    colformat_double(j = c("Beta", "R_Squared", "Dow_Beta"), digits = 4) %>%
-    autofit()
-
-  # Define the filename for the current table
-  image_filename <- file.path(output_folder, paste0("Table_", i, ".png"))
-
-  # Save the flextable as an image
-  save_as_image(ft, path = image_filename)
-}
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Load the required library
-library(flextable)
-
-# Inputs for CAPM
-risk_free_rate <- 0.03 # Example: 3% annualized (e.g., 10-year Treasury bond yield)
-market_return <- 0.08  # Example: 8% annualized expected return of Dow Jones index
-market_risk_premium <- market_return - risk_free_rate
-
-# Assuming `results` contains Beta values calculated earlier
-# Add Expected Return based on the CAPM formula
-results$Expected_Return <- risk_free_rate + (results$Beta * market_risk_premium)
-
-# Order by Expected Return (safest to riskiest stock based on CAPM)
-results <- results[order(results$Expected_Return), ]
-
-# Create a flextable for results
-ft <- flextable(results) %>%
-  # Add a title
-  set_caption("CAPM: Expected Returns vs Dow Beta") %>%
-  # Highlight the Dow Beta column in light blue
-  bg(j = "Dow_Beta", bg = "lightblue", part = "body") %>%
-  # Format numeric columns
-  colformat_double(j = c("Beta", "R_Squared", "Dow_Beta", "Expected_Return"), digits = 4) %>%
-  # Autofit column sizes
-  autofit()
-
-# Display the flextable
-ft
-
-# Save the results table to a CSV file (optional)
-write.csv(results, file = here("Week 6", "stock_capm_results.csv"), row.names = FALSE)
-
-# Print a message to indicate successful calculation
-cat("CAPM calculations and table generation completed successfully.\n")
-
-```
-
-```{r}
-
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-# Ensure MyData is a data frame
-MyData <- as.data.frame(MyData)
-
-# Load required libraries
-library(ggplot2)
-library(patchwork) # For arranging plots in a grid
-
-# Exclude "Dow" from the stock list
-stock_columns <- colnames(MyData)[colnames(MyData) != "Dow"]
-
-# Initialize an empty list to store plots
-plot_list <- list()
-
-# Loop through each stock to generate plots
-for (stock in stock_columns) {
-  # Create a data frame for the current stock and the Dow
-  plot_data <- data.frame(
-    Market_Excess_Return = MyData$Dow, # Dow is the market
-    Stock_Excess_Return = MyData[[stock]]
-  )
-
-  # Generate the scatter plot with regression line
-  p <- ggplot(plot_data, aes(x = Market_Excess_Return, y = Stock_Excess_Return)) +
-    geom_point(color = "blue", alpha = 0.6) +  # Scatter plot points
-    geom_smooth(method = "lm", color = "red", se = FALSE) +  # Regression line
-    labs(
-      title = paste(stock, "vs. Dow Jones"),
-      x = "Weekly Market Excess Return (%)",
-      y = "Weekly Stock Excess Return (%)"
-    ) +
-    theme_minimal() +
-    annotate(
-      "text",
-      x = max(plot_data$Market_Excess_Return, na.rm = TRUE) * 0.7,
-      y = max(plot_data$Stock_Excess_Return, na.rm = TRUE) * 0.7,
-      label = paste(
-        "Alpha =", round(coef(lm(Stock_Excess_Return ~ Market_Excess_Return, data = plot_data))[1], 4),
-        "\nBeta =", round(coef(lm(Stock_Excess_Return ~ Market_Excess_Return, data = plot_data))[2], 4)
-      ),
-      color = "black"
-    )
-
-  # Append the plot to the list
-  plot_list[[stock]] <- p
-}
-
-# Combine plots into a grid in batches of 3
-for (i in seq(1, length(plot_list), by = 3)) {
-  # Select up to 3 plots for the current batch
-  selected_plots <- plot_list[i:min(i + 2, length(plot_list))]
-
-  # Combine the selected plots into a 1 x 3 grid
-  combined_plot <- wrap_plots(selected_plots, ncol = 3)
-
-  # Save the combined plot as a PNG
-  ggsave(
-    filename = paste0("Week_6_Batch_", ceiling(i / 3), "_Graphs.png"),
-    plot = combined_plot,
-    width = 12, height = 4
-  )
-}
-
-```
-```{r}
-
-# Load required libraries
-library(flextable)
-library(here)
-library(webshot)
-
-# Install webshot dependencies if not already installed
-# webshot::install_phantomjs()
-
-# Inputs for CAPM
-risk_free_rate <- 0.03  # Example: 3% annualized (e.g., 10-year Treasury bond yield)
-market_return <- 0.08   # Example: 8% annualized expected return of Dow Jones index
-market_risk_premium <- market_return - risk_free_rate
-
-# Assuming `results` contains Beta values calculated earlier
-# Add Expected Return based on the CAPM formula
-results$Expected_Return <- risk_free_rate + (results$Beta * market_risk_premium)
-
-# Order by Expected Return (safest to riskiest stock based on CAPM)
-results <- results[order(results$Expected_Return), ]
-
-# Create a flextable for results with proper styling
-ft <- flextable(results) %>%
-  set_caption("CAPM: Expected Returns vs Dow Beta") %>%
-  bg(bg = "white", part = "all") %>%  # Set white background for the entire table
-  color(color = "black", part = "all") %>%  # Set text color to black
-  border_outer() %>%
-  colformat_double(j = c("Beta", "R_Squared", "Dow_Beta", "Expected_Return"), digits = 4) %>%
-  autofit()
-
-# Define the output folder for the image
-output_folder <- here("Week 6")
-if (!dir.exists(output_folder)) {
-  dir.create(output_folder, recursive = TRUE)
-}
-
-# Define the image path
-image_path <- file.path(output_folder, "CAPM_Results_Table.png")
-
-# Save the flextable as an image with proper rendering
-save_as_image(ft, path = image_path)
-
-# Print a message to indicate successful saving
-```
-
-```{r echo=FALSE, warning=FALSE, error=FALSE}
-
-# Stock Analysis Report
-
-# This report includes visualizations for the analysis of stock performance data. Below are the charts for each batch of stocks analyzed.
-
-## Visualizations
-
-# Load knitr package
-library(knitr)
-
-# Display all images in a single code chunk
-knitr::include_graphics(c(
-  "Week_6_Batch_1_Graphs.png",
-  "Week_6_Batch_2_Graphs.png",
-  "Week_6_Batch_3_Graphs.png",
-  "Week_6_Batch_4_Graphs.png",
-  "Week_6_Batch_5_Graphs.png",
-  "Week_6_Batch_6_Graphs.png",
-  "Week_6_Batch_7_Graphs.png",
-  "Week_6_Batch_8_Graphs.png",
-  "Week_6_Batch_9_Graphs.png",
-  "Week_6_Batch_10_Graphs.png"
-))
-
-```

### 2024-12-03 R Projects/CAPM.R

@@ -1,0 +1,84 @@
+# install.packages("tseries")
+# install.packages("quantmod")
+
+library(tseries)
+library(quantmod)
+library(here)
+library(tidyverse)
+
+# Go on yahoo finance, search GOOG (Google), FB (facebook), and ^GSPC (S&P 500). 
+# Obtain the prices from . 
+
+# Use the here function to construct the file path and import the dataset
+GoogData <- read.csv(here("GOOG.csv"), header = TRUE, sep = ",")
+FBData <- read.csv(here("FB.csv"), header = TRUE, sep = ",")
+SP500Data <- read.csv(here("^GSPC.csv"), header = TRUE, sep = ",")
+
+
+# Remove any missing values if needed
+ GoogData = na.omit(GoogData)
+
+# See what the data classes are
+str(GoogData)
+
+# Compute the returns and remove any missing values. 
+ReturnGoog = na.omit(Delt(GoogData[,5]))
+ReturnFB = na.omit(Delt(FBData[,5]))
+ReturnSP500 = na.omit(Delt(SP500Data[,5]))
+
+# Merge the returns of all three index into single data
+MyData = cbind(ReturnSP500,ReturnGoog,ReturnFB)
+
+# Monthly Return of Investment of Interest
+colnames(MyData) = c("SP500", "Google", "Facebook")
+head(MyData)
+
+# See how the data looks. You can see the risk is much lesser in the S&P as this 
+# represents the market, compared to the individual stock. 
+boxplot(MyData,main="Expected Return", xlab="Stock Picks", ylab="Return")
+
+# Compute mean and stdev for the returns.
+DataMean=apply(MyData, 2, mean)
+DataSD=apply(MyData, 2, sd)
+# Take a look at the means and standard deviations. 
+cbind(DataMean,DataSD)
+
+# According to the CAPM formula, we will first get the beta of each stock by 
+# regressions; then calculate the expected return of each stock and the covariance 
+# matrix of the four stocks; finally we can calculate the optimal asset allocations 
+# (weights) of the portfolio consisting of the 2 stocks.
+
+# Model for Google. Model fit can be judged by R-sq. Using LM here as interprtation 
+# is key. SVM is not so useful.
+lm.Goog<- lm(Google ~ SP500, data = as.data.frame(MyData))
+summary(lm.Goog)
+BetaGoog <- summary(lm.Goog)$coefficients[2, 1]
+
[…]
